---
title: "01b_Clustering_SCTransform_RegressionCellCycle"
author: "Julia Hitschfel"
date: "2026-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(dplyr)
library(ggplot2)
library(Seurat)
library(BPCells)
library(presto)
library(glmGamPoi)
library(hdf5r)
library(SingleCellExperiment)
library(scDblFinder)
library(scRepertoire)
library(RColorBrewer)

options(future.globals.maxSize= 734003200)
```


## Clustering of intestinal biopsy cells 

In this script we will analyze intestinal biopsy cells of patient CU1 until one year post therapy. The SCTransform workflow from Seurat will be used for analysis.  

Cells with very high nCount_RNA (>100,000), low nFeature_RNA (<200), and high frequency of mitochondrial reads (>12%) will be filtered out and Cell Cycle Effects on clustering will be assessed. The function quietVDJ genes of scRepertoire will be used to prevent effects of B and T cell receptor transcripts on cell type clustering: 


```{r Read Data}
All.data <- list.dirs("data/", recursive = F)

list.10x <- list()
for(i in All.data){
  rnaseq.data <- Read10X_h5(paste0(i, "/count/sample_filtered_feature_bc_matrix.h5"))
  pbmc <- CreateSeuratObject(counts = rnaseq.data, project = "colitis", min.cells = 3, 
                             min.features = 200)
  list.10x[[i]] <- pbmc
}
```

```{r Cell Counts before QC}
cell_counts_before_QC <- vector()
for(i in 1:length(list.10x)){
  cell_counts_before_QC[i] <- nrow(list.10x[[i]]@meta.data)
}
names(cell_counts_before_QC) <- substr(names(list.10x), 3, 100)
counts_before_qc <- as.data.frame(cell_counts_before_QC)
counts_before_qc
```


```{r QC and SCTransform}
qc_plots <- list()
qc_plots_doub_rem <- list()
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
for(i in 1:length(list.10x)){
  scobject <- list.10x[[i]]
  scobject[["percent.mt"]] <- PercentageFeatureSet(scobject, pattern = "^MT-")
  scobject@meta.data$tp <- substr(names(list.10x)[i], 3, 100)
  qc_graph <- ggplot(data = scobject@meta.data, aes(x = nCount_RNA, y = nFeature_RNA)) +
    geom_point(aes(colour = percent.mt))
  sce <- as.SingleCellExperiment(scobject)
  sce <- scDblFinder(sce, samples = sce@colData$tp)
  table(sce$scDblFinder.class)
  scobject$scDBLFinder <- sce$scDblFinder.class
  scobject <- subset(scobject, scDBLFinder == "singlet")
  qc_graph_doublet_rem <- ggplot(data = scobject@meta.data, aes(x = nCount_RNA, y = nFeature_RNA)) +
    geom_point(aes(colour = percent.mt))
  scobject <- subset(scobject, subset = nFeature_RNA > 200 & percent.mt < 12.5 & nCount_RNA < 100000)
  scobject <- NormalizeData(scobject)
  scobject <- FindVariableFeatures(scobject, selection.method = "vst")
  scobject <- ScaleData(scobject, features = rownames(scobject))
  scobject <- RunPCA(scobject, features = VariableFeatures(scobject))
  s.genes.filt <- intersect(s.genes, rownames(scobject))
  g2m.genes.filt <- intersect(g2m.genes, rownames(scobject))
  scobject <- CellCycleScoring(object = scobject, g2m.features = g2m.genes.filt, s.features = s.genes.filt)
  scobject <- SCTransform(scobject, vars.to.regress = c("S.Score", "G2M.Score"))
  list.10x[[i]] <- scobject
  qc_plots[[i]] <- qc_graph
  qc_plots_doub_rem[[i]] <- qc_graph_doublet_rem
}

scobject <- merge(x = list.10x[[1]], y = c(list.10x[2:length(list.10x)]))
ggplot(scobject@meta.data, aes(x = nCount_RNA, y = nFeature_RNA)) + 
  geom_point(aes(colour = percent.mt))
ggplot(scobject@meta.data, aes(x = tp, y = nCount_RNA)) + geom_boxplot(outlier.color = "red")
ggplot(scobject@meta.data, aes(x = tp, y = nFeature_RNA)) + geom_boxplot(outlier.color = "red")
ggplot(scobject@meta.data, aes(x = tp, y = percent.mt)) + geom_boxplot(outlier.color = "red")

sum_table <- scobject@meta.data %>% group_by(tp) %>% 
  summarise(RNA.count.median = median(nCount_RNA), RNA.feature.median = median(nFeature_RNA),
            MT.median = median(percent.mt))
sum_table
```

Filters for quality control are used before normalization with SCTransform, as data sets should not be subsetted after using SCTransform. 


```{r Cell counts after QC}
counts_after_qc <- table(scobject@meta.data$tp)
cell_counts <- cbind(counts_before_qc, counts_after_qc)
cell_counts
```


```{r Integration}
# Identify the 10 most highly variable genes
VariableFeatures(scobject[["SCT"]]) <- rownames(scobject[["SCT"]]@scale.data)
top10 <- head(VariableFeatures(scobject), 10)
top10

scobject <- RunPCA(scobject, features = VariableFeatures(object = scobject))

# Examine and visualize PCA results a few different ways
print(scobject[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(scobject, dims = 1:2, reduction = "pca")
DimPlot(scobject, reduction = "pca") + NoLegend()

#Check some initial confounding factors on the PCA
FeaturePlot(scobject, features = "nCount_RNA", reduction = "pca") 
FeaturePlot(scobject, features = "nCount_RNA", reduction = "pca", dims = c(1,3))
FeaturePlot(scobject, features = "nFeature_RNA", reduction = "pca") 
FeaturePlot(scobject, features = "nFeature_RNA", reduction = "pca", dims = c(1,3)) 
FeaturePlot(scobject, features = "percent.mt", reduction = "pca") 
FeaturePlot(scobject, features = "percent.mt", reduction = "pca", dims = c(1,3))

DimPlot(scobject, reduction = "pca", group.by = "tp", split.by = "tp", ncol = 3) 

DimHeatmap(scobject, dims = 1, cells = 500, balanced = TRUE)
DimHeatmap(scobject, dims = 1:15, cells = 500, balanced = TRUE)

ElbowPlot(scobject, ndims = 25)

# Integration
scobject_int <- IntegrateLayers(object = scobject, method = CCAIntegration, 
                                normalization.method = "SCT", verbose = FALSE)

# re-join layers after integration
scobject_int[["RNA"]] <- JoinLayers(scobject_int[["RNA"]])

# QC on integrated data

# Examine and visualize Integration results a few different ways
print(scobject_int[["integrated.dr"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(scobject_int, dims = 1:2, reduction = "integrated.dr")
DimPlot(scobject_int, reduction = "integrated.dr") + NoLegend()

#Check some initial confounding factors on the integrated dimensionality reduction
FeaturePlot(scobject_int, features = "nCount_RNA", reduction = "integrated.dr") 
FeaturePlot(scobject_int, features = "nCount_RNA", reduction = "integrated.dr", dims = c(1,3))
FeaturePlot(scobject_int, features = "nFeature_RNA", reduction = "integrated.dr") 
DimPlot(scobject_int, reduction = "integrated.dr", group.by = "Phase", split.by = "Phase") 

FeaturePlot(scobject_int, features = "nFeature_RNA", reduction = "integrated.dr", dims = c(1,3)) 
FeaturePlot(scobject_int, features = "percent.mt", reduction = "integrated.dr") 
FeaturePlot(scobject_int, features = "percent.mt", reduction = "integrated.dr", dims = c(1,3))

DimPlot(scobject_int, reduction = "integrated.dr", group.by = "tp", split.by = "tp", ncol = 3) 

DimHeatmap(scobject_int, dims = 1, cells = 500, balanced = TRUE)
DimHeatmap(scobject_int, dims = 1:15, cells = 500, balanced = TRUE)

ElbowPlot(scobject_int, ndims = 30)

scobject_int <- FindNeighbors(scobject_int, reduction = "integrated.dr", dims = 1:25)
scobject_int <- FindClusters(scobject_int, resolution = c(0.4,0.6,0.8,1))
scobject_int <- RunUMAP(scobject_int, dims = 1:25, reduction = "integrated.dr")
saveRDS(scobject_int, file = "integrated_seurat_cellcycle_regression_CU1_1year.rds")

# Visualization
DimPlot(scobject_int, reduction = "umap", group.by = "tp")
DimPlot(scobject_int, reduction = "umap", group.by = "tp", split.by = "tp", ncol = 3)

DimPlot(scobject_int, reduction = "umap", group.by = "SCT_snn_res.0.4", label = T)
DimPlot(scobject_int, reduction = "umap", group.by = "SCT_snn_res.0.6", label = T)
```

```{r Cells over time and marker expression}
Cells_per_cluster <- table(scobject_int@meta.data$SCT_snn_res.0.4)
Cells_per_cluster

Cells_per_cluster_06 <- table(scobject_int@meta.data$SCT_snn_res.0.6)
Cells_per_cluster_06

scobject_int <- PrepSCTFindMarkers(scobject_int)
Idents(scobject_int) <- "SCT_snn_res.0.4"
Cluster_markers <- FindAllMarkers(scobject_int, logfc.threshold = 0.25, min.pct = 0.5, only.pos = T)
Cluster_top10_markers <- Cluster_markers %>% group_by(cluster) %>% slice_head(n = 10)
Cluster_top10_markers
write.csv(Cluster_markers, 
       file = "Cluster_marker_genes_res04_cca_integrated_cellcycle_regress.csv")

# Resolution 0.6
Idents(scobject_int) <- "SCT_snn_res.0.6"
Cluster_markers_06 <- FindAllMarkers(scobject_int, logfc.threshold = 0.25, min.pct = 0.5, only.pos = T)
Cluster_top10_markers_06 <- Cluster_markers_06 %>% group_by(cluster) %>% slice_head(n = 10)
Cluster_top10_markers_06
write.csv(Cluster_markers_06, 
       file = "Cluster_marker_genes_res06_cca_integrated_cellcycle_regress.csv")
```

```{r Examine ribosomal transcripts}
scobject_int[["percent.rps"]] <- PercentageFeatureSet(scobject_int, pattern = "^RPS")
FeaturePlot(scobject_int, reduction = "pca", features = "percent.rps")
FeaturePlot(scobject_int, reduction = "pca", dims = c(1,3), features = "percent.rps")
FeaturePlot(scobject_int, reduction = "umap", features = "percent.rps")
FeaturePlot(scobject_int, reduction = "integrated.dr", features = "percent.rps")

VlnPlot(scobject_int, features = "percent.rps", group.by = "tp")

scobject_int[["percent.rpl"]] <- PercentageFeatureSet(scobject_int, pattern = "^RPL")
FeaturePlot(scobject_int, reduction = "pca", features = "percent.rpl")
FeaturePlot(scobject_int, reduction = "pca", dims = c(1,3), features = "percent.rpl")
FeaturePlot(scobject_int, reduction = "umap", features = "percent.rpl")
FeaturePlot(scobject_int, reduction = "integrated.dr", features = "percent.rpl")


VlnPlot(scobject_int, features = "percent.rpl", group.by = "tp")

VlnPlot(scobject_int, features = "nCount_RNA", group.by = "SCT_snn_res.0.4")
VlnPlot(scobject_int, features = "nFeature_RNA", group.by = "SCT_snn_res.0.4")
VlnPlot(scobject_int, features = "percent.mt", group.by = "SCT_snn_res.0.4")
```

```{r Cluster over time}
Cluster_over_time <- scobject_int@meta.data %>% group_by(tp, SCT_snn_res.0.4) %>% 
  summarise(n = n()) %>% mutate(Freq = n/sum(n))

farbenblind15 <- c("#000000","#004949","#009292","#FF6DB6","#FFB6DB",
                  "#490092","#006DDB","#B66DFF","#6DB6FF","#B6DBFF",
                  "#920000","#924900","#DB6D00","#24FF24","#FFFF6D") 
palette1 <- brewer.pal(8, "Set2")  # First palette
palette2 <- brewer.pal(8, "Dark2")

combined_palette <- c(farbenblind15, palette1, palette2)  # Combine the palettes

ggplot(Cluster_over_time, aes(x = tp, y = Freq, fill = SCT_snn_res.0.4)) + 
  geom_col(color = "black") + scale_fill_manual(values = combined_palette) + theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

# Resolution 0.6
Cluster_over_time <- scobject_int@meta.data %>% group_by(tp, SCT_snn_res.0.6) %>% 
  summarise(n = n()) %>% mutate(Freq = n/sum(n))

ggplot(Cluster_over_time, aes(x = tp, y = Freq, fill = SCT_snn_res.0.6)) + 
  geom_col(color = "black") + scale_fill_manual(values = combined_palette) + theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```
