---
title: "04_Tcell_Subclustering"
author: "Julia Hitschfel"
date: "2026-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(Seurat)
library(dplyr)
library(ggplot2)
library(BPCells)
library(presto)
library(glmGamPoi)
library(hdf5r)
library(knitr)
library(pheatmap)
library(RColorBrewer)

farbenblind15 <- c("#000000","#004949","#009292","#FF6DB6","#FFB6DB",
                  "#490092","#006DDB","#B66DFF","#6DB6FF","#B6DBFF",
                  "#920000","#924900","#DB6D00","#00A651","#FFFF6D") 
```


# T Cell subclustering analysis with Integration (CCA)

For the T cell subclustering, it is likely also useful to perform integration. As for the whole dataset, we will use SCTransform for normalization of raw data and CCA workflow of Seurat for integration. 
Since Doublet detection was performed in the whole dataset, by filtering on the barcodes that remained we will automatically exclude doublets from the T cell subclustering and won't perform a separate doublet detection. 

For subclustering analysis, the main T cell clusters of CD8+ T cells and CD4+ T cells will be included, as well as the CD45+ immune cell cluster to identify T cells contained in it. 

```{r Read in Data}
scobject <- readRDS("integrated_seurat_cellcycle_regression_CU1_1year_res04_annotated.rds")

All_Tcells_comb <- WhichCells(scobject, idents = c("CD8+ T cells","CD4+ T cells","CD45+ immune cells"))

FeaturePlot(scobject, features = "CD3E", cells = All_Tcells_comb, slot = "counts", max.cutoff = 3)
```
Almost all cells annotated as T cells express CD3D or CD3E. However, there are a few cells that don't express these markers and potentially are not T cells. 


```{r Establish filters}
# Canonical T cell marker genes
FeaturePlot(scobject, features = "CD3D", slot = "counts", max.cutoff = 5, cells = All_Tcells_comb)
FeaturePlot(scobject, features = "CD3E", slot = "counts", max.cutoff = 5, cells = All_Tcells_comb)
FeaturePlot(scobject, features = "CD247", slot = "counts", max.cutoff = 5, cells = All_Tcells_comb) #CD3 zeta
FeaturePlot(scobject, features = "CD2", slot = "counts", max.cutoff = 5, cells = All_Tcells_comb)
FeaturePlot(scobject, features = "CD5", slot = "counts", max.cutoff = 5, cells = All_Tcells_comb)
FeaturePlot(scobject, features = "PTPRC", slot = "counts", max.cutoff = 5, cells = All_Tcells_comb)

#KRT8 is a marker for epithelial cells. It will be used here as a filter to remove epithelial cells contaminating the T cell clusters
FeaturePlot(scobject, features = "KRT8", slot = "counts", max.cutoff = 5, cells = All_Tcells_comb)

#CD14 and CD68 expression is restricted to the myeloid cell cluster. These will be used as filters to remove contaminating monocytes and myeloid cells without losing T cells  
FeaturePlot(scobject, features = "CD14", slot = "counts", max.cutoff = 5, cells = All_Tcells_comb)
FeaturePlot(scobject, features = "CD68", slot = "counts", max.cutoff = 5, cells = All_Tcells_comb)

genes_to_filter <- c('CD3D', 'CD3E', 'CD247', 'CD2', 'CD5', 'PTPRC')
Non_Tcells <- WhichCells(scobject, expression = CD3D < 1 & CD3E < 1 & CD247 < 1 & CD2 < 1 & 
                           CD5 < 1 & PTPRC < 1, slot = "counts")

No_Tcells_Seurat <- subset(scobject, cells = Non_Tcells)

DimPlot(No_Tcells_Seurat, reduction = "umap")
```
Using these filters includes most cells from the T cell clusters and seems to be good in excluding other contaminating cell types. 

```{r Integration and Clustering}
All.data <- list.dirs("data/", recursive = F)
list.10x <- list()
counter <- 1
for(i in All.data){
  rnaseq.data <- Read10X_h5(paste0(i, "/count/sample_filtered_feature_bc_matrix.h5"))
  pbmc <- CreateSeuratObject(counts = rnaseq.data, project = "Colitis", min.cells = 3, 
                             min.features = 200)
  pbmc <- RenameCells(pbmc, new.names = paste0(colnames(pbmc), "_", counter))
  pbmc <- subset(pbmc, cells = All_Tcells_comb)
  Tcells <- WhichCells(pbmc, expression = CD3D > 0 | CD3E > 0 | CD247 > 0|
                   CD2 > 0 |  CD5 > 0 | PTPRC > 0, slot = "counts")
  Epithelial_Cells <- WhichCells(pbmc, expression = KRT8 > 0 | KRT86 > 1, slot = "counts")
  Monocytes <- WhichCells(pbmc, expression = CD14 > 0 | CD68 > 0, slot = "counts")
  APCs <- WhichCells(pbmc, expression = `HLA-DQA1` > 1, slot = "counts")
  Tcells_filtered <- setdiff(Tcells, Epithelial_Cells)
  Tcells_filtered <- setdiff(Tcells_filtered, Monocytes)
  Tcells_filtered <- setdiff(Tcells_filtered, APCs)
  pbmc <- subset(pbmc, cells = Tcells_filtered)
  pbmc@meta.data$percent.mt <- PercentageFeatureSet(pbmc, pattern = "^MT")
  pbmc@meta.data$percent.rps <- PercentageFeatureSet(pbmc, pattern = "^RPS")
  pbmc@meta.data$percent.rpl <- PercentageFeatureSet(pbmc, pattern = "^RPL")
 
  pbmc@meta.data$tp <- substr(i, 3, 100)
  pbmc <- SCTransform(pbmc, verbose = FALSE)
  list.10x[[i]] <- pbmc
  counter <- counter + 1
}

integ_features <- SelectIntegrationFeatures(object.list = list.10x, nfeatures = 3000) 
scTcells <- merge(x = list.10x[[1]], y = c(list.10x[2:length(list.10x)]), merge.data = T)
DefaultAssay(scTcells) <- "SCT"
VariableFeatures(scTcells) <- integ_features
scTcells <- RunPCA(scTcells, assay = "SCT", npcs = 50)

scTcells <- IntegrateLayers(object = scTcells, method = CCAIntegration, 
                            normalization.method = "SCT", verbose = F)
scTcells <- FindNeighbors(scTcells, reduction = "integrated.dr", dims = 1:30)
scTcells <- FindClusters(scTcells, resolution = c(0.4,0.6,0.8))

scTcells <- RunUMAP(scTcells, reduction = "integrated.dr", assay = "SCT", 
                             dims = 1:30)

DimPlot(scTcells, reduction = "umap", group.by = "SCT_snn_res.0.8", label = T, cols = farbenblind15, 
        label.size = 6)
DimPlot(scTcells, reduction = "umap", group.by = "SCT_snn_res.0.6", label = T, cols = farbenblind15,
        label.size = 6)
DimPlot(scTcells, reduction = "umap", group.by = "SCT_snn_res.0.4", label = T, cols = farbenblind15,
        label.size = 6)

DimPlot(scTcells, reduction = "umap", group.by = "tp")
DimPlot(scTcells, reduction = "umap", group.by = "tp", split.by = "tp", ncol = 3)
```

On a first look, the resolution of 0.6 looks best. This resolution will be examined first to see whether it gives a good resoltuion for annotating T cell subsets. 

```{r Cell population identification}
Idents(scTcells) <- scTcells$SCT_snn_res.0.6

scTcells <- PrepSCTFindMarkers(scTcells)
Tcell_markers_06 <- FindAllMarkers(scTcells, logfc.threshold = 0.25, min.pct = 0.25, 
                                   only.pos = T)

FeaturePlot(scTcells, reduction = "umap", features = "PTPRC")
FeaturePlot(scTcells, reduction = "umap", features = "CD2")
FeaturePlot(scTcells, reduction = "umap", features = "CD5")
FeaturePlot(scTcells, reduction = "umap", features = "CD3D")
FeaturePlot(scTcells, reduction = "umap", features = "CD3E")
FeaturePlot(scTcells, reduction = "umap", features = "CD3G")
FeaturePlot(scTcells, reduction = "umap", features = "CD8A")
FeaturePlot(scTcells, reduction = "umap", features = "CD8B")
FeaturePlot(scTcells, reduction = "umap", features = "CD4")
```

Clusters 0, 2, 5, and 10 mostly consist of CD8+ T cells and Clusters 1, 3, 6, 7, and 8 probably mostly contain CD4+ T cells. With Cluster 4 and 9 this is less obvious. 


```{r Cluster 0}
VlnPlot(scTcells, features = "CCL5", cols = farbenblind15)
VlnPlot(scTcells, features = "CD160", cols = farbenblind15)
VlnPlot(scTcells, features = "ITGAE", cols = farbenblind15)
VlnPlot(scTcells, features = "TRBV27", cols = farbenblind15)
```
Cluster 0 is likely a tissue-resident memory cluster. 
Based on ITGAE (CD103) expression, cluster 10 could also contain TRM cells. 

```{r Identify Tregs}
VlnPlot(scTcells, features = "IL1R1", cols = farbenblind15)
VlnPlot(scTcells, features = "IL1R2", cols = farbenblind15)
VlnPlot(scTcells, features = "FOXP3", cols = farbenblind15)
VlnPlot(scTcells, features = "BATF", cols = farbenblind15)
VlnPlot(scTcells, features = "CTLA4", cols = farbenblind15)
```
Clusters 3 and 8 are likely regulatory T cells. 

```{r Identify TCM}
VlnPlot(scTcells, features = "KLF2", cols = farbenblind15)
VlnPlot(scTcells, features = "BACH2", cols = farbenblind15)
VlnPlot(scTcells, features = "IL7R", cols = farbenblind15)
VlnPlot(scTcells, features = "TCF7", cols = farbenblind15)
```

Cluster 1 could be central memory T cells. But also clusters 5-7 and 9 partially express TCM marker genes. 

```{r Identify TPH}
VlnPlot(scTcells, features = "TNFSF8", cols = farbenblind15)
VlnPlot(scTcells, features = "ICOS", cols = farbenblind15)
VlnPlot(scTcells, features = "CXCR4", cols = farbenblind15)
```
Cluster 6 shows expression of TNFSF8, together with ICOS and CXCR4. This could be TPH cells, however more marker genes should be examined in the list of differentially expressed genes between clusters. 

```{r GzmK}
VlnPlot(scTcells, features = "GZMK", cols = farbenblind15)
VlnPlot(scTcells, features = "PRF1", cols = farbenblind15)
VlnPlot(scTcells, features = "CCL4", cols = farbenblind15)
```
Cluster 5 expresses high levels of GzmK, in contrast Cluster 9 expresses much Perforin. CCL4 is also increased in Cluster 5, which fits with the GzmK+ T cell annotation. 

```{r }
VlnPlot(scTcells, features = "FYN", cols = farbenblind15)
VlnPlot(scTcells, features = "SIK3", cols = farbenblind15)
VlnPlot(scTcells, features = "MALAT1", cols = farbenblind15)
```

Cluster 4 expressed mostly FYN, SIK3 and MALAT1. 

```{r Identify Th17}
VlnPlot(scTcells, features = "IL4I1", cols = farbenblind15)
VlnPlot(scTcells, features = "RORA", cols = farbenblind15)
VlnPlot(scTcells, features = "RORC", cols = farbenblind15)
VlnPlot(scTcells, features = "CCL20", cols = farbenblind15)
VlnPlot(scTcells, features = "IL17A", cols = farbenblind15)
```
Cluster 7 seem like Th17 differentiated T helper cells. 

```{r Treg}
VlnPlot(scTcells, features = "FOXP3", cols = farbenblind15)
VlnPlot(scTcells, features = "CD177", cols = farbenblind15)
VlnPlot(scTcells, features = "CCR8", cols = farbenblind15)
VlnPlot(scTcells, features = "IKZF2", cols = farbenblind15)
VlnPlot(scTcells, features = "TNFRSF4", cols = farbenblind15)
VlnPlot(scTcells, features = "TNFRSF18", cols = farbenblind15) #GITR
```
Cluster 3 and 8 are both Tregs, however, they show distinct marker gene expression. 

```{r Cluster 2}
VlnPlot(scTcells, features = "KLF6", cols = farbenblind15)
VlnPlot(scTcells, features = "FOS", cols = farbenblind15)
VlnPlot(scTcells, features = "JUNB", cols = farbenblind15)
VlnPlot(scTcells, features = "CD69", cols = farbenblind15)
```

```{r Check NK cells}
VlnPlot(scTcells, features = "TYROBP", cols = farbenblind15)
VlnPlot(scTcells, features = "FCER1G", cols = farbenblind15)
VlnPlot(scTcells, features = "KLRC1", cols = farbenblind15)
```
```{r Proliferation}
VlnPlot(scTcells, features = "MKI67", cols = farbenblind15)
VlnPlot(scTcells, features = "CDK1", cols = farbenblind15)
VlnPlot(scTcells, features = "TYMS", cols = farbenblind15)
```


Preliminary Annotation: 
Cluster 0 - TRM
Cluster 1 - TCM
Cluster 2 - Activated T cells?
Cluster 3 - Treg
Cluster 4 - Undefined 
Cluster 5 - GzmK+ T cells
Cluster 6 - TPH?
Cluster 7 - Th17?
Cluster 8 - Treg IKZF2+
Cluster 9 - NK cells? 
Cluster 10 - Proliferating T cells 


```{r Rename Idents}
scTcells <- RenameIdents(scTcells, '0' = 'CD8+ TRM', '1' = 'TCM', '2' = 'Activated T cells?', '3' = 'Treg', '4' = 'Undefined', 
                         '5' = 'CD8+ GzmK+', '6' = 'TPH?', '7' = 'Th17', '8' = 'IKZF2+ Treg', '9' = 'NK/NKT cells', 
                         '10' = 'Proliferating T')

scTcells$Annotation <- Idents(scTcells)
Cluster_over_time <- scTcells@meta.data %>% group_by(tp, Annotation) %>% 
  summarise(n = n()) %>% mutate(Freq = n/sum(n))

ggplot(Cluster_over_time, aes(x = tp, y = Freq, fill = Annotation)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15)
```


```{r Marker figures, fig.height=12, fig.width=10}
# Basic Heatmap with T cell marker genes 
top5_genes <- Tcell_markers_06 %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 5) %>% ungroup()

DoHeatmap(scTcells, features = unique(top5_genes$gene), assay = "SCT", slot = "scale.data",
          group.colors = farbenblind15) + NoLegend()

DotPlot(scTcells, features = unique(top5_genes$gene), group.by = "SCT_snn_res.0.6", 
        assay = "SCT") + coord_flip()

saveRDS(scTcells, file = "Tcell_subclustering_CCA_Integration_annotated.rds")
```
