---
title: "05_Tcell_Cytokine_Expression"
author: "Julia Hitschfel"
date: "2026-05-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(Seurat)
library(SeuratObject)
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(colorspace)
library(RColorBrewer)
library(viridis)
library(pheatmap)
```


## T Cell Effector molecule expression

Here the expression of canonical T cell cytokines and chemokines will be examined to better identify the T cell subpopulations in the subclustering analysis. 

```{r Import dataset}
Tcell_scobject <- readRDS(file = "Tcell_subclustering_CCA_Integration_annotated.rds")

Effector_list <- c("IL1A", "IL1B","IL2", "IL4", "IL5", "IL6", "IL7", "CXCL8", "IL10", "IL12A", "TXLNA",
                   "IL15", "IL16", "IL17A", "IL18", "IL19", "IL21", "IL22", "IL26", "EBI3", "IL32",
                   "IL33", "GZMB", "PRF1", "GNLY", "GZMA", "GZMK", "TGFB1", "IFNG", "CXCL10",
                   "CXCL13", "CCL4", "CCL5", "TNF", "LTA", "XCL1", "XCL2")

DoHeatmap(Tcell_scobject, features = unique(Effector_list), assay = "SCT", slot = "scale.data") +
  NoLegend()
```

Some of the features are not found in the SCT Assay, so expression will be plotted from the normalized and scaled RNA assay data: 

```{r Scale RNA Data for Heatmap}
DefaultAssay(Tcell_scobject) <- "RNA"
Tcell_scobject <- NormalizeData(Tcell_scobject)
Tcell_scobject <- ScaleData(Tcell_scobject, features = rownames(Tcell_scobject))

DoHeatmap(Tcell_scobject, features = unique(Effector_list), assay = "RNA", slot = "scale.data") +
  NoLegend()
```

Many cytokines are lowly expressed but can be detected in some of the clusters. 
Interleukin-1A and 1B, and Interleukin-2 only show weak expression. 
Interleukin-3 was not detected in the data. 
Interleukin-4, -5 and -6 also barely show expression in the T cell clusters. 
CXCL8 shows partial expression in Cluster 1.
Interleukin-9 was not detected in T cells, as was Interleukin-13.
TXLNA (IL14) was only expressed in very few cells.
EBI3 and IL27p28 are the genes encoding for IL-27. Only EBI3 expression is weakly detected. 
IL28 and IL29 (IFN lambda) are not expressed in this dataset.

The highest cytokine expression can be seen with IL7, IL10, IL16, IL17A, and IL26. 
IL-32 also shows increased expression, as does IL-16.

```{r Cytokines as Heatmap, fig.height=10, fig.width=10}
pseudo_Tcells <- AggregateExpression(Tcell_scobject, assays = "RNA", return.seurat = T, 
                                     group.by = "ident")

DoHeatmap(pseudo_Tcells, features = unique(Effector_list), assay = "RNA", slot = "scale.data") +
  NoLegend()

Effector_gene_rows <- which(rownames(pseudo_Tcells@assays$RNA$scale.data) %in% Effector_list)
Effector_genes <- pseudo_Tcells@assays$RNA$scale.data[Effector_gene_rows,]

pheatmap(Effector_genes)

pheatmap(Effector_genes, color = viridis(100))
```

```{r Cluster over time}
Cluster_over_time <- Tcell_scobject@meta.data %>% group_by(tp, Annotation) %>% 
  summarise(n = n()) %>% mutate(Freq = n/sum(n))

farbenblind15 <- c("#000000","#004949","#009292","#FF6DB6","#FFB6DB",
                  "#490092","#006DDB","#B66DFF","#6DB6FF","#B6DBFF",
                  "#920000","#924900","#DB6D00","#00A651","#FFFF6D") 

ggplot(Cluster_over_time, aes(x = tp, y = Freq, fill = Annotation)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15)
```
CD8+ TRM are increased after therapy. They decline a little bit after the last timepoint, but still remain elevated.
CD8+ GzmK+ T cells also expand over time. They produce GzmK, IL10, IL19 and IL12A, which could be associated with a tissue repair program.
CD8+ TCM cells show a constant frequency over time and express little cytokines except for some IL16. 
Activated T cells also remain constant over time. They express IL2 and TNFa, as well as IFNg, which fits with their annotation as activated cells. 
Regulatory T cells decline over time. They produce IL-10 - a classical Treg marker - but also CXCL8, IL17A, IL26, LTA and EBI3, perhaps indicating a pathogenic potential. 
The second IKZF2+ Tregs (or CCR8+ Tregs) continue to decline, they show weaker expression of IL-10 but express inflammation-associated cytokines.  
Annotated TPH cells show expressing of CXCL13 and IL-21, which supports their annotation as TPH cells since these are classical B cell help associated cytokines. 
Th17 T cells also remain at constant frequencies and express IL15, IL17A, IL22 and IL26, supporting their annotation. 
The undefined population shows little cytokine expression, only some CXCL10 and IL4. IL4 is associated with Th2 cells, while CXCL10 is not usually associated with T cells. These could be Quiescent T cells or Memory T cells without much activation or differentiation. 
NK/NKT cells show a decline, as do the proliferating T cells. 

