---
title: "08_Epithelial_Cell_Annotation"
author: "Julia Hitschfel"
date: "2026-05-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(Seurat)
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(BPCells)
library(presto)
library(glmGamPoi)
library(hdf5r)
library(knitr)
library(pheatmap)
library(RColorBrewer)

farbenblind15 <- c("#000000","#004949","#009292","#FF6DB6","#FFB6DB",
                  "#490092","#006DDB","#B66DFF","#6DB6FF","#B6DBFF",
                  "#920000","#924900","#DB6D00","#00A651","#FFFF6D") 
```

## Violin Plots for marker genes of epithelial cell subpopulations 


```{r Import Data}
scEpithelial <- readRDS(file = "Epithelial_Cells.rds")

Idents(scEpithelial) <- scEpithelial@meta.data$SCT_snn_res.0.4

scEpithelial <- PrepSCTFindMarkers(scEpithelial)

Cluster_marker_04 <- FindAllMarkers(scEpithelial, logfc.threshold = 0.25, min.pct = 0.25, 
                                    only.pos = T)

write.csv(Cluster_marker_04, file = "Markergenes_Subpopulations/Epithelial_cells_res04.csv")
```

### Intestinal stem cells 

```{r Intestinal stem cells}
VlnPlot(scEpithelial, features = "LGR5")

VlnPlot(scEpithelial, features = "ASCL2")

VlnPlot(scEpithelial, features = "VDR")

VlnPlot(scEpithelial, features = "OLFM4")

VlnPlot(scEpithelial, features = "SMOC2")

VlnPlot(scEpithelial, features = "SOX9")
```
Cluster 3 shows slightly elevated expression of intestinal stem cell marker genes LGR5 and ASCL2 and an increased expression of OLFM4, which supports an early differentiation stage. 

### Goblet Cell marker genes 

```{r Goblet cell marker}
VlnPlot(scEpithelial, features = "TFF3")

VlnPlot(scEpithelial, features = "ITLN1")

VlnPlot(scEpithelial, features = "SPINK4")

```

Cluster 0 shows the highest expression of TFF3 and thus is likely a fully differentiated goblet cell population. In addition, cluster 7 and 9 show increased expression of TFF3 and could be goblet cells or precursors thereof.  


### INFLARE marker genes from PanGI Paper

```{r INFLAREs}
VlnPlot(scEpithelial, features = "MUC6")

VlnPlot(scEpithelial, features = "MUC5AC")

VlnPlot(scEpithelial, features = "BPIFB1")

VlnPlot(scEpithelial, features = "PGC")

VlnPlot(scEpithelial, features = "AQP5")

```

Cluster 3 contains cells that were described to be associated with IBD in the PanGI atlas paper of Sarah Teichmann's group.The paper also described these cells as having a stem cell phenotype, which would be fitting with the potential stem cell identity of this cluster. 

### Absorptive epithelial cell marker genes 

```{r Absorptive}
VlnPlot(scEpithelial, features = "FABP1")

VlnPlot(scEpithelial, features = "CDHR5")

VlnPlot(scEpithelial, features = "PHGR1")

VlnPlot(scEpithelial, features = "ANPEP")

VlnPlot(scEpithelial, features = "IL32")
```

Cluster 2 shows an increased expression of marker genes that are associated with fully differentiated absorptive enterocytes/colonocytes. Cluster 5 shows moderate expression of these markers and perhaps thus represents an earlier differentiation state. 

### Transit Amplifying Cells

```{r Transit Amplifying}
VlnPlot(scEpithelial, features = "MKI67")

VlnPlot(scEpithelial, features = "STMN1")

VlnPlot(scEpithelial, features = "PCLAF")

VlnPlot(scEpithelial, features = "PTTG1")

VlnPlot(scEpithelial, features = "RARRES2")

VlnPlot(scEpithelial, features = "MYB")

VlnPlot(scEpithelial, features = "AQP1")

```

There seem to be three clusters of cycling epithelial cells. Cluster 6-8 all show expression of proliferation markers that are associated with Transit Amplifying Cells. Since cluster 7 also expresses some markers of secretory epithelial cells/Goblet cells, perhaps these are proliferating cells on a differentiation trajectory towards goblet cells. 

### Enteroendocrine

```{r Enteroendocrine}
VlnPlot(scEpithelial, features = "CHGA")

VlnPlot(scEpithelial, features = "PCSK1N")

VlnPlot(scEpithelial, features = "SCGN")

```

Cluster 10 represent a cluster of enteroendocrine cells. 

### Epithelial cell progenitor cells 

```{r Epithelial progenitor}
VlnPlot(scEpithelial, features = "APOE")

VlnPlot(scEpithelial, features = "APOA1")

VlnPlot(scEpithelial, features = "GPC3")
```

No population clearly expresses marker genes of progenitor cells. 

```{r Unidentified Clusters}
VlnPlot(scEpithelial, features = "KRT8")

VlnPlot(scEpithelial, features = "NME2")

VlnPlot(scEpithelial, features = "PIEZO1")

VlnPlot(scEpithelial, features = "S100A10")

VlnPlot(scEpithelial, features = "HES6")

# Cluster 8
VlnPlot(scEpithelial, features = "CCNB1")

VlnPlot(scEpithelial, features = "CDC20")

VlnPlot(scEpithelial, features = "CENPF")

# Cluster 9
VlnPlot(scEpithelial, features = "NRG1")

VlnPlot(scEpithelial, features = "CXCL14")

VlnPlot(scEpithelial, features = "COL1A2")

# Cluster 11
VlnPlot(scEpithelial, features = "PRSS2")

VlnPlot(scEpithelial, features = "DEFA5")

VlnPlot(scEpithelial, features = "PLA2G2A")

# Cluster 12 
VlnPlot(scEpithelial, features = "SPIB")

VlnPlot(scEpithelial, features = "BEST4")

VlnPlot(scEpithelial, features = "PROX1")

```

Preliminary Annotation: 

```{r Annotation}
Cluster_over_time_04 <- scEpithelial@meta.data %>% group_by(tp, SCT_snn_res.0.4) %>% 
  summarise(n = n()) %>% mutate(Freq = n/sum(n))

ggplot(Cluster_over_time_04, aes(x = tp, y = Freq, fill = SCT_snn_res.0.4)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15)

scEpithelial <- RenameIdents(scEpithelial, '0' = 'Goblet Cells', '1' = 'Stress-response epithelial cells', '2' = 'Mature Enterocytes', 
                             '3' = 'Intestinal Stem Cells', '4' = 'Injury-response epithelial cells', 
                             '5' = 'Enterocytes (early differentiated)',
                             '6' = 'Cycling Transit Amplifying 1', '7' = 'Immature Goblet Cells', '8' = 'Transit Amplifying 2', 
                             '9' = 'EMT-like secretory Epithelial', '10' = 'Enteroendocrine', '11' = 'Paneth-like Cells', 
                             '12' = 'BEST4+ Enterocytes')

scEpithelial$Annotation <- Idents(scEpithelial)

Cluster_over_time_04_ann <- scEpithelial@meta.data %>% group_by(tp, Annotation) %>% 
  summarise(n = n()) %>% mutate(Freq = n/sum(n))

ggplot(Cluster_over_time_04_ann, aes(x = tp, y = Freq, fill = Annotation)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15)

write.csv(Cluster_over_time_04_ann, 
          file = "Epithelial_Cell_subsets_in_epithelial.csv")

saveRDS(scEpithelial, file = "Epithelial_Cells_annotated.rds")

```
```{r Frequencies Epithelial in all cells}
All_cells_scobject <- readRDS(file = "integrated_seurat_cellcycle_regression_CU1_1year_res04_annotated.rds")

All_cells_TP <- All_cells_scobject@meta.data %>% group_by(tp) %>% summarise(All.n = n())
All_cells_TP

# Frequencies in All Cells
EpithelialCluster_All <- merge(Cluster_over_time_04_ann, All_cells_TP, by = "tp")
EpithelialCluster_All_Freq <- EpithelialCluster_All %>% mutate(Freq.All = (n/All.n)*100)
write.csv(EpithelialCluster_All_Freq, "Epithelial_cells_in_Total.csv")
```
