---
title: "11_Bcell_Receptor_Analysis"
author: "Julia Hitschfel"
date: "2026-05-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(SingleCellExperiment)
library(SingleR)
library(celldex)
library(Seurat)
library(SeuratObject)
library(dplyr)
library(ggplot2)
library(scRepertoire)

farbenblind15 <- c("#000000","#004949","#009292","#FF6DB6","#FFB6DB",
                  "#490092","#006DDB","#B66DFF","#6DB6FF","#B6DBFF",
                  "#920000","#924900","#DB6D00","#00A651","#FFFF6D")
```


## B cell receptor sequence analysis 

```{r Import data}
scobject.BCR <- readRDS(file = "Bcell_subclustering_BCRs_integrated_annotated.rds")

scobject <- readRDS(file = "integrated_seurat_cellcycle_regression_CU1_1year_res04_annotated.rds")

# Read in VDJ Libraries
all.tps <- list.dirs("data/", recursive = F)
contig.path <- "/vdj_b/filtered_contig_annotations.csv"
contig.list <- list()
counter <- 1

for(i in all.tps){
  vdj_b_contig <- read.csv(paste0(i, contig.path))
  contig.list[[counter]] <- vdj_b_contig
  counter <- counter + 1
}

combined.BCR <- combineBCR(contig.list, 
                           samples = c("T1",  "T2", "T3", "T4", "T5", "T6", "T7", "T8", "T9"),
                           removeNA = FALSE, 
                           removeMulti = FALSE, 
                           filterMulti = FALSE)

head(combined.BCR[[1]])
```

```{r Make BCs compatible in scobject}
old_BCs <- colnames(scobject@assays$RNA$counts)
old_BCs <- ifelse(grepl("_1", old_BCs) == T, paste0("T1_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_2", old_BCs) == T, paste0("T2_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_3", old_BCs) == T, paste0("T3_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_4", old_BCs) == T, paste0("T4_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_5", old_BCs) == T, paste0("T5_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_6", old_BCs) == T, paste0("T6_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_7", old_BCs) == T, paste0("T7_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_8", old_BCs) == T, paste0("T8_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_9", old_BCs) == T, paste0("T9_",old_BCs), old_BCs) 
new_BCs <- substr(old_BCs, 0, nchar(old_BCs) - 2)

scobject <- RenameCells(scobject, new.names = new_BCs)
```


## VDJ without GEX

```{r VDJ without GEX}
BCR.barcodes <- vector()
for(i in 1:length(combined.BCR)){
  barcodes <- combined.BCR[[i]]$barcode
  BCR.barcodes <- c(BCR.barcodes, barcodes)
}

VDJ.without.Bcell.GEX <- intersect(BCR.barcodes, rownames(scobject.BCR@meta.data))
length(VDJ.without.Bcell.GEX)

VDJ.without.total.GEX <- intersect(BCR.barcodes, rownames(scobject@meta.data))
length(VDJ.without.total.GEX)
```


```{r Add timepoints for VDJ analysis}
# Barcodes in contig.list do not have the addition of the timepoint, so we need to create that addition in the list to filter on cells with associated gene expression data: 

date_BCs <- colnames(scobject.BCR@assays$RNA$counts)

tps <- c("T1",  "T2", "T3", "T4", "T5", "T6",  "T7", "T8", "T9")

tp.to.exclude <- vector()

for(i in 1:length(contig.list)){
  contig_filter <- contig.list[[i]]
  contig_filter$barcode <- paste0(tps[i], "_", contig_filter$barcode)
  contig_filter <- subset(contig_filter, contig_filter$barcode %in% date_BCs)
  contig_filter$barcode <- substr(contig_filter$barcode, 4, 100)
  if(nrow(contig_filter) > 0){
  contig.list[[i]] <- contig_filter
  } else{
    print(paste0("Exclude timepoint ", i, " from BCR analysis."))
    tp.to.exclude[i] <- i 
  }
}

contig.list[tp.to.exclude] <- NULL

combined.BCR.GEX <- combineBCR(contig.list,
                           samples = c("T1",  "T2", "T3", "T4", "T5", "T6", "T7", "T8", "T9"),
                           removeNA = FALSE,
                           removeMulti = FALSE,
                           filterMulti = FALSE)

head(combined.BCR.GEX[[1]])
```


## Basic Clonal Analysis BCRs

```{r }
clonalQuant(combined.BCR.GEX, 
            cloneCall="strict", 
            chain = "both", 
            scale = TRUE)

clonalAbundance(combined.BCR.GEX, 
                cloneCall = "gene", 
                scale = FALSE)

clonalCompare(combined.BCR.GEX, 
                  top.clones = 10, 
                  samples = c("T1", "T2", "T3"), 
                  cloneCall="aa", 
                  graph = "alluvial")

clonalCompare(combined.BCR.GEX, 
                  top.clones = 10, 
                  samples = c("T3", "T4", "T5"), 
                  cloneCall="aa", 
                  graph = "alluvial")

clonalCompare(combined.BCR.GEX, 
                  top.clones = 10, 
                  samples = c("T5", "T7"), 
                  cloneCall="aa", 
                  graph = "alluvial")

clonalCompare(combined.BCR.GEX, 
                  top.clones = 10, 
                  samples = c("T7", "T8", "T9"), 
                  cloneCall="aa", 
                  graph = "alluvial")

clonalProportion(combined.BCR.GEX, 
                 cloneCall = "strict") 

percentAA(combined.BCR.GEX, 
          chain = "IGH", 
          aa.length = 25)

clonalDiversity(combined.BCR.GEX, 
                cloneCall = "strict")

clonalOverlap(combined.BCR.GEX, 
              cloneCall = "strict", 
              method = "raw")

clonalOverlap(combined.BCR.GEX, 
              cloneCall = "strict", 
              method = "morisita")
```

At the early timepoints there are a few B cell clones that can be detected at multiple timepoints. 
Following therapy, there are few overlaps between timepoints. 


```{r Clones in B cell clusters}
#Define color palette 
colorblind_vector <- hcl.colors(n=7, palette = "inferno", fixup = TRUE)

DimPlot(scobject.BCR, group.by = "cloneSize") +
    scale_color_manual(values=rev(colorblind_vector[c(1,3,5,7)]))

clonalOccupy(scobject.BCR, 
              x.axis = "Annotation")

clonalOccupy(scobject.BCR, 
             x.axis = "Annotation", 
             proportion = TRUE, 
             label = FALSE)

# Use the vegan package to calculate diversity across clusters:
BCR.meta <- scobject.BCR@meta.data %>% group_by(Annotation, CTstrict) %>% summarise(n = n()) %>%
  filter(!is.na(CTstrict))

BCR.meta.shannon <- BCR.meta %>% group_by(Annotation) %>% dplyr::mutate(shannon_index =
                                                                               vegan::diversity(n))

ggplot(BCR.meta.shannon, aes(x = Annotation, y = shannon_index)) + geom_boxplot() + ylim(0,8)
```

Expanded clonotypes are mostly present in the GC-like B cell cluster and also within cluster Plasmacells_1. 
This makes biologic sense, as B cells in germinal centers proliferate following activation through their BCR. 
Whether these B cell clones are reactive against colitis-associated antigens would need further evaluation. 

```{r Isotypes over time}
VlnPlot(scobject.BCR, features = "IGHG1", group.by = "date") + 
  scale_fill_manual(values = farbenblind15, drop = FALSE)

VlnPlot(scobject.BCR, features = "IGHG2", group.by = "date") + 
  scale_fill_manual(values = farbenblind15, drop = FALSE)

VlnPlot(scobject.BCR, features = "IGHG4", group.by = "date") + 
  scale_fill_manual(values = farbenblind15, drop = FALSE)

VlnPlot(scobject.BCR, features = "IGHA1", group.by = "date") + 
  scale_fill_manual(values = farbenblind15, drop = FALSE)

VlnPlot(scobject.BCR, features = "IGHA2", group.by = "date") + 
  scale_fill_manual(values = farbenblind15, drop = FALSE)
```

IGHG1 expressing B cells vanishing following therapy and also seem to not come back at the early timepoints of B cell reconstitution. 
IGHA1 expressing cells are also depleted, however, there are a few B cells remaining that express this isotype and these are also earlier to replenish the repertoire. 

### Isotypes on T8

```{r }
unique(combined.BCR.GEX$`T8`$CTgene)
```


### Expanded Clones on T7

```{r B cell receptor expansion}
length(unique(combined.BCR.GEX$`T7`$CTstrict))

nrow(combined.BCR.GEX$`T7`)
```


### Expanded Clones on T5

```{r Expanded clones T5}
length(unique(combined.BCR.GEX$`T5`$CTstrict))

nrow(combined.BCR.GEX$`T5`)
```


### Expanded Clones on T4

```{r Expanded clones d10}
length(unique(combined.BCR.GEX$`T4`$CTstrict))

nrow(combined.BCR.GEX$`T4`)

print("Potentially expanded B cell receptor clones on T4:")

combined.BCR.GEX$`T4`$CTstrict[which(duplicated(combined.BCR.GEX$`T4`$CTstrict))]

combined.BCR.GEX$`T4`$CTaa[which(duplicated(combined.BCR.GEX$`T4`$CTstrict))]
```


### Expanded Clones on T3

```{r Expanded clones pre CAR T }
length(unique(combined.BCR.GEX$`T3`$CTstrict))

nrow(combined.BCR.GEX$`T3`)

print("Potentially expanded B cell receptor clones on T3:")

combined.BCR.GEX$`T3`$CTstrict[which(duplicated(combined.BCR.GEX$`T3`$CTstrict))]

combined.BCR.GEX$`T3`$CTaa[which(duplicated(combined.BCR.GEX$`T3`$CTstrict))]
```


### Expanded Clones on T2

```{r Expanded clones post T2}
length(unique(combined.BCR.GEX$`T2`$CTstrict))

nrow(combined.BCR.GEX$`T2`)

print("Potentially expanded B cell receptor clones on T2:")

unique(combined.BCR.GEX$`T2`$CTstrict[which(duplicated(combined.BCR.GEX$`T2`$CTstrict))])

unique(combined.BCR.GEX$`T2`$CTaa[which(duplicated(combined.BCR.GEX$`T2`$CTstrict))])
```


### Expanded Clones on T1

```{r Expanded clones pre Blina}
length(unique(combined.BCR.GEX$`T1`$CTstrict))

nrow(combined.BCR.GEX$`T1`)

print("Potentially expanded B cell receptor clones on T1:")

unique(combined.BCR.GEX$`T1`$CTstrict[which(duplicated(combined.BCR.GEX$`T1`$CTstrict))])

unique(combined.BCR.GEX$`T1`$CTaa[which(duplicated(combined.BCR.GEX$`T1`$CTstrict))])
```
