---
title: "12_Myeloid_Cell_Subclustering"
author: "Julia Hitschfel"
date: "2026-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(Seurat)
library(SeuratObject)
library(dplyr)
library(ggplot2)
library(colorspace)
library(RColorBrewer)
library(SingleR)
library(celldex)

farbenblind15 <- c("#000000","#004949","#009292","#FF6DB6","#FFB6DB",
                  "#490092","#006DDB","#B66DFF","#6DB6FF","#B6DBFF",
                  "#920000","#924900","#DB6D00","#00A651","#FFFF6D")
```


## Subclustering myeloid cells 

```{r Import data and explore marker genes}

scobject <- readRDS(file = "integrated_seurat_cellcycle_regression_CU1_1year_res04_annotated.rds")

DimPlot(scobject, reduction = "umap", label = T)
scobject$Annotation <- Idents(scobject)
DimPlot(scobject, reduction = "umap", group.by = "Annotation", label = T)

# Top markers in myeloid cell Cluster (Macs): TYROBP, AIF1, IFI30, C1QA, FCER1G, HLA class II, CD68, MS4A7, LYZ, MS4A6A, IL1B

# Top marker in myeloid cell Cluster (DCs): CORO1A, HLA-DQA1, TMSB4X, PFN1, RPL and RPS transkripte, LCP1, CD52

All_Myeloid_comb <- WhichCells(scobject, idents = c("Myeloid cells (Macs)", "Myeloid cells (DCs)"))

FeaturePlot(scobject, features = "CD14", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "FCER1G", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "LYZ", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "IFI30", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "TYROBP", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "HLA-DPB1", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "CD68", slot = "counts", max.cutoff = 5)

FeaturePlot(scobject, features = "CORO1A", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "HLA-DQA1", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "TMSB4X", slot = "counts", max.cutoff = 35)
FeaturePlot(scobject, features = "LCP1", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "CD52", slot = "counts", max.cutoff = 5)
```

The expression of CORO1A is not a reliable marker for macrophages and myeloid cells. Also expression of CD52 and LCP1 is not exclusive to myeloid cells and thus should not be used for filtering. 
Lysozyme is strongly expressed in myeloid cells, but also shows some expression in intestinal epithelial cells, which suggests that it is not a definitive marker. FCER1G, IFI30, TYROBP, CD68, and also HLA class II are primarily restricted to myeloid cell clusters and seem to be good inclusion filters for myeloid cells. 

```{r Check filters}
genes_to_filter <- c('FCER1G', 'LYZ', 'IFI30', 'HLA-DPB1')
Non_Myeloid_Cells <- WhichCells(scobject, expression = FCER1G < 3 & LYZ < 3 & IFI30 < 2 &
                                     `HLA-DPB1` < 3 & TYROBP < 2, slot = "counts")

Non_Myeloid_Seurat <- subset(scobject, cells = Non_Myeloid_Cells)

DimPlot(Non_Myeloid_Seurat, reduction = "umap")

#Classical monocyte markers
FeaturePlot(Non_Myeloid_Seurat, features = "CD14")
FeaturePlot(Non_Myeloid_Seurat, features = "CD68")
FeaturePlot(Non_Myeloid_Seurat, features = "FCGR1A")

#Classical macrophage markers
FeaturePlot(Non_Myeloid_Seurat, features = "ITGAM")
FeaturePlot(Non_Myeloid_Seurat, features = "FCGR3A")
FeaturePlot(Non_Myeloid_Seurat, features = "CCR5")
```

To further exclude immune cell-"contamination" - especially B cells - the inclusion filters for B cells will now be used here as exclusion filters: MS4A1 (CD20) >= 1 count, IGHM >= 1 count, MZB1 >= 1 count, CD19 >= 1 count and CD79A >= 1 count.
T cells will be excluded via CD3D and CD3E >= 1 UMI counts. 

```{r Read in data for myeloid cells}
All.data <- list.dirs("data/", recursive = F)
list.10x <- list()
counter <- 1
for(i in All.data){
  rnaseq.data <- Read10X_h5(paste0(i, "/count/sample_filtered_feature_bc_matrix.h5"))
  all.cells <- CreateSeuratObject(counts = rnaseq.data, project = "Colitis", min.cells = 3, 
                             min.features = 200)
  all.cells <- RenameCells(all.cells, new.names = paste0(colnames(all.cells), "_", counter))
  monos <- subset(all.cells, cells = All_Myeloid_comb)
  list.10x[[i]] <- monos
  counter <- counter + 1
}


print("Myeloid Cell Numbers before Filtering:")

for(i in list.10x){print(nrow(i@meta.data))}
```
```{r Filter for myeloid cells}
for(i in 1:length(list.10x)){  
  monos <- list.10x[[i]]
  monos@meta.data$tp <- substr(names(list.10x)[i], 6, 100)
  Myeloid_Cells <- WhichCells(monos, expression = FCER1G > 2 | LYZ > 2 | IFI30 > 1 | `HLA-DPB1` > 2 | 
                                TYROBP > 1, slot = "counts")
  Bcells <- vector()
  if("MS4A1" %in% rownames(monos@assays$RNA$counts)){
    MS4A1_Bcells <- WhichCells(monos, expression = MS4A1 > 0, slot = "counts")
  }
    if(exists("Bcells")){Bcells <- c(Bcells, MS4A1_Bcells)}
  
  if("IGHM" %in% rownames(monos@assays$RNA$counts)){
    IGHM_Bcells <- WhichCells(monos, expression = IGHM > 0, slot = "counts")
  }
   if(exists("Bcells")){Bcells <- c(Bcells, IGHM_Bcells)}
  
  if("MZB1" %in% rownames(monos@assays$RNA$counts)){
    MZB1_Bcells <- WhichCells(monos, expression = MZB1 > 0, slot = "counts")
  }
   if(exists("Bcells")){Bcells <- c(Bcells,  MZB1_Bcells)}
  
  if("CD19" %in% rownames(monos@assays$RNA$counts)){
    CD19_Bcells <- WhichCells(monos, expression = CD19 > 0, slot = "counts")
  }
   if(exists("Bcells")){Bcells <- c(Bcells,  CD19_Bcells)}
  
  if("CD79A" %in% rownames(monos@assays$RNA$counts)){
    CD79A_Bcells <- WhichCells(monos, expression = CD79A > 0, slot = "counts")
  }
  if(exists("Bcells")){Bcells <- c(Bcells,  CD79A_Bcells)}
  
  Tcells <- WhichCells(monos, expression = CD3D > 0 | CD3E > 0, slot = "counts")
  
  Myeloid_Cells <- setdiff(Myeloid_Cells, Bcells)
  Myeloid_Cells <- setdiff(Myeloid_Cells, Tcells)
  monos <- subset(monos, cells = Myeloid_Cells)
  monos@meta.data$percent.mt <- PercentageFeatureSet(monos, pattern = "^MT")
  monos@meta.data$percent.rps <- PercentageFeatureSet(monos, pattern = "^RPS")
  monos@meta.data$percent.rpl <- PercentageFeatureSet(monos, pattern = "^RPL")
  monos <- SCTransform(monos, verbose = FALSE)
  list.10x[[i]] <- monos
}

print("Myeloid Cell Numbers after Filtering:")

for(i in list.10x){print(nrow(i@meta.data))}
```

At the early timepoints, the filters remove a number of cells, which likely are B cells. After administration of treatment, only few cells are filtered our from the myeloid cell subclustering analysis. 

CCA Integration workflow does not work here, as there are fewer than 100 cells in the first sample. Instead, RPCA Integration workflow will be used here, which allows finer adjustment of k.weights. 
The RPCA workflow documentation can be found here: https://satijalab.org/seurat/articles/integration_rpca.html

```{r RPCA workflow for integration}
features <- SelectIntegrationFeatures(object.list = list.10x, nfeatures = 3000)
list.10x <- PrepSCTIntegration(object.list = list.10x, anchor.features = features)
list.10x <- lapply(X = list.10x, FUN = RunPCA, features = features)

immune.anchors <- FindIntegrationAnchors(object.list = list.10x, normalization.method = "SCT",
    anchor.features = features, dims = 1:20, reduction = "rpca", k.anchor = 20)
scMyeloid <- IntegrateData(anchorset = immune.anchors, normalization.method = "SCT", 
                                     dims = 1:20, k.weight = 62)

scMyeloid <- RunPCA(scMyeloid, verbose = FALSE)
scMyeloid <- RunUMAP(scMyeloid, reduction = "pca", dims = 1:20)

# Visualization
p1 <- DimPlot(scMyeloid, reduction = "umap", group.by = "tp")
p1 


scMyeloid <- FindNeighbors(scMyeloid, reduction = "pca", dims = 1:20)
scMyeloid <- FindClusters(scMyeloid, resolution = c(0.2, 0.4, 0.6))

DimPlot(scMyeloid, reduction = "umap", label = T, group.by = "integrated_snn_res.0.4",
        cols = farbenblind15, label.size = 6)

DimPlot(scMyeloid, reduction = "umap", group.by = "integrated_snn_res.0.2", label = T,
        cols = farbenblind15, label.size = 6)

DimPlot(scMyeloid, reduction = "umap", group.by = "tp", split.by = "tp", ncol = 3)
```

```{r Examine QC metrics}
Idents(scMyeloid) <- "integrated_snn_res.0.4"
sum_table <- scMyeloid@meta.data %>% group_by(tp) %>% 
  summarise(RNA.count.median = median(nCount_RNA), RNA.feature.median = median(nFeature_RNA), 
            MT.freq.median = median(percent.mt), RPS.freq.median = median(percent.rps), 
            RPL.freq.median = median(percent.rpl))

sum_table

DimPlot(scMyeloid, reduction = "pca") + NoLegend()

FeaturePlot(scMyeloid, reduction = "pca", features = "percent.mt")
FeaturePlot(scMyeloid, reduction = "pca", features = "percent.rps")
FeaturePlot(scMyeloid, reduction = "pca", features = "percent.rpl")
FeaturePlot(scMyeloid, features = "nCount_RNA", reduction = "pca", max.cutoff = 5000) 
DimHeatmap(scMyeloid, dims = 1:15, cells = 500, balanced = TRUE)
```

```{r Examine cluster over time}
# Cluster Resolution 0.4
table(scMyeloid@meta.data$tp, scMyeloid@meta.data$integrated_snn_res.0.4)

Cluster_over_time_04 <- scMyeloid@meta.data %>% group_by(tp, integrated_snn_res.0.4) %>% 
  summarise(n = n()) %>% mutate(Freq = n/sum(n))

ggplot(Cluster_over_time_04, aes(x = tp, y = Freq, fill = integrated_snn_res.0.4)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15)

# Cluster Resolution 0.2
table(scMyeloid@meta.data$tp, scMyeloid@meta.data$integrated_snn_res.0.2)

Cluster_over_time_04 <- scMyeloid@meta.data %>% group_by(tp, integrated_snn_res.0.2) %>% 
  summarise(n = n()) %>% mutate(Freq = n/sum(n))

ggplot(Cluster_over_time_04, aes(x = tp, y = Freq, fill = integrated_snn_res.0.2)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15)
```

```{r Find markers, fig.height=12, fig.width=8}
scMyeloid <- JoinLayers(scMyeloid, assay = "RNA", layers = "counts")

# For better visualization, normalization and scaling will be performed here
scMyeloid <- NormalizeData(scMyeloid, assay = "RNA")
scMyeloid <- ScaleData(scMyeloid, assay = "RNA")

DefaultAssay(scMyeloid) <- "RNA"

# Cluster Resolution 0.4
Cluster_marker_04 <- FindAllMarkers(scMyeloid, logfc.threshold = 0.25, min.pct = 0.25, 
                                    only.pos = T)

write.csv(Cluster_marker_04, file = "Myeloid_cells_marker_genes_res04.csv")

top10_genes <- Cluster_marker_04 %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 10) %>% ungroup()

DoHeatmap(scMyeloid, features = unique(top10_genes$gene), assay = "RNA", slot = "scale.data") +
  NoLegend()

DotPlot(scMyeloid, features = unique(top10_genes$gene), group.by = "integrated_snn_res.0.4") +
  coord_flip()

Cluster_marker_04_noRPS <- Cluster_marker_04[which(grepl("^RPS*", Cluster_marker_04$gene) == F),]
Cluster_marker_04_noRPL <- Cluster_marker_04_noRPS[which(grepl("^RPL*", Cluster_marker_04_noRPS$gene) == F),]
top10_genes_noRPS <- Cluster_marker_04_noRPL %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 10) %>% ungroup()

DoHeatmap(scMyeloid, features = unique(top10_genes_noRPS$gene), assay = "RNA", slot = "scale.data") +
  NoLegend() + labs(title = "Heatmap without RPS transcripts")

DotPlot(scMyeloid, features = unique(top10_genes_noRPS$gene), group.by = "integrated_snn_res.0.4") +
  coord_flip() + labs(title = "Dotplot without RPS transcripts")

# Top 50 Marker without RPS/RPL
top50_genes_04 <- Cluster_marker_04_noRPL %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 50) %>% ungroup()

write.csv(top50_genes_04, file = "Myeloid_cells_top50_marker_genes_res04.csv")

# Cluster Resolution 0.2
Idents(scMyeloid) <- "integrated_snn_res.0.2"
Cluster_marker_02 <- FindAllMarkers(scMyeloid, logfc.threshold = 0.25, min.pct = 0.25, 
                                    only.pos = T)

write.csv(Cluster_marker_02, file = "Myeloid_cells_marker_genes_res02.csv")

top10_genes <- Cluster_marker_02 %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 10) %>% ungroup()

DoHeatmap(scMyeloid, features = unique(top10_genes$gene), assay = "RNA", slot = "scale.data") +
  NoLegend()

DotPlot(scMyeloid, features = unique(top10_genes$gene), group.by = "integrated_snn_res.0.2") +
  coord_flip()

Cluster_marker_02_noRPS <- Cluster_marker_02[which(grepl("^RPS*", Cluster_marker_02$gene) == F),]
Cluster_marker_02_noRPL <- Cluster_marker_02_noRPS[which(grepl("^RPL*", Cluster_marker_02_noRPS$gene) == F),]
top10_genes_noRPS <- Cluster_marker_02_noRPL %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 10) %>% ungroup()

DoHeatmap(scMyeloid, features = unique(top10_genes_noRPS$gene), assay = "RNA", slot = "scale.data") +
  NoLegend() + labs(title = "Heatmap without RPS transcripts")

DotPlot(scMyeloid, features = unique(top10_genes_noRPS$gene), group.by = "integrated_snn_res.0.2") +
  coord_flip() + labs(title = "Dotplot without RPS transcripts")

#Top 50 Marker
top50_genes_02 <- Cluster_marker_02_noRPL %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 50) %>% ungroup()

write.csv(top50_genes_02, file = "Myeloid_cells_top50_marker_genes_res02.csv")
```

```{r Annotation myeloid cell populations}
Myo.sce <- as.SingleCellExperiment(scMyeloid, assay = "integrated")
Myo.sce
ls('package:celldex')
ref.set <- celldex::HumanPrimaryCellAtlasData()
head(unique(ref.set$label.fine),20)
pred.cnts <- SingleR::SingleR(test = Myo.sce, ref = ref.set, labels = ref.set$label.fine)

lbls.keep <- table(pred.cnts$labels)>10
scMyeloid$SingleR.labels <- ifelse(lbls.keep[pred.cnts$labels], pred.cnts$labels, 'Other')

DimPlot(scMyeloid, reduction='umap', group.by='SingleR.labels', label = T) + NoLegend()

# Potential Idents for Res 0.4
Myos_annotations <- scMyeloid@meta.data %>% group_by(integrated_snn_res.0.4, SingleR.labels) %>% 
  summarise(potential.ident = n())

Myos_top_ann_04 <- Myos_annotations %>% group_by(integrated_snn_res.0.4) %>% 
  slice_max(potential.ident, n = 3)

print(Myos_top_ann_04)

# Potential Idents for Res 0.2
Myos_annotations_02 <- scMyeloid@meta.data %>% group_by(integrated_snn_res.0.2, SingleR.labels) %>% 
  summarise(potential.ident = n())

Myos_top_ann_02 <- Myos_annotations_02 %>% group_by(integrated_snn_res.0.2) %>% 
  slice_max(potential.ident, n = 3)

print(Myos_top_ann_02)


saveRDS(scMyeloid, file = "Myeloid_Cells.rds")
```

## Preliminary Annotation 

### Resolution 0.4

Cluster 0 - Macrophages
Cluster 1 - Macrophages
Cluster 2 - Monocytes
Cluster 3 - iPSCs or B cells
Cluster 4 - endothelial cells?
Cluster 5 - Other and NK cells
Cluster 6 - Other 
Cluster 7 - Neutrophils


### Resolution 0.2 

Cluster 0 - Macrophages
Cluster 1 - Monocytes 
Cluster 2 - Macrophages 
Cluster 3 - B cells? 
Cluster 4 - Other
Cluster 5 - Neutrophils


Since SingleR does not always reliably identify all cell types, the marker gene expression will be examined more closely to determine the subpopulation identity: 

## Marker genes resolution 0.4

### Monocyte Marker Genes Res 0.4

```{r Violin Plots Monocyte markers}
Idents(scMyeloid) <- "integrated_snn_res.0.4"

#Monocyte marker
VlnPlot(scMyeloid, features = "CD14", layer = "data")
VlnPlot(scMyeloid, features = "FCGR3A", layer = "data") #CD16a (NK cells, macrophages)
VlnPlot(scMyeloid, features = "S100A8", layer = "data")
```

### Activated innate Immune cells 0.4

```{r Monocyte markers 2}
VlnPlot(scMyeloid, features = "NEAT1", layer = "data")
VlnPlot(scMyeloid, features = "CD83", layer = "data")
VlnPlot(scMyeloid, features = "TNFAIP3", layer = "data")
VlnPlot(scMyeloid, features = "NFKB1", layer = "data")
```

### Macrophages marker genes

```{r Macrophage Marker genes}
VlnPlot(scMyeloid, features = "CCR1", layer = "data")
VlnPlot(scMyeloid, features = "LYZ", layer = "data")
VlnPlot(scMyeloid, features = "APOE", layer = "data")
VlnPlot(scMyeloid, features = "CD4", layer = "data")
VlnPlot(scMyeloid, features = "ITGAM", layer = "data") #CD11b
VlnPlot(scMyeloid, features = "CSF1R", layer = "data") #CD115
VlnPlot(scMyeloid, features = "CD163", layer = "data") # (M2 Macs)
VlnPlot(scMyeloid, features = "MRC1", layer = "data") #CD206(M2 Mac)
VlnPlot(scMyeloid, features = "CD209", layer = "data") #(M2 Mac)
VlnPlot(scMyeloid, features = "TGFB1", layer = "data") #(M2 Mac)
VlnPlot(scMyeloid, features = "MAF", layer = "data") #(M2 Mac)
```


### DC Marker Gene

```{r Violins for DC marker genes}
VlnPlot(scMyeloid, features = "ID2") # cDC1 and cDC2 marker

VlnPlot(scMyeloid, features = "IRF8") # cDC1 marker
VlnPlot(scMyeloid, features = "CLEC9A") # cDC1 marker
VlnPlot(scMyeloid, features = "LY75") #CD205 # cDC1 marker
VlnPlot(scMyeloid, features = "ITGAE") #CD103 # cDC1 marker
VlnPlot(scMyeloid, features = "CD8A") #cDC1 marker
VlnPlot(scMyeloid, features = "CADM1") #cDC1 marker
VlnPlot(scMyeloid, features = "ITGAX") #CD11C

VlnPlot(scMyeloid, features = "IRF4") # cDC2 marker
VlnPlot(scMyeloid, features = "CD1C") #cDC2 marker
VlnPlot(scMyeloid, features = "FCER1A") # cDC2 Marker
VlnPlot(scMyeloid, features = "CLEC10A") # cDC2 Marker
VlnPlot(scMyeloid, features = "FCGR2B") #CD32b # cDC2 Marker
```


### pDC Marker Gene

```{r pDC marker}
VlnPlot(scMyeloid, features = "IL3RA") #CD123 
VlnPlot(scMyeloid, features = "NRP1") #CD304
```


### Neutrophil marker

```{r Neutrophil violins}
VlnPlot(scMyeloid, features = "FCGR3B") #CD16b
VlnPlot(scMyeloid, features = "CXCR2")
VlnPlot(scMyeloid, features = "IL17RA")
```

### B cell marker genes

```{r B cell marker}
VlnPlot(scMyeloid, features = "CD52")
VlnPlot(scMyeloid, features = "CD37")
VlnPlot(scMyeloid, features = "IGHA1")
```

# Inflammatory marker genes

```{r inflammation marker}
VlnPlot(scMyeloid, features = "IL1B")
VlnPlot(scMyeloid, features = "IL6")
VlnPlot(scMyeloid, features = "CCL24")
VlnPlot(scMyeloid, features = "MMP12")
```


```{r Annotate myeloid subpopulations}
scMyeloid <- RenameIdents(scMyeloid, '0' = 'M2 Macrophages',
                          '1' = 'M1 Macrophages', 
                          '2' = 'Inflammatory monocyte-derived Macrophages', 
                          '3' = 'cDC1', 
                          '4' = 'Epithelial?', 
                          '5' = 'cDC2', 
                          '6' = 'pDCs?',
                          '7' = 'Neutrophils')

DimPlot(scMyeloid, reduction = "umap", label = T)

scMyeloid$Annotation <- Idents(scMyeloid)

Cluster_over_time_ann_04 <- scMyeloid@meta.data %>% group_by(tp, Annotation) %>% 
  summarise(n = n()) %>% mutate(Freq = (n/sum(n))*100)

ggplot(Cluster_over_time_ann_04, aes(x = tp, y = Freq, fill = Annotation)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15) + labs(y = "% of myeloid cells") + 
  theme_minimal()

DimPlot(scMyeloid, reduction = "umap", cols = farbenblind15) + coord_equal(ratio = 1.5)

ggsave("Myeloid_cells_umap_res04.tiff", plot = last_plot(),
       device = "tiff", dpi = 300, width = 10, height = 7, units = "in", compression = "lzw")

saveRDS(scMyeloid, file = "Myeloid_Cells_annotated.rds")
```
