---
title: "T Cell Receptors"
author: "Julia Hitschfel"
date: "2025-09-08"
output: html_document
---

# TCR Analysis

In this script we take a first look at T cell receptor sequences. 
For this purpos we use the R package scRepertoire. 
Documentation for this package is available at the link below: 
https://www.borch.dev/uploads/screpertoire/articles/attaching_sc


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(scRepertoire)
library(Seurat)
library(dplyr)
library(tidyr)
```

```{r Load Contigs}
# Directory example
all.tps <- list.dirs("/data/", recursive = F)

contig.path <- "/vdj_t/filtered_contig_annotations.csv"
contig.list <- list()
counter <- 1

for(i in all.tps){
  vdj_t_contig <- read.csv(paste0(i, contig.path))
  contig.list[[counter]] <- vdj_t_contig
  counter <- counter + 1
}

combined.TCR <- combineTCR(contig.list, 
                           samples = c("2024_07_15",  "2024_11_28", "2025_02_17", 
                                            "2025_03_07","2025_03_24", "2025_08_06", "2025_10_22"),
                           removeNA = FALSE, 
                           removeMulti = FALSE, 
                           filterMulti = FALSE)

head(combined.TCR[[1]])

clonalQuant(combined.TCR, 
            cloneCall="strict", 
            chain = "both", 
            scale = TRUE)

clonalAbundance(combined.TCR, 
                cloneCall = "gene", 
                scale = TRUE)

clonalCompare(combined.TCR, 
                  top.clones = 3, 
                  cloneCall="aa", 
                  graph = "alluvial")

clonalScatter(combined.TCR, 
              cloneCall ="gene", 
              x.axis = "2025_03_24", 
              y.axis = "2025_08_06",
              dot.size = "total",
              graph = "proportion")

clonalScatter(combined.TCR, 
              cloneCall ="gene", 
              x.axis = "2025_08_06", 
              y.axis = "2025_10_22",
              dot.size = "total",
              graph = "proportion")
```

```{r Clonal Dynamics}
clonalHomeostasis(combined.TCR, 
                  cloneCall = "gene")

clonalProportion(combined.TCR, 
                 cloneCall = "gene") 
```

```{r Combine with scobject}
scobject <- readRDS("integrated_seurat_refined_annotations.rds")

#the cell barcodes are different here, so we need to change the barcodes in scobject to be able to combine the expression 

old_BCs <- colnames(scobject@assays$RNA$counts)
old_BCs <- ifelse(grepl("_1", old_BCs) == T, paste0("2024_07_15_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_2", old_BCs) == T, paste0("2024_11_28_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_3", old_BCs) == T, paste0("2025_02_17_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_4", old_BCs) == T, paste0("2025_03_07_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_5", old_BCs) == T, paste0("2025_03_24_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_6", old_BCs) == T, paste0("2025_08_06_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_7", old_BCs) == T, paste0("2025_10_22_",old_BCs), old_BCs) 
new_BCs <- substr(old_BCs, 0, nchar(old_BCs) - 2)

scobject <- RenameCells(scobject, new.names = new_BCs)

sce <- as.SingleCellExperiment(scobject, assay = "SCT")

sce <- combineExpression(combined.TCR, 
                         sce, 
                         cloneCall="gene", 
                         group.by = "sample", 
                         proportion = TRUE)

#Define color palette 
colorblind_vector <- hcl.colors(n=7, palette = "inferno", fixup = TRUE)

scobject_TCR <- combineExpression(combined.TCR, 
                                   scobject, 
                                   cloneCall="gene", 
                                   group.by = "sample", 
                                   proportion = FALSE, 
                                   cloneSize=c(Single=1, Small=5, Medium=20, Large=100,
                                               Hyperexpanded=500))

Seurat::DimPlot(scobject_TCR, group.by = "cloneSize") +
    scale_color_manual(values=rev(colorblind_vector[c(1,3,4,5,7)]))
```


```{r Overlay TCRs on T cell scobject}
Tcell_scobject <- readRDS(file = "Tcells_CCAintegration_annotation.rds")

old_BCs <- colnames(Tcell_scobject@assays$RNA$counts)
old_BCs <- ifelse(grepl("_1", old_BCs) == T, paste0("2024_07_15_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_2", old_BCs) == T, paste0("2024_11_28_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_3", old_BCs) == T, paste0("2025_02_17_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_4", old_BCs) == T, paste0("2025_03_07_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_5", old_BCs) == T, paste0("2025_03_24_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_6", old_BCs) == T, paste0("2025_08_06_",old_BCs), old_BCs) 
old_BCs <- ifelse(grepl("_7", old_BCs) == T, paste0("2025_10_22_",old_BCs), old_BCs) 
new_BCs <- substr(old_BCs, 0, nchar(old_BCs) - 2)

Tcell_scobject <- RenameCells(Tcell_scobject, new.names = new_BCs)

Tcell_scobject_TCR <- combineExpression(combined.TCR, 
                                   Tcell_scobject, 
                                   cloneCall="gene", 
                                   group.by = "sample", 
                                   proportion = FALSE, 
                                   cloneSize=c(Single=1, Small=5, Medium=20, Large=100,
                                               Hyperexpanded=500))

Seurat::DimPlot(Tcell_scobject_TCR, group.by = "cloneSize") +
    scale_color_manual(values=rev(colorblind_vector[c(1,3,4,5,7)]))

saveRDS(Tcell_scobject_TCR, file = "Tcells_CCAintegration_annotation_TCRs.rds")


DimPlot(Tcell_scobject_TCR, group.by = "SCT_snn_res.0.6", label = T)
```


```{r Expanded Clone}
clonalCompare(combined.TCR, 
                  top.clones = 5, 
              samples = c("2025_03_24", "2025_08_06"),
                  cloneCall="aa", 
                  graph = "alluvial")

clonalCompare(combined.TCR, 
              chain = "TRB",
                  top.clones = 10, 
              samples = c("2025_03_24", "2025_08_06"),
                  cloneCall="aa", 
                  graph = "alluvial")

clonalCompare(combined.TCR, 
              chain = "TRB",
                  top.clones = 10, 
              samples = c("2025_08_06", "2025_10_22"),
                  cloneCall="aa", 
                  graph = "alluvial")
```

### Examine persisting T cell clones

```{r Persisting T cell clones}
# Klon "CASSLEGAGEQYF"
Tcell_scobject_TCR@meta.data$date[grep("CASSLEGAGEQYF", Tcell_scobject_TCR@meta.data$CTaa)]
scobject_TCR@meta.data$date[grep("CASSLEGAGEQYF", scobject_TCR@meta.data$CTaa)]

# Klon "CASSRDLSSYEQYF"
unique(Tcell_scobject_TCR@meta.data$date[grep("CASSRDLSSYEQYF", Tcell_scobject_TCR@meta.data$CTaa)])

grep("CASSRDLSSYEQYF", combined.TCR$`2025_08_06`$cdr3_aa2)
grep("CASSRDLSSYEQYF", combined.TCR$`2025_03_24`$cdr3_aa2)
combined.TCR$`2025_03_24`$barcode[744]
grep("2025_03_24_GTTCTCGGTTACCGAT-1", rownames(scobject_TCR@meta.data))

# Klon "CASSRDLSSYENQFF"
unique(Tcell_scobject_TCR@meta.data$date[grep("CASSRDLSSYNEQFF", Tcell_scobject_TCR@meta.data$CTaa)])


# Klone nur am 06.08. und 22.10. detektiert:
unique(Tcell_scobject_TCR@meta.data$date[grep("CASTSGAGTGDGYTF", Tcell_scobject_TCR@meta.data$CTaa)])
unique(Tcell_scobject_TCR@meta.data$date[grep("CASSVQINTEAFF", Tcell_scobject_TCR@meta.data$CTaa)])
unique(Tcell_scobject_TCR@meta.data$date[grep("CASSLRTGGNTEAFF", Tcell_scobject_TCR@meta.data$CTaa)])
unique(Tcell_scobject_TCR@meta.data$date[grep("CASSGRTGSYSPLHF", Tcell_scobject_TCR@meta.data$CTaa)])
unique(Tcell_scobject_TCR@meta.data$date[grep("CASSFSTGQSYEQYF", Tcell_scobject_TCR@meta.data$CTaa)])
```


```{r Clonal Compare}
clonalCompare(combined.TCR, 
              chain = "TRB",
                  top.clones = 10, 
                  cloneCall="aa", 
                  graph = "alluvial")
```


```{r Diversity Index}
combined.TCR <- addVariable(combined.TCR, 
                            variable.name = "Timepoint", 
                            variables = c("Pre_CART", "Pre_CART", "Pre_CART", "Early_Post_CART",
                                          "Early_Post_CART", "Late_Post_CART", "Late_Post_CART"))

clonalDiversity(combined.TCR, 
                cloneCall = "aa", 
                chain = "TRB")

clonalDiversity(combined.TCR, 
                cloneCall = "aa",
                chain = "TRB",
                x.axis = "Timepoint",
                metric = "inv.simpson")

inv.simpson.table <- clonalDiversity(combined.TCR, 
                cloneCall = "aa",
                chain = "TRB",
                x.axis = "Timepoint",
                metric = "inv.simpson", exportTable = T)


write.csv(inv.simpson.table, file = "inv_simpson_TP.csv")

clonalDiversity(combined.TCR, 
                cloneCall = "aa",
                chain = "TRB",
                x.axis = "Timepoint", 
                metric = "shannon")

shannon.table <- clonalDiversity(combined.TCR, 
                cloneCall = "aa",
                chain = "TRB",
                x.axis = "Timepoint", 
                metric = "shannon", exportTable = T)

write.csv(shannon.table, file = "shannon_TP.csv")

clonalOverlap(combined.TCR, 
              chain = "TRB",
              cloneCall = "aa", 
              method = "raw")

clonalOverlap(combined.TCR, 
              chain = "TRB",
              cloneCall = "aa", 
              method = "morisita")

MHSI <- clonalOverlap(combined.TCR, 
              chain = "TRB",
              cloneCall = "aa", 
              method = "morisita", exportTable = T)

write.csv(MHSI, file = "morisita_horn.csv")
```

Schauen wir auch nochmal kurz als Bestätigung der Daten, für wie viele Barcodes die wir in der VDJ Library gefunden haben, wir auch assoziierte Barcodes in der Genexpression haben: 

```{r}
VDJ.barcodes <- vector()
for(i in names(combined.TCR)){
  VDJ.barcodes.tp <- combined.TCR[[i]]$barcode
  VDJ.barcodes <- c(VDJ.barcodes, VDJ.barcodes.tp)
}

GEX.Tcell.barcodes <- rownames(Tcell_scobject@meta.data)

Only.VDJ.Tcell.barcodes <- setdiff(VDJ.barcodes, GEX.Tcell.barcodes)

print(paste0("There are ", length(Only.VDJ.Tcell.barcodes), " barcodes that were only found in the VDJ library and not in the GEX library in the T cell subclustering."))

GEX.All.barcodes <- rownames(scobject@meta.data)

Only.VDJ.barcodes <- setdiff(VDJ.barcodes, GEX.All.barcodes)

print(paste0("There are ", length(Only.VDJ.barcodes), " barcodes that were only found in the VDJ library and not in the GEX library in the whole dataset."))
```


```{r VDJ Transcripts without GEX in T cell subclustering}
Maybe.T.cells <- setdiff(Only.VDJ.Tcell.barcodes, Only.VDJ.barcodes)

#check if they are from a specific timepoint
timepoints <- unique(substr(Maybe.T.cells, 0, 10))
timepoints

# This is not the case, they belong to different timepoints

# Derived from specific cluster? 
test.VDJ <- scobject@meta.data 
test.VDJ$barcode <- rownames(test.VDJ)
test.VDJ.sub <- subset(test.VDJ, barcode %in% Maybe.T.cells)

Cluster.VDJ.missing <- test.VDJ.sub %>% group_by(Annotations.0.4) %>% summarise(n_Cluster = n())
```


```{r Early Timepoints}

clonalCompare(combined.TCR, 
              chain = "TRB",
                  top.clones = 20, 
              samples = c("2024_07_15", "2024_11_28"),
                  cloneCall="aa", 
                  graph = "alluvial")

clonalCompare(combined.TCR, 
              chain = "TRB",
                  top.clones = 20, 
              samples = c("2024_07_15", "2024_11_28", "2025_02_17"),
                  cloneCall="aa", 
                  graph = "alluvial")

```

### Track specific T cell clones 

```{r Clonal overlap from 15.07. and 06.08.}

clonalCompare(combined.TCR, 
              chain = "TRB",
                  top.clones = 10, 
              samples = c("2024_07_15", "2025_08_06"),
                  cloneCall="aa", 
                  graph = "alluvial")


# Klon "CASSQDLSRVHEQFF"
unique(Tcell_scobject_TCR@meta.data$date[grep("CASSQDLSRVHEQFF", Tcell_scobject_TCR@meta.data$CTaa)])

# Klon "CASSRDLSSYEQYF"
unique(Tcell_scobject_TCR@meta.data$date[grep("CASSRDLSSYEQYF", Tcell_scobject_TCR@meta.data$CTaa)])

# Klon "CASSRDLSSYENQFF"
unique(Tcell_scobject_TCR@meta.data$date[grep("CASSRDLSSYNEQFF", Tcell_scobject_TCR@meta.data$CTaa)])

```


```{r Clusters clones}
grep("CASSRDLSSYEQYF", combined.TCR$`2025_10_22`$cdr3_aa2)
grep("CASSRDLSSYEQYF", combined.TCR$`2025_08_06`$cdr3_aa2)
grep("CASSRDLSSYEQYF", combined.TCR$`2024_07_15`$cdr3_aa2)

Clone1_2210_BCs <- combined.TCR$`2025_10_22`$barcode[grep("CASSRDLSSYEQYF",
                                                             combined.TCR$`2025_10_22`$cdr3_aa2)]
Clone1_0608_BCs <- combined.TCR$`2025_08_06`$barcode[grep("CASSRDLSSYEQYF",
                                                             combined.TCR$`2025_08_06`$cdr3_aa2)]
Clone1_1507_BCs <- combined.TCR$`2024_07_15`$barcode[grep("CASSRDLSSYEQYF",
                                                             combined.TCR$`2024_07_15`$cdr3_aa2)]

grep("CASSRDLSSYNEQFF", combined.TCR$`2025_10_22`$cdr3_aa2)
grep("CASSRDLSSYNEQFF", combined.TCR$`2025_08_06`$cdr3_aa2)
grep("CASSRDLSSYNEQFF", combined.TCR$`2024_07_15`$cdr3_aa2)

Clone2_2210_BCs <- combined.TCR$`2025_10_22`$barcode[grep("CASSRDLSSYNEQFF",
                                                             combined.TCR$`2025_10_22`$cdr3_aa2)]
Clone2_0608_BCs <- combined.TCR$`2025_08_06`$barcode[grep("CASSRDLSSYNEQFF",
                                                             combined.TCR$`2025_08_06`$cdr3_aa2)]
Clone2_1507_BCs <- combined.TCR$`2024_07_15`$barcode[grep("CASSRDLSSYNEQFF",
                                                             combined.TCR$`2024_07_15`$cdr3_aa2)]
```


```{r Overlap early and late Cluster}
Tcell_scobject_TCR@meta.data$Expanded.clones <- ifelse(rownames(Tcell_scobject_TCR@meta.data) %in%
                                                        Clone1_2210_BCs, "Clone_1_CASSRDLSSYEQYF", 
                                                      NA)

Tcell_scobject_TCR@meta.data$Expanded.clones <- ifelse(rownames(Tcell_scobject_TCR@meta.data) %in%
                                                        Clone1_0608_BCs, "Clone_1_CASSRDLSSYEQYF", 
                                                      Tcell_scobject_TCR@meta.data$Expanded.clones)

Tcell_scobject_TCR@meta.data$Expanded.clones <- ifelse(rownames(Tcell_scobject_TCR@meta.data) %in%
                                                        Clone1_1507_BCs, "Clone_1_CASSRDLSSYEQYF", 
                                                      Tcell_scobject_TCR@meta.data$Expanded.clones)

Tcell_scobject_TCR@meta.data$Expanded.clones <- ifelse(rownames(Tcell_scobject_TCR@meta.data) %in%
                                                        Clone2_2210_BCs, "Clone_2_CASSRDLSSYNEQFF", 
                                                      Tcell_scobject_TCR@meta.data$Expanded.clones)

Tcell_scobject_TCR@meta.data$Expanded.clones <- ifelse(rownames(Tcell_scobject_TCR@meta.data) %in%
                                                        Clone2_0608_BCs, "Clone_2_CASSRDLSSYNEQFF", 
                                                      Tcell_scobject_TCR@meta.data$Expanded.clones)

Tcell_scobject_TCR@meta.data$Expanded.clones <- ifelse(rownames(Tcell_scobject_TCR@meta.data) %in%
                                                        Clone2_1507_BCs, "Clone_2_CASSRDLSSYNEQFF", 
                                                      Tcell_scobject_TCR@meta.data$Expanded.clones)

Tcell_scobject_TCR@meta.data$Expanded.clones <- factor(Tcell_scobject_TCR@meta.data$Expanded.clones,
                                                         levels = c("NA", "Clone_1_CASSRDLSSYEQYF", 
                                                                    "Clone_2_CASSRDLSSYNEQFF"))

DimPlot(Tcell_scobject_TCR, reduction = "umap", group.by = "Expanded.clones", pt.size = 1, 
        cols = c("red", "blue", "lightgrey"))

Clones_summary <- table(Tcell_scobject_TCR@meta.data$Expanded.clones,
                        Tcell_scobject_TCR@meta.data$date)
Clones_summary

Clones_summary_cluster <- table(Tcell_scobject_TCR@meta.data$Expanded.clones,
                        Tcell_scobject_TCR@meta.data$date, 
                        Tcell_scobject_TCR@meta.data$SCT_snn_res.0.6)
```

### Clonal overlap pre CAR T Therapy

```{r }
#ggraph needs to be loaded due to issues with ggplot
library(ggraph)

Tcell_scobject_TCR$Annotations <- Idents(Tcell_scobject_TCR)

#No Identity filter
clonalNetwork(Tcell_scobject_TCR, 
              reduction = "umap", 
              group.by = "Annotations",
              filter.clones = NULL,
              filter.identity = NULL,
              cloneCall = "CTaa")

clonalOccupy(Tcell_scobject_TCR, 
              x.axis = "SCT_snn_res.0.6")

clonalOccupy(Tcell_scobject_TCR, 
             x.axis = "SCT_snn_res.0.6", 
             proportion = TRUE, 
             label = FALSE)

levels(Idents(Tcell_scobject_TCR))
```


```{r Clonal Overlap with MHSI}
clonalOverlap(combined.TCR,
              chain = "TRB",
              cloneCall = "aa", 
              method = "morisita")

clonalOverlap(combined.TCR, 
              chain = "TRB",
              cloneCall = "aa", 
              method = "raw")
```


```{r Overlap with minimum Doublets only}
for(i in 1:length(combined.TCR)){
Min_Doublet <- which(duplicated(combined.TCR[[i]]$cdr3_aa2))
TRB_AA_Seq_Dbl <- unique(combined.TCR[[i]]$cdr3_aa2[Min_Doublet])
combined.TCR[[i]] <- subset(combined.TCR[[i]], combined.TCR[[i]]$cdr3_aa2 %in% TRB_AA_Seq_Dbl)
}

clonalOverlap(combined.TCR,
              chain = "TRB",
              cloneCall = "aa", 
              method = "morisita")

clonalOverlap(combined.TCR, 
              chain = "TRB",
              cloneCall = "aa", 
              method = "raw")
```


```{r Expanded Clones before CAR T}

TRBVs_before_CAR <- c(combined.TCR$`2024_07_15`$cdr3_aa2, combined.TCR$`2024_11_28`$cdr3_aa2,
                      combined.TCR$`2025_02_17`$cdr3_aa2)

TRBVs_before_CAR_NAfilt <- TRBVs_before_CAR[!is.na(TRBVs_before_CAR)]
TRBVs_before_CAR_NAfilt <- unique(TRBVs_before_CAR_NAfilt)

# Before CAR T cell Therapy, there are 355 unique TCR beta chains detected by amino acid sequence 

TRBVs_after_CAR <- c(combined.TCR$`2025_03_07`$cdr3_aa2, combined.TCR$`2025_03_24`$cdr3_aa2,
                      combined.TCR$`2025_08_06`$cdr3_aa2, combined.TCR$`2025_10_22`$cdr3_aa2)

TRBVs_after_CAR_NAfilt <- TRBVs_after_CAR[!is.na(TRBVs_after_CAR)]
TRBVs_after_CAR_NAfilt <- unique(TRBVs_after_CAR_NAfilt)

# After CAR T cell therapy there are 613 unique TCR beta chains detected by amino acid sequence

TRBVs_overlap <- intersect(TRBVs_before_CAR_NAfilt, TRBVs_after_CAR_NAfilt)
# There are 31 TCR beta chains that overlap before and after CAR T cell therapy 

TRBVs_depleted <- setdiff(TRBVs_before_CAR_NAfilt, TRBVs_after_CAR_NAfilt)
# There are 324 TCR beta chains that are present before CAR T cell therapy, which are not present after CAR T cell therapy 

TRBVs_new <- setdiff(TRBVs_after_CAR_NAfilt, TRBVs_before_CAR_NAfilt)
# There are 582 TCR beta chains that are found after CAR T cell therapy, which are not detected prior to therapy 

# We can now examine if these clonotypes fall within specific clusters, but my guess is that this is all stochastic 
Tcell_scobject_TCR@meta.data <- Tcell_scobject_TCR@meta.data %>% 
  separate(CTaa, into = c("aa_alpha", "aa_beta"), sep = "_")

Cell_IDs_overlap <- rownames(Tcell_scobject_TCR@meta.data)[which(Tcell_scobject_TCR$aa_beta %in%
                                                                   TRBVs_overlap)]
Tcell_scobject_TCR$CAR.comparison <- rep(NA)
Tcell_scobject_TCR$CAR.comparison <- ifelse(rownames(Tcell_scobject_TCR@meta.data) %in%
                                              Cell_IDs_overlap, "Persisting",
                                            Tcell_scobject_TCR$CAR.comparison)
# Clones depleted
Cell_IDs_depleted <- rownames(Tcell_scobject_TCR@meta.data)[which(Tcell_scobject_TCR$aa_beta %in%
                                                                   TRBVs_depleted)]

Tcell_scobject_TCR$CAR.comparison <- ifelse(rownames(Tcell_scobject_TCR@meta.data) %in%
                                              Cell_IDs_depleted, "Depleted",
                                            Tcell_scobject_TCR$CAR.comparison)

# Clones new
Cell_IDs_new <- rownames(Tcell_scobject_TCR@meta.data)[which(Tcell_scobject_TCR$aa_beta %in%
                                                                   TRBVs_new)]

Tcell_scobject_TCR$CAR.comparison <- ifelse(rownames(Tcell_scobject_TCR@meta.data) %in%
                                              Cell_IDs_new, "New Clone",
                                            Tcell_scobject_TCR$CAR.comparison)

DimPlot(Tcell_scobject_TCR, reduction = "umap", group.by = "CAR.comparison", 
        cols = c("blue", "lightblue", "yellow", "lightgrey"))

DimPlot(Tcell_scobject_TCR, reduction = "umap", label = T)

Persisting_Info <- Tcell_scobject_TCR@meta.data %>% filter(CAR.comparison == "Persisting") %>% 
  select("CTgene", "CTnt", "aa_alpha", "aa_beta")

write.csv(Persisting_Info, file = "Persisting_Tcell_clones.csv")

Depleted_Info <- Tcell_scobject_TCR@meta.data %>% filter(CAR.comparison == "Depleted") %>% 
  select("CTgene", "CTnt", "aa_alpha", "aa_beta")

write.csv(Depleted_Info, file = "Depleted_Tcell_clones.csv")

Depleted_expanded_beta_GEX <- Depleted_Info %>% group_by(aa_beta) %>% summarise(n = n())
write.csv(Depleted_expanded_beta_GEX, file = "Depleted_expanded_TRBV_count.csv")

New_Info <- Tcell_scobject_TCR@meta.data %>% filter(CAR.comparison == "New Clones") %>% 
  select("CTgene", "CTnt", "aa_alpha", "aa_beta")

write.csv(New_Info, file = "New_Tcell_clones.csv")
```

Most of the T cell receptor clones that persist or newly expand post CAR T cell therapy fall in cluster 0 (CD8+ TRM)

```{r Bias in TRBV gene usage}
clonalOverlap(combined.TCR, 
              chain = "TRB",
              cloneCall = "gene", 
              method = "raw")

clonalOverlap(combined.TCR, 
              chain = "TRB",
              cloneCall = "gene", 
              method = "morisita")

```
