---
title: "B Cell Subclustering"
author: "Julia Hitschfel"
date: "2025-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Seurat)
library(SeuratObject)
library(scDblFinder)
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(colorspace)
library(RColorBrewer)
```

## B Cell Subclustering 


```{r Import Dataset and Examine B cell Frequenices}
scobject <- readRDS(file = "integrated_seurat_refined_annotations.rds")

Idents(scobject) <- scobject$SCT_snn_res.0.4
```

```{r B Zellen pro Zeitpunkt}
B.cells <- scobject@meta.data %>% filter(SCT_snn_res.0.4 == 1 | SCT_snn_res.0.4 == 9 | 
                                           SCT_snn_res.0.4 == 11 | SCT_snn_res.0.4 == 12)

B.cells.per.TP <- B.cells %>% group_by(date, SCT_snn_res.0.4) %>% summarise(B.count = n())

ggplot(B.cells.per.TP, aes(x = date, y = B.count, fill = SCT_snn_res.0.4)) + 
  geom_col(position = "dodge") + labs(title = "Absolute Cell Counts in Clusters with B cells") + 
  ylab(label = "Cell Count") + 
  scale_fill_manual(name = "Clusters containing B cells", 
                    values = c("lightblue", "black", "grey", "darkblue")) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
# Approximate B cell frequencies 
All.cells.per.TP <- scobject@meta.data %>% group_by(date) %>% summarise(Cell.count = n())
B.cell.freq.per.TP <- left_join(B.cells.per.TP, All.cells.per.TP, by = "date")
B.cell.freq.per.TP <- B.cell.freq.per.TP %>% mutate(Freq = (B.count/Cell.count)*100)

ggplot(B.cell.freq.per.TP, aes(x = date, y = Freq, fill = SCT_snn_res.0.4)) + 
  geom_col(position = "dodge") + labs(title = "Cell Frequencies Clusters with B cells") + 
  ylab(label = "Cell Frequencies") + 
  scale_fill_manual(name = "Clusters containing B cells", 
                    values = c("lightblue", "black", "grey", "darkblue")) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r Import B Cells}
All_Bcells_comb <- WhichCells(scobject, idents = c(1,9,11,12))

All.data <- list.dirs("/data/", recursive = F)

list.10x <- list()
counter <- 1
for(i in All.data){
  rnaseq.data <- Read10X_h5(paste0(i, "/count/sample_filtered_feature_bc_matrix.h5"))
  pbmc <- CreateSeuratObject(counts = rnaseq.data, project = "Colitis", min.cells = 3, 
                             min.features = 200)
  pbmc <- RenameCells(pbmc, new.names = paste0(colnames(pbmc), "_", counter))
  pbmc <- subset(pbmc, cells = All_Bcells_comb)
  list.10x[[i]] <- pbmc
  counter <- counter + 1
}


print("Cell Numbers before Filtering:")
for(i in list.10x){print(nrow(i@meta.data))}
```

```{r B cell QC}
for(i in 1:length(list.10x)){
  pbmc <- list.10x[[i]]
  pbmc@meta.data$date <- substr(names(list.10x)[i], 25, 100)
  pbmc@meta.data$percent.mt <- PercentageFeatureSet(pbmc, pattern = "^MT")
  pbmc@meta.data$percent.rps <- PercentageFeatureSet(pbmc, pattern = "^RPS")
  pbmc@meta.data$percent.rpl <- PercentageFeatureSet(pbmc, pattern = "^RPL")
  list.10x[[i]] <- pbmc
}

scobject <- merge(x = list.10x[[1]], y = c(list.10x[2:length(list.10x)]))

sum_table <- scobject@meta.data %>% group_by(date) %>% 
  summarise(RNA.count.median = median(nCount_RNA), RNA.feature.median = median(nFeature_RNA), 
            MT.freq.median = median(percent.mt), RPS.freq.median = median(percent.rps), 
            RPL.freq.median = median(percent.rpl))

sum_table

ggplot(scobject@meta.data, aes(x = date, y = nCount_RNA)) + geom_boxplot()

ggplot(scobject@meta.data, aes(x = date, y = nFeature_RNA)) + geom_boxplot()

```

```{r B Cell Clustering}
scobject <- NormalizeData(scobject, normalization.method = "LogNormalize", scale.factor = 10000)

scobject <- FindVariableFeatures(scobject, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(scobject), 10)
top10

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(scobject)  + NoLegend()
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE) + NoLegend()
plot1 + plot2

all.genes <- rownames(scobject)
scobject <- ScaleData(scobject, features = all.genes)

# Cluster after Regression 
scobject <- RunPCA(scobject, features = VariableFeatures(object = scobject))

# Examine and visualize PCA results a few different ways
print(scobject[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(scobject, dims = 1:2, reduction = "pca")
DimPlot(scobject, reduction = "pca") + NoLegend()

#Check some initial confounding factors on the PCA
FeaturePlot(scobject, reduction = "pca", features = "percent.mt")
FeaturePlot(scobject, reduction = "pca", features = "percent.rps")
FeaturePlot(scobject, reduction = "pca", features = "percent.rpl")

FeaturePlot(scobject, features = "nCount_RNA", reduction = "pca", max.cutoff = 5000) #+ NoLegend()
FeaturePlot(scobject, features = "nCount_RNA", reduction = "pca", dims = c(1,3), max.cutoff = 5000) +
  NoLegend()

FeaturePlot(scobject, features = "nFeature_RNA", reduction = "pca") + NoLegend()
FeaturePlot(scobject, features = "nFeature_RNA", reduction = "pca", dims = c(1,3)) + NoLegend()

DimHeatmap(scobject, dims = 1, cells = 500, balanced = TRUE)
DimHeatmap(scobject, dims = 1:15, cells = 500, balanced = TRUE)

#Cluster the cells
scobject <- FindNeighbors(scobject, dims = 1:25)
scobject <- FindClusters(scobject, resolution = c(0.4, 0.6, 0.8))
scobject <- RunUMAP(scobject, dims = 1:25)
saveRDS(scobject, file = "Bcells_regression.rds")


DimPlot(scobject, reduction = "umap", group.by = "seurat_clusters", label = T)
DimPlot(scobject, reduction = "umap", group.by = "date")
DimPlot(scobject, reduction = "umap", group.by = "date", split.by = "date", ncol = 3)
```

There seem to be visible batch effects, as seen for the total cell clustering and the T cell clustering. We will also attempt an integration here for this data, which could be challenging because of the low cell count at the later timepoints. 
There also are epithelial cells and myeloid cells contained in the clustering. Filters will be included to remove these cells from downstream analysis. Variable Features of epithelial cells are KRT17, MUC2 and die the strongest driver for PC1 are PIGR and KRT8. 


```{r B Cell Clustering Regression}
for(i in 1:length(list.10x)){
  pbmc <- list.10x[[i]]
  EpithelialCells <- WhichCells(pbmc, expression = KRT18 > 0 | KRT8 > 0 | PIGR > 1, slot = "counts")
  Monocytes <- WhichCells(pbmc, expression = CD14 > 1, slot = "counts")
  T_Cells <- WhichCells(pbmc, expression = CD3D > 0 | CD3E > 0, slot = "counts")
  Erys <- WhichCells(pbmc, expression = HBA1 > 0, slot = "counts")
  if("LILRB1" %in% rownames(pbmc@assays$RNA$counts)){
    pDCs <- WhichCells(pbmc, expression = LILRB1 > 1, slot = "counts")
  }
  Non_EpithelialCells <- setdiff(rownames(pbmc@meta.data), EpithelialCells)
  Non_Monocytes <- setdiff(Non_EpithelialCells, Monocytes)
  Non_Tcells <- setdiff(Non_Monocytes, T_Cells)
  Non_Erys <- setdiff(Non_Tcells, Erys)
  if(exists("pDCs")){Non_Erys <- setdiff(Non_Erys, pDCs)}
  pbmc <- subset(pbmc, cells = Non_Erys)
  list.10x[[i]] <- pbmc
}

scobject <- merge(x = list.10x[[1]], y = c(list.10x[2:length(list.10x)]))

print("B Cell Numbers after Filtering:")
for(i in list.10x){print(nrow(i@meta.data))}

sum_table <- scobject@meta.data %>% group_by(date) %>%
  summarise(RNA.count.median = median(nCount_RNA), RNA.feature.median = median(nFeature_RNA),
            MT.freq.median = median(percent.mt), RPS.freq.median = median(percent.rps),
            RPL.freq.median = median(percent.rpl))

sum_table

scobject <- NormalizeData(scobject, normalization.method = "LogNormalize", scale.factor = 10000)

scobject <- FindVariableFeatures(scobject, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(scobject), 10)
top10

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(scobject)  + NoLegend()
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE) + NoLegend()
plot1 + plot2

all.genes <- rownames(scobject)
scobject <- ScaleData(scobject, features = all.genes)

# Cluster after Regression 
scobject <- RunPCA(scobject, features = VariableFeatures(object = scobject))

# Examine and visualize PCA results a few different ways
print(scobject[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(scobject, dims = 1:2, reduction = "pca")
DimPlot(scobject, reduction = "pca") + NoLegend()

#Check some initial confounding factors on the PCA
FeaturePlot(scobject, reduction = "pca", features = "percent.mt")
FeaturePlot(scobject, reduction = "pca", features = "percent.rps")
FeaturePlot(scobject, reduction = "pca", features = "percent.rpl")

FeaturePlot(scobject, features = "nCount_RNA", reduction = "pca", max.cutoff = 5000) 
FeaturePlot(scobject, features = "nCount_RNA", reduction = "pca", dims = c(1,3), max.cutoff = 5000) +
  NoLegend()

FeaturePlot(scobject, features = "nFeature_RNA", reduction = "pca") + NoLegend()
FeaturePlot(scobject, features = "nFeature_RNA", reduction = "pca", dims = c(1,3)) + NoLegend()

DimHeatmap(scobject, dims = 1, cells = 500, balanced = TRUE)
DimHeatmap(scobject, dims = 1:15, cells = 500, balanced = TRUE)
```
 

```{r Clustering B cells}
#Cluster the cells
ElbowPlot(scobject, ndims = 25)
scobject <- FindNeighbors(scobject, dims = 1:20)
scobject <- FindClusters(scobject, resolution = c(0.4, 0.6, 0.8))
scobject <- RunUMAP(scobject, dims = 1:20)
saveRDS(scobject, file = "Bcells_subclustering.rds")

DimPlot(scobject, reduction = "umap", group.by = "RNA_snn_res.0.8", label = T)
DimPlot(scobject, reduction = "umap", group.by = "RNA_snn_res.0.6", label = T)
DimPlot(scobject, reduction = "umap", group.by = "RNA_snn_res.0.4", label = T)

DimPlot(scobject, reduction = "umap", group.by = "date")
DimPlot(scobject, reduction = "umap", group.by = "date", split.by = "date", ncol = 3)
```

```{r Identify markers}
scobject <- JoinLayers(scobject)
Idents(scobject) <- "RNA_snn_res.0.4"
Markers_clusters <- FindAllMarkers(scobject, logfc.threshold = 0.25, min.pct = 0.25, only.pos = T)
```


```{r Markers Violins}
VlnPlot(scobject, "MS4A1")
 
VlnPlot(scobject, "CD19")
 
VlnPlot(scobject, "MKI67")

VlnPlot(scobject, "FCRL5")

VlnPlot(scobject, "CD74")

VlnPlot(scobject, "CD79A")

VlnPlot(scobject, "CR2") #CD21
 
VlnPlot(scobject, "CD24")

VlnPlot(scobject, "LILRA4")
 
VlnPlot(scobject, "ITGAX")


VlnPlot(scobject, "IGHG1")
VlnPlot(scobject, "IGHA1")
VlnPlot(scobject, "JCHAIN")

VlnPlot(scobject, "CD22")
VlnPlot(scobject, "CD27")
VlnPlot(scobject, "IGHM")
VlnPlot(scobject, "IGHD")

VlnPlot(scobject, "MZB1")
VlnPlot(scobject, "PRDM1")
VlnPlot(scobject, "SDC1")
VlnPlot(scobject, "TNFRSF13C")

VlnPlot(scobject, "IDO1")
VlnPlot(scobject, "LYZ")
VlnPlot(scobject, "TYROBP")
VlnPlot(scobject, "LGALS4")
```

## Summary
We likely need to improve filters and include IDO1 and LGALS4, so just B cells are retained. 

For Myeloid Cells: IDO1 > 1, for epithelial cells: LGALS > 1, for T cells FYN > 1 as filters. 

## Additional Filters

```{r Filters and SCTransform}
list.10x.filt <- list()
for(i in 1:length(list.10x)){
  pbmc <- list.10x[[i]]
  MyeloidCells <- WhichCells(pbmc, expression = IDO1 > 0 | LYZ > 0, slot = "counts")
  T_Cells <- WhichCells(pbmc, expression = FYN > 1, slot = "counts")
  SECs <- WhichCells(pbmc, expression = LGALS4 > 1, slot = "counts")
  Non_MyeloidCells <- setdiff(rownames(pbmc@meta.data), MyeloidCells)
  Non_Tcells <- setdiff(Non_MyeloidCells, T_Cells)
  Non_SECs <- setdiff(Non_Tcells, SECs)
  pbmc <- subset(pbmc, cells = Non_SECs)
  list.10x.filt[[i]] <- pbmc
}

scBcells <- merge(x = list.10x.filt[[1]], y = c(list.10x.filt[2:length(list.10x.filt)]))

scBcells <- NormalizeData(scBcells, normalization.method = "LogNormalize", scale.factor = 10000)

scBcells <- FindVariableFeatures(scBcells, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(scBcells), 10)
top10

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(scBcells)  + NoLegend()
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE) + NoLegend()
plot1 + plot2

all.genes <- rownames(scBcells)
scBcells <- ScaleData(scBcells, features = all.genes)

scBcells <- RunPCA(scBcells, features = VariableFeatures(object = scBcells))

# Examine and visualize PCA results a few different ways
print(scBcells[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(scBcells, dims = 1:2, reduction = "pca")
DimPlot(scBcells, reduction = "pca") + NoLegend()

#Check some initial confounding factors on the PCA
FeaturePlot(scBcells, reduction = "pca", features = "percent.mt")
FeaturePlot(scBcells, reduction = "pca", features = "percent.rps")
FeaturePlot(scBcells, reduction = "pca", features = "percent.rpl")

FeaturePlot(scBcells, features = "nCount_RNA", reduction = "pca", max.cutoff = 5000) #+ NoLegend()
FeaturePlot(scBcells, features = "nCount_RNA", reduction = "pca", dims = c(1,3), max.cutoff = 5000) +
  NoLegend()

FeaturePlot(scBcells, features = "nFeature_RNA", reduction = "pca") + NoLegend()
FeaturePlot(scBcells, features = "nFeature_RNA", reduction = "pca", dims = c(1,3)) + NoLegend()

DimHeatmap(scBcells, dims = 1, cells = 500, balanced = TRUE)
DimHeatmap(scBcells, dims = 1:15, cells = 500, balanced = TRUE)

ElbowPlot(scBcells, ndims = 25)
scBcells <- FindNeighbors(scBcells, dims = 1:15)
scBcells <- FindClusters(scBcells, resolution = c(0.4, 0.6, 0.8))
scBcells <- RunUMAP(scBcells, dims = 1:15)
saveRDS(scBcells, file = "D:/Sequencing_data/analysis_all_tp/Bcells_subclustering.rds")

DimPlot(scBcells, reduction = "umap", group.by = "RNA_snn_res.0.8", label = T)
DimPlot(scBcells, reduction = "umap", group.by = "RNA_snn_res.0.6", label = T)
DimPlot(scBcells, reduction = "umap", group.by = "RNA_snn_res.0.4", label = T)

DimPlot(scBcells, reduction = "umap", group.by = "date")
DimPlot(scBcells, reduction = "umap", group.by = "date", split.by = "date", ncol = 3)

scBcells <- JoinLayers(scBcells)
Idents(scBcells) <- "RNA_snn_res.0.4"
Markers_clusters_filt <- FindAllMarkers(scBcells, logfc.threshold = 0.25, min.pct = 0.25, 
                                        only.pos = T)
write.csv(Markers_clusters_filt, file = "Bcell_cluster_marker_genes_filt.csv")

VlnPlot(scBcells, "MS4A1")
 
VlnPlot(scBcells, "CD19")
 
VlnPlot(scBcells, "MKI67")

VlnPlot(scBcells, "FCRL5")

VlnPlot(scBcells, "CD74")

VlnPlot(scBcells, "CD79A")

VlnPlot(scBcells, "CR2") #CD21
 
VlnPlot(scBcells, "CD24")

VlnPlot(scBcells, "LILRA4")
 
VlnPlot(scBcells, "ITGAX")


VlnPlot(scBcells, "IGHG1")
VlnPlot(scBcells, "IGHA1")
VlnPlot(scBcells, "JCHAIN")

VlnPlot(scBcells, "CD22")
VlnPlot(scBcells, "CD27")
VlnPlot(scBcells, "IGHM")
VlnPlot(scBcells, "IGHD")

VlnPlot(scBcells, "MZB1")
VlnPlot(scBcells, "PRDM1")
VlnPlot(scBcells, "SDC1")
VlnPlot(scBcells, "TNFRSF13C")

VlnPlot(scBcells, "LYZ")
VlnPlot(scBcells, "TYROBP")
```


```{r Cluster over time}
Cluster_over_time_filt <- scBcells@meta.data %>% group_by(date, RNA_snn_res.0.4) %>% 
  summarise(n = n()) %>% mutate(Freq = n/sum(n))

farbenblind15 <- c("#000000","#004949","#009292","#FF6DB6","#FFB6DB",
                  "#490092","#006DDB","#B66DFF","#6DB6FF","#B6DBFF",
                  "#920000","#924900","#DB6D00","#24FF24","#FFFF6D") 

ggplot(Cluster_over_time_filt, aes(x = date, y = Freq, fill = RNA_snn_res.0.4)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15)
```


```{r Examine Cluster over time}
Cluster_0 <- Cluster_over_time_filt %>% filter(RNA_snn_res.0.4 == 0)
Cluster_0_gg <- ggplot(Cluster_0, aes(x = date, y = Freq)) + geom_col(fill = "#000000") +  
  labs(title = "Cluster 0 (PB/PC) Frequencies") + theme_minimal()
Cluster_0_gg

Cluster_2 <- Cluster_over_time_filt %>% filter(RNA_snn_res.0.4 == 2)
Cluster_2_gg <- ggplot(Cluster_2, aes(x = date, y = Freq)) + geom_col(fill = "#009292") +  
  labs(title = "Cluster 2 (PB/PC) Frequencies") + theme_minimal()
Cluster_2_gg

Cluster_4 <- Cluster_over_time_filt %>% filter(RNA_snn_res.0.4 == 4)
Cluster_4_gg <- ggplot(Cluster_4, aes(x = date, y = Freq)) + geom_col(fill = "#FFB6DB") +  
  labs(title = "Cluster 4 (unsure?) Frequencies") + theme_minimal()
Cluster_4_gg

Cluster_5 <- Cluster_over_time_filt %>% filter(RNA_snn_res.0.4 == 5)
Cluster_5_gg <- ggplot(Cluster_5, aes(x = date, y = Freq)) + geom_col(fill = "#490092") +  
  labs(title = "Cluster 5 (unsure?) Frequencies") + theme_minimal()
Cluster_5_gg

Cluster_6 <- Cluster_over_time_filt %>% filter(RNA_snn_res.0.4 == 6)
Cluster_6_gg <- ggplot(Cluster_6, aes(x = date, y = Freq)) + geom_col(fill = "#006DDB") +  
  labs(title = "Cluster 6 (Memory o PB) Frequencies") + theme_minimal()
Cluster_6_gg

Cluster_7 <- Cluster_over_time_filt %>% filter(RNA_snn_res.0.4 == 7)
Cluster_7_gg <- ggplot(Cluster_7, aes(x = date, y = Freq)) + geom_col(fill = "#B66DFF") +  
  labs(title = "Cluster 7 (Other?) Frequencies") + theme_minimal()
Cluster_7_gg

Cluster_8 <- Cluster_over_time_filt %>% filter(RNA_snn_res.0.4 == 8)
Cluster_8_gg <- ggplot(Cluster_8, aes(x = date, y = Freq)) + geom_col(fill = "#6DB6FF") +  
  labs(title = "Cluster 8 (Naive) Frequencies") + theme_minimal()
Cluster_8_gg

Cluster_9 <- Cluster_over_time_filt %>% filter(RNA_snn_res.0.4 == 9)
Cluster_9_gg <- ggplot(Cluster_9, aes(x = date, y = Freq)) + geom_col(fill = "#B6DBFF") +  
  labs(title = "Cluster 9 (IgG+ Plasmablasts) Frequencies") + theme_minimal()
Cluster_9_gg
```


```{r Heatmap Cluster Marker, fig.width=8, fig.height=14}
top10_genes <- Markers_clusters_filt %>% group_by(cluster) %>%  slice_head(n = 10) %>% ungroup()

DoHeatmap(scBcells, features = unique(top10_genes$gene), assay = "RNA", slot = "scale.data") + NoLegend()

DotPlot(scBcells, features = unique(top10_genes$gene), group.by = "RNA_snn_res.0.4") + coord_flip()

```
