---
title: "GSVA and Annotation Epithelial Cells"
author: "Julia Hitschfel"
date: "2025-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(BiocManager)
library(GSVA)
library(GSEABase)
library(Seurat)
library(dplyr)
library(ggplot2)
```

### Import Cell Type Gene Set Collection from MSigDb (C8)

```{r Import Gene Sets}
C8.path <- "D:/Sequencing_data/references/MSigDB_GMTs/c8.all.v2025.1.Hs.symbols.gmt"

C8.gmt <- readGMT(C8.path)

#Filter out small gene sets with fewer than 10 genes
# Read the GMT file
gmt_data <- readLines(C8.path)

# Parse the GMT file and calculate the size of each gene set
gene_set_sizes <- sapply(gmt_data, function(line) {
  genes <- strsplit(line, "\t")[[1]][-c(1, 2)]  # Exclude the first two columns (name and description)
  length(genes)  # Count the number of genes
})

# Create a named vector with gene set names and their sizes
gene_set_names <- sapply(gmt_data, function(line) strsplit(line, "\t")[[1]][1])
names(gene_set_sizes) <- gene_set_names

gene_set_sizes_filt <- gene_set_sizes[gene_set_sizes > 9]

gene_sets_to_keep <- which(names(C8.gmt) %in% names(gene_set_sizes_filt))
C8.gmt.filt <- C8.gmt[gene_sets_to_keep]
```

### Import Seurat Objects

```{r Import Dataset and prep for GSVA}
scEpithelial <- readRDS(file = "Epithelial_Cells.rds")

# Pseudobulking based on Cluster Identity
pseudo_Epithelial <- AggregateExpression(scEpithelial, assays = "RNA", return.seurat = T, 
                                     group.by = "ident")

pseudo_Epithelial_mat <- GetAssayData(pseudo_Epithelial, layer = "data")

# GSVA Analysis
Epithelial_param <- gsvaParam(pseudo_Epithelial_mat, C8.gmt.filt)
Epithelial_gsva <- gsva(Epithelial_param, verbose = F)
```

```{r Check Enrichments for interesting populations, fig.height=12, fig.width=8}
Epithelial_gsva_df <- as.data.frame(Epithelial_gsva)
Epithelial_gsva_df$set.names <- rownames(Epithelial_gsva_df)
  
Epithelial.0.top.10 <- Epithelial_gsva_df %>% slice_max(g0, n = 15)
ggplot(Epithelial.0.top.10, aes(x = g0, y = set.names)) + geom_col()

Epithelial.1.top.10 <- Epithelial_gsva_df %>% slice_max(g1, n = 15)
ggplot(Epithelial.1.top.10, aes(x = g1, y = set.names)) + geom_col()

Epithelial.2.top.10 <- Epithelial_gsva_df %>% slice_max(g2, n = 15)
ggplot(Epithelial.2.top.10, aes(x = g2, y = set.names)) + geom_col()

Epithelial.3.top.10 <- Epithelial_gsva_df %>% slice_max(g3, n = 15)
ggplot(Epithelial.3.top.10, aes(x = g3, y = set.names)) + geom_col()

Epithelial.4.top.10 <- Epithelial_gsva_df %>% slice_max(g4, n = 15)
ggplot(Epithelial.4.top.10, aes(x = g4, y = set.names)) + geom_col()

Epithelial.5.top.10 <- Epithelial_gsva_df %>% slice_max(g5, n = 15)
ggplot(Epithelial.5.top.10, aes(x = g5, y = set.names)) + geom_col()

Epithelial.6.top.10 <- Epithelial_gsva_df %>% slice_max(g6, n = 15)
ggplot(Epithelial.6.top.10, aes(x = g6, y = set.names)) + geom_col()

Epithelial.7.top.10 <- Epithelial_gsva_df %>% slice_max(g7, n = 15)
ggplot(Epithelial.7.top.10, aes(x = g7, y = set.names)) + geom_col()

Epithelial.8.top.10 <- Epithelial_gsva_df %>% slice_max(g8, n = 15)
ggplot(Epithelial.8.top.10, aes(x = g8, y = set.names)) + geom_col()

Epithelial.9.top.10 <- Epithelial_gsva_df %>% slice_max(g9, n = 15)
ggplot(Epithelial.9.top.10, aes(x = g9, y = set.names)) + geom_col()
```

## Violin Plots für Markergene der Epithelzell Subpopulationen 

### Intestinal Stem cells 

```{r Stem cell marker}
VlnPlot(scEpithelial, features = "LGR5")
VlnPlot(scEpithelial, features = "ASCL2")
VlnPlot(scEpithelial, features = "VDR")
VlnPlot(scEpithelial, features = "OLFM4")
VlnPlot(scEpithelial, features = "SLC26A3")
```


### Goblet Cell Marker

```{r Goblet cells}
VlnPlot(scEpithelial, features = "TFF3")
```

### INFLARE Marker from PanGI Paper

```{r Inflares}
VlnPlot(scEpithelial, features = "MUC6")
VlnPlot(scEpithelial, features = "MUC5AC")
VlnPlot(scEpithelial, features = "CEACAM7")
VlnPlot(scEpithelial, features = "CEACAM1")
VlnPlot(scEpithelial, features = "BPIFB1")
VlnPlot(scEpithelial, features = "PGC")
VlnPlot(scEpithelial, features = "AQP5")
VlnPlot(scEpithelial, features = "MKI67")
```

### Paneth Cells

```{r Paneth Cell Secreted Molecules and marker genes}
# Secreted
VlnPlot(scEpithelial, features = "EGFR")
VlnPlot(scEpithelial, features = "TGFA")
VlnPlot(scEpithelial, features = "WNT3")

# Marker
VlnPlot(scEpithelial, features = "DEFA5")
VlnPlot(scEpithelial, features = "DEFA6")
VlnPlot(scEpithelial, features = "REG3A")

```


### Intestinal Mesenchymal Cells

```{r IMCs}
VlnPlot(scEpithelial, features = "PDGFRA")
VlnPlot(scEpithelial, features = "GLI1")
VlnPlot(scEpithelial, features = "FOXL1")
```


### Chemokines

```{r Chemokines}
VlnPlot(scEpithelial, features = "CXCL16")
VlnPlot(scEpithelial, features = "CXCL2")
VlnPlot(scEpithelial, features = "CXCL3")
VlnPlot(scEpithelial, features = "CXCL17")
VlnPlot(scEpithelial, features = "CXCL1")
```

## Marker for recovery of gut epithelia

```{r Recovery}
VlnPlot(scEpithelial, features = "VDR", group.by = "date")
VlnPlot(scEpithelial, features = "CXCL16", group.by = "date")
VlnPlot(scEpithelial, features = "MUC6", group.by = "date")
VlnPlot(scEpithelial, features = "MUC5AC", group.by = "date")

```

## Comparison to markers from Sarah Teichmann's GI Atlas

Im vorherigen Skript von der Analyse wo wir den 15.10.24 noch mit eingeschlossen hatten habe ich mit dem folgenden Code die Marker Gene der Zellannotationen aus Sarah Teichmann's Pan GI Atlas exportiert: 
```{r }
PanGI.Atlas <- readRDS("/references/1_Healthy_Pan-GI_atlas_all_lineages_20241119.rds")

PanGI.Epithelial <- subset(PanGI.Atlas, subset = level_1_annot == "Epithelial")

Idents(PanGI.Epithelial) <- "level_2_annot"

EC_Markers <- FindAllMarkers(PanGI.Epithelial, logfc.threshold = 0.25, min.pct = 0.25, only.pos = T)

write.csv(EC_Markers, "Pan_GI_Atlas_Marker_Genes_lv2.csv")
```


### Absorptive Cell Marker

```{r Absorptive Cells}
VlnPlot(scEpithelial, features = "FABP1")
VlnPlot(scEpithelial, features = "CDHR5")
VlnPlot(scEpithelial, features = "PHGR1")
VlnPlot(scEpithelial, features = "ANPEP")
VlnPlot(scEpithelial, features = "IL32")
```


### Basal Cells
```{r Basal}
VlnPlot(scEpithelial, features = "KRT15")
VlnPlot(scEpithelial, features = "S100A2")
VlnPlot(scEpithelial, features = "CXCL14")
VlnPlot(scEpithelial, features = "COL17A1")
VlnPlot(scEpithelial, features = "IFITM3")
```


### Cycling Epithelial 
```{r Cycling}
VlnPlot(scEpithelial, features = "STMN1")
VlnPlot(scEpithelial, features = "PCLAF")
VlnPlot(scEpithelial, features = "PTTG1")
```


### Enteroendokrine 
```{r enteroendokrine}
VlnPlot(scEpithelial, features = "CHGA")
VlnPlot(scEpithelial, features = "PCSK1N")
VlnPlot(scEpithelial, features = "SCGN")
```


### Epithelial Progenitor
```{r progenitor}
VlnPlot(scEpithelial, features = "APOE")
VlnPlot(scEpithelial, features = "APOA1")
VlnPlot(scEpithelial, features = "GPC3")
```

### Sekretorische Zellen 
```{r secretory}
VlnPlot(scEpithelial, features = "ITLN1")
VlnPlot(scEpithelial, features = "SPINK4")
```

### Transit Amplifying
```{r transit}
VlnPlot(scEpithelial, features = "RARRES2")
VlnPlot(scEpithelial, features = "MYB")
VlnPlot(scEpithelial, features = "AQP1")
VlnPlot(scEpithelial, features = "NEB")
```


```{r Rename Idents}
scEpithelial <- RenameIdents(scEpithelial, 
                             '0' = 'Goblet Cells',
                             '1' = 'Enterocytes_1',
                             '2' = 'Enterocytes_2',
                             '3' = 'Inflammatory Epithelial',
                             '4' = 'Cycling Epithelial',
                             '5' = 'Basal Cells',
                             '6' = 'Immature Enterocytes', 
                             '7' = 'Stromal Cells',
                             '8' = 'Enteroendocrine', 
                             '9' = 'BEST4+ Enterocytes')

saveRDS(scEpithelial, file = "Epithelial_Cells_annotated.rds")
```
