---
title: "DESeq2"
author: "Julia Hitschfel"
date: "2026-01-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Seurat)
library(dplyr)
library(DESeq2)
library(ggplot2) 
library(clusterProfiler)
library(tidyverse)
library(msigdbr)
library(ggforce)
library(ggrepel)
library(tidytext)
library(plotly)
```

## DESeq2 Pseudobulk timepoints

In this script I am generating pseudobulk datensets for the different celltypes and timepoint to do a differential gene expression analysis with DESeq2. The comparison will be between the three timepoints pre CART cell therapy to the four timepoints post CART cell therapy. 
This will be done for the total dataset first and then for the subclustering datasets of B cells, T cells, and epithelial cells.  


```{r Read in Data}
all.cells.seurat <- readRDS("integrated_seurat_refined_annotations.rds")

all.cells.pseudo <- AggregateExpression(all.cells.seurat, assays = "RNA", 
                                        group.by = c("date", "Annotations.0.4"), slot = "counts",
                                        return.seurat = F)

# Manipulate the matrix to fit DESeq2 input
matrix.t <- t(all.cells.pseudo$RNA)
count.data.frame <- as.data.frame(matrix.t)
splitRows <- gsub('.*_', '', rownames(count.data.frame))
cell.type.list <- split.data.frame(count.data.frame, f = factor(splitRows))

#Retranspose the matrix and fix column names for all elements of the list
cell.type.list.t <- lapply(cell.type.list, function(x){
	rownames(x) <- gsub('(.*)_.*', '\\1', rownames(x))
	t(x)
})

# Create meta.data for count matrix
col.data <- data.frame(samples = colnames(cell.type.list.t$`T cells-1`))

col.data$treatment <- ifelse(col.data$samples == "g2024-07-15" | col.data$samples == "g2024-11-28" |
                               col.data$samples == "g2025-02-17", "pre_CART", "post_CART")

# Perform DESeq2 for Tcells-1
count.data.Tcells1 <- cell.type.list.t$`T cells-1`
class(count.data.Tcells1) <- "integer"


dds <- DESeqDataSetFromMatrix(countData = count.data.Tcells1, 
				colData =  col.data,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds)

resultsNames(dds)

#Generate results object
res <- results(dds, name = "treatment_pre_CART_vs_post_CART")
res 

res_df <- as.data.frame(res)

# Sort dataframe for adjusted p value and show top20 results 
res_df_sort <- res_df[order(res_df$padj),]
head(res_df_sort, 20)
```


```{r DESeq2 for Tcells2}
# Perform DESeq2 for Tcells-2
count.data.Tcells2 <- cell.type.list.t$`T cells-2`
class(count.data.Tcells2) <- "integer"


dds.Tcells2 <- DESeqDataSetFromMatrix(countData = count.data.Tcells2, 
				colData =  col.data,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds.Tcells2)) >= 10
dds.Tcells2 <- dds.Tcells2[keep,]

dds.Tcells2 <- DESeq(dds.Tcells2)

resultsNames(dds.Tcells2)

#Generate results object
res.Tcells2 <- results(dds.Tcells2, name = "treatment_pre_CART_vs_post_CART")
res.Tcells2 

res.Tcells2_df <- as.data.frame(res.Tcells2)

# Sort dataframe for adjusted p value and show top20 results 
res.Tcells2_df_sort <- res.Tcells2_df[order(res.Tcells2_df$padj),]
head(res.Tcells2_df_sort, 20)
```


### DESeq2 for the filtered datasets

Hier verwenden wir nochmal DESeq2 aber auf den einzelnen T Zell und Epithelzell Datensätzen, um zu schauen ob es dort weniger Einfluss von den B Zell Transkripten gibt: 

```{r DESeq2 for T cell data}

Tcells <- readRDS("Tcells_CCAintegration_annotation_TCRs.rds")

Tcells.pseudo <- AggregateExpression(Tcells, assays = "RNA", 
                                        group.by = "date", slot = "counts",
                                        return.seurat = F)

# Create meta.data for count matrix
Tcell.numbers <- Tcells@meta.data %>% group_by(date) %>% summarise(Tcell.count = n())
Tcell.numbers$date <- paste0("g", Tcell.numbers$date)
Tcell.numbers$treatment <- ifelse(Tcell.numbers$date == "g2024_07_15" | 
                                    Tcell.numbers$date == "g2024_11_28" |
                               Tcell.numbers$date == "g2025_02_17", "pre_CART", "post_CART")

# Visualize cell counts
ggplot(Tcell.numbers, aes(x=date, y=Tcell.count, fill=treatment)) +
    geom_bar(stat="identity", color="black") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(x="Timepoint", y="Number of T cells") +
    geom_text(aes(label=Tcell.count), vjust=-0.5)

# Perform DESeq2 for Tcell pseudobulk
count.data.Tcells <- as.matrix(Tcells.pseudo$RNA)
IG.transcripts <- c(which(grepl("^IGH", rownames(count.data.Tcells)) == T), 
                    which(grepl("^IGL", rownames(count.data.Tcells)) == T), 
                    which(grepl("^IGK", rownames(count.data.Tcells)) == T))
count.data.Tcells.noIG <- count.data.Tcells[-IG.transcripts,]
class(count.data.Tcells.noIG) <- "integer"


dds.Tcells <- DESeqDataSetFromMatrix(countData = count.data.Tcells.noIG, 
				colData =  Tcell.numbers,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds.Tcells)) >= 10
dds.Tcells <- dds.Tcells[keep,]

dds.Tcells <- DESeq(dds.Tcells)

resultsNames(dds.Tcells)

#Generate results object
res.Tcells <- results(dds.Tcells, name = "treatment_pre_CART_vs_post_CART")
res.Tcells 

res.Tcells_df <- as.data.frame(res.Tcells)

# Sort dataframe for adjusted p value and show top20 results 
res.Tcells_df_sort <- res.Tcells_df[order(res.Tcells_df$padj),]
head(res.Tcells_df_sort, 20)

dir.create(file.path("DESeq2"))
write.csv(res.Tcells_df_sort, "/DESeq2/DEGs_Tzellen.csv")
```

## DESeq2 for Epithelial cells

```{r DESeq2 for Epithelial Cells}
ECs <- readRDS("Epithelial_Cells_annotated.rds")

ECs.pseudo <- AggregateExpression(ECs, assays = "RNA", 
                                        group.by = "date", slot = "counts",
                                        return.seurat = F)

# Create meta.data for count matrix
ECs.numbers <- ECs@meta.data %>% group_by(date) %>% summarise(ECs.count = n())
ECs.numbers$date <- paste0("g", ECs.numbers$date)
ECs.numbers$treatment <- ifelse(ECs.numbers$date == "g2024_07_15" | 
                                    ECs.numbers$date == "g2024_11_28" |
                               ECs.numbers$date == "g2025_02_17", "pre_CART", "post_CART")

# Visualize cell counts
ggplot(ECs.numbers, aes(x=date, y=ECs.count, fill=treatment)) +
    geom_bar(stat="identity", color="black") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(x="Timepoint", y="Number of Epithelial cells") +
    geom_text(aes(label=ECs.count), vjust=-0.5)

# Perform DESeq2 for ECs pseudobulk
count.data.ECs <- as.matrix(ECs.pseudo$RNA)

IG.transcripts <- c(which(grepl("^IGH", rownames(count.data.ECs)) == T), 
                    which(grepl("^IGL", rownames(count.data.ECs)) == T), 
                    which(grepl("^IGK", rownames(count.data.ECs)) == T))
count.data.ECs.noIG <- count.data.ECs[-IG.transcripts,]

class(count.data.ECs.noIG) <- "integer"


dds.ECs <- DESeqDataSetFromMatrix(countData = count.data.ECs.noIG, 
				colData =  ECs.numbers,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds.ECs)) >= 10
dds.ECs <- dds.ECs[keep,]

dds.ECs <- DESeq(dds.ECs)

resultsNames(dds.ECs)

#Generate results object
res.ECs <- results(dds.ECs, name = "treatment_pre_CART_vs_post_CART")
res.ECs 

res.ECs_df <- as.data.frame(res.ECs)

# Sort dataframe for adjusted p value and show top20 results 
res.ECs_df_sort <- res.ECs_df[order(res.ECs_df$padj),]
head(res.ECs_df_sort, 20)

write.csv(res.ECs_df_sort, "/DESeq2/DEGs_Epithelzellen.csv")
```

### Volcano Plot for DEGs Epithelial cells

```{r Create Volcano Plot to explore genes}
EC.volcano.data <- res.ECs_df
EC.volcano.data$log10.padj <- -log10(EC.volcano.data$padj)
EC.volcano.data$diffexpressed <- ifelse(EC.volcano.data$log10.padj > -log10(0.05) & 
                                        abs(EC.volcano.data$log2FoldChange) > 1, "Up or Down", "NS")
EC.volcano.data$gene <- rownames(EC.volcano.data)

# Volcano plot
EC_volcano_plot <- ggplot(EC.volcano.data, aes(x = log2FoldChange, y = log10.padj, 
                       color = diffexpressed, text = gene)) +
  geom_point(size = 1.5) +
  scale_color_manual(values = c("lightgrey", "darkblue")) + 
  labs(title = "Volcano Plot", x = "Log2(Fold Change)", y = "-log10(adj. P-value)") + 
  theme_minimal()
  
ggplotly(EC_volcano_plot, tooltip = "text")
```

### Volcano with categories

```{r Labeled volcano plot}
Inflammation.gene.list <- c("TNF", "IL1A", "IL1B", "IL18", "IL23A", "IL32", "CSF2", "CXCL1", "CXCL3", "CXCL11", "CCL25")
Secretory.function.gene.list <- c("WFDC2", "CLCA1", "ZG16", "B4GALNT2", "GP2")
Barrier.dysfunction.gene.list <- c("CLDN1", "CLDN2", "DEFB4A", "LTF", "GPX2", "S100A2", "S100A3", "SOD3", "CA9")
Mucosal.barrier.gene.list <- c("CLDN15", "ITLN2", "ZG16", "B3GNT6", "ST6GALNAC1", "ST6GALNAC3", "LGALS2")

EC.volcano.data$Category <- ifelse(EC.volcano.data$gene %in% Inflammation.gene.list, "Inflammation/Leukocyte Recruitment", "Other")
EC.volcano.data$Category <- ifelse(EC.volcano.data$gene %in% Secretory.function.gene.list, "Secretory Function", EC.volcano.data$Category)
EC.volcano.data$Category <- ifelse(EC.volcano.data$gene %in% Barrier.dysfunction.gene.list, "Barrier Dysfunction", EC.volcano.data$Category)
EC.volcano.data$Category <- ifelse(EC.volcano.data$gene %in% Mucosal.barrier.gene.list, "Mucosal Barrier Restoration", EC.volcano.data$Category)
EC.volcano.data$Category <- factor(EC.volcano.data$Category, levels = c("Barrier Dysfunction", "Inflammation/Leukocyte Recruitment", 
                                                                        "Mucosal Barrier Restoration", "Secretory Function", "Other"))

points_to_label <- subset(EC.volcano.data, Category != "Other")

# Volcano plot
EC_volcano_plot <- ggplot(EC.volcano.data, aes(x = log2FoldChange, y = log10.padj)) +
  geom_point(data = subset(EC.volcano.data, Category == "Other"), color = "snow2", size = 1) +
  geom_point(data = subset(EC.volcano.data, Category == "Barrier Dysfunction"), aes(color = 'Barrier Dysfunction'), size = 1.25) +
  geom_point(data = subset(EC.volcano.data, Category == "Inflammation/Leukocyte Recruitment"), 
             aes(color = 'Inflammation/Leukocyte Recruitment'), size = 1.25) +
  geom_point(data = subset(EC.volcano.data, Category == "Mucosal Barrier Restoration"), 
             aes(color = 'Mucosal Barrier Restoration'), size = 1.25) +
  geom_point(data = subset(EC.volcano.data, Category == "Secretory Function"), aes(color = 'Secretory Function'), size = 1.25) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "cornsilk2") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "cornsilk2")  +
  geom_vline(xintercept = -1, linetype = "dashed", color = "cornsilk2") +
  geom_text_repel(data = points_to_label, aes(label = gene), color = "black", size = 3) +
  coord_fixed(ratio = 3.44) +
      scale_color_manual(name = 'Category', 
                       breaks = c('Barrier Dysfunction', 'Inflammation/Leukocyte Recruitment', 'Mucosal Barrier Restoration', 
                                  'Secretory Function'), 
                       values = c('Barrier Dysfunction' = "darkviolet", 'Mucosal Barrier Restoration' = "dodgerblue1", 
                                  'Inflammation/Leukocyte Recruitment' ="firebrick1", 'Secretory Function' = "chartreuse4")) +
  labs(x = expression("Log"[2] ~ "(Fold Change)"), y = expression("-log"[10] ~ "(adj. P-value)")) + 
  theme_classic()

EC_volcano_plot

dir.create(file.path("Figures"))
ggsave("/Figures/Volcano_Epithelzellen_Labels.tiff", plot = EC_volcano_plot,
       device = "tiff", dpi = 300, width = 8, height = 6, units = "in", compression = "lzw")
```



## Pre-ranked GSEA with clusterProfiler

Since for T cells the ranking with -log10(adj. p value) creates too many duplicate values, -log10(p value) will be used instead for ranking

### GSEA for T cells

```{r GSEA mir clusterProfiler}
res.Tcells_df_sort$sign.log.10.p <-
  sign(res.Tcells_df_sort$log2FoldChange)*(-log10(res.Tcells_df_sort$pvalue))
res.Tcells_df_sort <- res.Tcells_df_sort %>% filter(!is.na(sign.log.10.p))
res.Tcells_df_sort <- res.Tcells_df_sort[order(res.Tcells_df_sort$sign.log.10.p, decreasing = T),]

ranked_genes_T <- as.vector(res.Tcells_df_sort$sign.log.10.p)
names(ranked_genes_T) <- rownames(res.Tcells_df_sort)

# Import the MSigDB Hallmark Collection
hallmark<- msigdbr(collection = "H") %>% 
  dplyr::select(gs_name, gene_symbol)

head(hallmark)

h2 <- msigdbr(collection = "H") %>% 
  dplyr::select(gs_name, gs_description)

#set seed
set.seed(123)
#run GSEA
Tcell.HALLMARK <- GSEA(ranked_genes_T, TERM2GENE = hallmark, TERM2NAME = h2, seed=TRUE)

head(Tcell.HALLMARK)

print(paste0(Tcell.HALLMARK@result$ID, " - NES:  ", Tcell.HALLMARK@result$NES,  " - padj:  ",
             Tcell.HALLMARK@result$p.adjust))

Tcell.Hallmark.df <- data.frame(ID = Tcell.HALLMARK@result$ID, 
                             NES = Tcell.HALLMARK@result$NES, 
                             padj = Tcell.HALLMARK@result$p.adjust)

Tcell.Hallmark.df <- Tcell.Hallmark.df[order(Tcell.Hallmark.df$NES),]
Tcell.Hallmark.df$ID <- factor(Tcell.Hallmark.df$ID, levels = Tcell.Hallmark.df$ID)

ggplot(Tcell.Hallmark.df, aes(x = ID, y = NES, fill = padj)) + geom_col() + 
  scale_fill_gradient() + coord_flip() + theme_minimal()

```

### GSEA for T cells with GO and REACTOME and select Gene Sets

```{r GSEA pre-ranked mit GO und REACTOME, fig.width=10, fig.height=8}
GO_BP <- msigdbr(collection = "C5", subcollection = "GO:BP") %>% 
  dplyr::select(gs_name, gene_symbol)

GO_BP_t2n <- msigdbr(collection = "C5", subcollection = "GO:BP") %>% 
  dplyr::select(gs_name, gs_description)

#run GSEA
Tcell.GO.BP <- GSEA(ranked_genes_T, TERM2GENE = GO_BP, TERM2NAME = GO_BP_t2n, seed=TRUE)

Tcell.GO.BP.df <- data.frame(ID = Tcell.GO.BP@result$ID, 
                             NES = Tcell.GO.BP@result$NES, 
                             padj = Tcell.GO.BP@result$p.adjust)

Tcell.GO.BP.df.top10.up <- Tcell.GO.BP.df %>% filter(NES > 0) %>% slice_min(padj, n = 10)
Tcell.GO.BP.df.top10.dn <- Tcell.GO.BP.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
Tcell.GO.BP.df.top20 <- rbind(Tcell.GO.BP.df.top10.up, Tcell.GO.BP.df.top10.dn)

Tcell.GO.BP.df.top20$ID <- ifelse(Tcell.GO.BP.df.top20$ID ==
 "GOBP_ADAPTIVE_IMMUNE_RESPONSE_BASED_ON_SOMATIC_RECOMBINATION_OF_IMMUNE_RECEPTORS_BUILT_FROM_IMMUNOGLOBULIN_SUPERFAMILY_DOMAINS", 
 "GOBP_SOMATIC_RECOMBINATION_OF_IMMUNE_RECEPTORS", Tcell.GO.BP.df.top20$ID)

Tcell.GO.BP.df.top20$ID <- ifelse(Tcell.GO.BP.df.top20$ID ==
"GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_EXOGENOUS_PEPTIDE_ANTIGEN_VIA_MHC_CLASS_II",
 "GOBP_PRESENTATION_OF_EXOGENOUS_PEPTIDE_ANTIGEN_VIA_MHC_CLASS_II", Tcell.GO.BP.df.top20$ID)

Tcell.GO.BP.df.top20 <- Tcell.GO.BP.df.top20[order(Tcell.GO.BP.df.top20$NES),]
Tcell.GO.BP.df.top20$ID <- factor(Tcell.GO.BP.df.top20$ID, levels = Tcell.GO.BP.df.top20$ID)

ggplot(Tcell.GO.BP.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() + 
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "GO:BP Top 20 enriched Gene Sets in T cells")


## GSEA mit REACTOME
REACTOME <- msigdbr(collection = "C2", subcollection = "CP:REACTOME") %>% 
  dplyr::select(gs_name, gene_symbol)

REACTOME_t2n <- msigdbr(collection = "C2", subcollection = "CP:REACTOME") %>% 
  dplyr::select(gs_name, gs_description)

#run GSEA
Tcell.REACTOME <- GSEA(ranked_genes_T, TERM2GENE = REACTOME, TERM2NAME = REACTOME_t2n, seed=TRUE)

Tcell.REACTOME.df <- data.frame(ID = Tcell.REACTOME@result$ID, 
                             NES = Tcell.REACTOME@result$NES, 
                             padj = Tcell.REACTOME@result$p.adjust)

Tcell.REACTOME.df.top10.up <- Tcell.REACTOME.df %>% filter(NES > 0) %>% slice_min(padj, n = 10) %>%
  slice_max(NES, n = 10)
Tcell.REACTOME.df.top10.dn <- Tcell.REACTOME.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
Tcell.REACTOME.df.top20 <- rbind(Tcell.REACTOME.df.top10.up, Tcell.REACTOME.df.top10.dn)

Tcell.REACTOME.df.top20$ID <- ifelse(Tcell.REACTOME.df.top20$ID ==
  "REACTOME_IMMUNOREGULATORY_INTERACTIONS_BETWEEN_A_LYMPHOID_AND_A_NON_LYMPHOID_CELL",
  "REACTOME_IMMUNOREGULATORY_INTERACTIONS_LYMPHOID_AND_NON_LYMPHOID_CELL", Tcell.REACTOME.df.top20$ID)

Tcell.REACTOME.df.top20$ID <- ifelse(Tcell.REACTOME.df.top20$ID ==
    "REACTOME_ACTIVATION_OF_THE_MRNA_UPON_BINDING_OF_THE_CAP_BINDING_COMPLEX_AND_EIFS_AND_SUBSEQUENT_BINDING_TO_43S",
    "REACTOME_CAP_DEP_MRNA_ACTIVATION_eIF_AND_43S_BINDING", Tcell.REACTOME.df.top20$ID)

Tcell.REACTOME.df.top20 <- Tcell.REACTOME.df.top20[order(Tcell.REACTOME.df.top20$NES),]
Tcell.REACTOME.df.top20$ID <- factor(Tcell.REACTOME.df.top20$ID, levels = Tcell.REACTOME.df.top20$ID)

ggplot(Tcell.REACTOME.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "REACTOME Top 20 enriched Gene Sets in T cells")

```

```{r Bind results and create dotplot, fig.width=8, fig.height=8}
all.GSEA.T <- rbind(Tcell.Hallmark.df, Tcell.GO.BP.df.top20, Tcell.REACTOME.df.top20)
all.GSEA.T$DB <- ifelse(grepl("HALLMARK", all.GSEA.T$ID), "HALLMARK", NA)
all.GSEA.T$DB <- ifelse(grepl("GOBP", all.GSEA.T$ID), "GOBP", all.GSEA.T$DB)
all.GSEA.T$DB <- ifelse(grepl("REACTOME", all.GSEA.T$ID), "REACTOME", all.GSEA.T$DB)
all.GSEA.T$DB <- factor(all.GSEA.T$DB, levels = c("HALLMARK", "GOBP", "REACTOME"))

all.GSEA.T$Status <- ifelse(all.GSEA.T$NES < 0, "post CART", NA)
all.GSEA.T$Status <- ifelse(all.GSEA.T$NES > 0, "pre CART", all.GSEA.T$Status)

all.GSEA.T.preCAR <- all.GSEA.T %>% filter(Status == "pre CART") %>% group_by(DB) %>% slice_max(NES, n = 5)
all.GSEA.T.postCAR <- all.GSEA.T %>% filter(Status == "post CART") %>% group_by(DB) %>% slice_min(NES, n = 5)
all.GSEA.T.top5 <- rbind(all.GSEA.T.preCAR, all.GSEA.T.postCAR)
all.GSEA.T.top5$Status <- factor(all.GSEA.T.top5$Status, levels = c("pre CART", "post CART"))

all.GSEA.T.top5$ID <- gsub("GOBP", "", all.GSEA.T.top5$ID)
all.GSEA.T.top5$ID <- gsub("REACTOME", "", all.GSEA.T.top5$ID)
all.GSEA.T.top5$ID <- gsub("HALLMARK", "", all.GSEA.T.top5$ID)
all.GSEA.T.top5$ID <- gsub("_", " ", all.GSEA.T.top5$ID)

all.GSEA.T.top5 <- all.GSEA.T.top5 %>% mutate(log.10.padj = -log10(padj))

All.GSEA.T.top5.gg <- ggplot2::ggplot(all.GSEA.T.top5, aes(x = Status, y = reorder_within(ID, NES, DB))) + 
  geom_point(aes(color = log.10.padj, size = abs(NES))) + scale_y_reordered() + scale_size(range = c(3,6)) +
  scale_x_discrete(expand = c(0.5,0.5)) +
  ggforce::facet_col(~DB, scales = "free_y", space = "free") +
   scale_color_gradient(low = "darkblue", high = "purple") + 
  labs(color = "-log10\n(adj. P-value)", size = "NES") +
  theme_classic(base_size = 13) +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_text(size = 10),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "right", 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

All.GSEA.T.top5.gg

ggplot2::ggsave("/Figures/GSEA_summary_Tcells.tiff", 
       plot = All.GSEA.T.top5.gg, device = "tiff", dpi = 300, width = 8, height = 8, units = "in", compression = "lzw")
```


### GSEA for epithelial cells 

```{r GSEA clusterProfiler Epithelzellen, fig.width = 10, fig.height=8}
res.ECs_df_sort$sign.log.10.p <- sign(res.ECs_df_sort$log2FoldChange)*(-log10(res.ECs_df_sort$pvalue))
res.ECs_df_sort <- res.ECs_df_sort %>% filter(!is.na(sign.log.10.p))
res.ECs_df_sort <- res.ECs_df_sort[order(res.ECs_df_sort$sign.log.10.p, decreasing = T),]

ranked_genes_EC <- as.vector(res.ECs_df_sort$sign.log.10.p)
names(ranked_genes_EC) <- rownames(res.ECs_df_sort)

#run GSEA
EC.HALLMARK <- GSEA(ranked_genes_EC, TERM2GENE = hallmark, TERM2NAME = h2, seed=TRUE)

head(EC.HALLMARK)

print(paste0(EC.HALLMARK@result$ID, " - NES:  ", EC.HALLMARK@result$NES,  " - padj:  ",
             EC.HALLMARK@result$p.adjust))

EC.Hallmark.df <- data.frame(ID = EC.HALLMARK@result$ID, 
                             NES = EC.HALLMARK@result$NES, 
                             padj = EC.HALLMARK@result$p.adjust)

EC.Hallmark.df <- EC.Hallmark.df[order(EC.Hallmark.df$NES),]
EC.Hallmark.df$ID <- factor(EC.Hallmark.df$ID, levels = EC.Hallmark.df$ID)

EC.Hallmark.df.top10.up <- EC.Hallmark.df %>% filter(NES > 0) %>% slice_min(padj, n = 10)
EC.Hallmark.df.top10.dn <- EC.Hallmark.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
EC.Hallmark.df.top20 <- rbind(EC.Hallmark.df.top10.up, EC.Hallmark.df.top10.dn)

ggplot(EC.Hallmark.df, aes(x = ID, y = NES, fill = padj)) + geom_col() + 
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "HALLMARK Enrichments for Epithelial Cells")

#run GSEA with GO:BP gene sets
EC.GO.BP <- GSEA(ranked_genes_EC, TERM2GENE = GO_BP, TERM2NAME = GO_BP_t2n, seed=TRUE)

# head(EC.GO.BP, n = 20)

# print(paste0(EC.GO.BP@result$ID, " - NES:  ", EC.GO.BP@result$NES,  " - padj:  ",
#              EC.GO.BP@result$p.adjust))

EC.GO.BP.df <- data.frame(ID = EC.GO.BP@result$ID, 
                             NES = EC.GO.BP@result$NES, 
                             padj = EC.GO.BP@result$p.adjust)

EC.GO.BP.df.top10.up <- EC.GO.BP.df %>% filter(NES > 0) %>% slice_min(padj, n = 10)
EC.GO.BP.df.top10.dn <- EC.GO.BP.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
EC.GO.BP.df.top20 <- rbind(EC.GO.BP.df.top10.up, EC.GO.BP.df.top10.dn)

EC.GO.BP.df.top20$ID <- ifelse(EC.GO.BP.df.top20$ID ==
 "GOBP_ADAPTIVE_IMMUNE_RESPONSE_BASED_ON_SOMATIC_RECOMBINATION_OF_IMMUNE_RECEPTORS_BUILT_FROM_IMMUNOGLOBULIN_SUPERFAMILY_DOMAINS",
 "GOBP_SOMATIC_RECOMBINATION_OF_IMMUNE_RECEPTORS", EC.GO.BP.df.top20$ID)

EC.GO.BP.df.top20 <- EC.GO.BP.df.top20[order(EC.GO.BP.df.top20$NES),]
EC.GO.BP.df.top20$ID <- factor(EC.GO.BP.df.top20$ID, levels = EC.GO.BP.df.top20$ID)

ggplot(EC.GO.BP.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() + 
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "GO:BP top 20 enriched gene sets in epithelial cells")


#run GSEA with REACTOME on epithelial cells
EC.REACTOME <- GSEA(ranked_genes_EC, TERM2GENE = REACTOME, TERM2NAME = REACTOME_t2n, seed=TRUE)

EC.REACTOME.df <- data.frame(ID = EC.REACTOME@result$ID, 
                             NES = EC.REACTOME@result$NES, 
                             padj = EC.REACTOME@result$p.adjust)

EC.REACTOME.df.top10.up <- EC.REACTOME.df %>% filter(NES > 0) %>% slice_min(padj, n = 10)
EC.REACTOME.df.top10.dn <- EC.REACTOME.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
EC.REACTOME.df.top20 <- rbind(EC.REACTOME.df.top10.up, EC.REACTOME.df.top10.dn)

EC.REACTOME.df.top20$ID <- ifelse(EC.REACTOME.df.top20$ID ==
  "REACTOME_SRP_DEPENDENT_COTRANSLATIONAL_PROTEIN_TARGETING_TO_MEMBRANE",
  "REACTOME_SRP_DEPENDENT_PROTEIN_TARGETING_TO_MEMBRANE", EC.REACTOME.df.top20$ID)

EC.REACTOME.df.top20 <- EC.REACTOME.df.top20[order(EC.REACTOME.df.top20$NES),]
EC.REACTOME.df.top20$ID <- factor(EC.REACTOME.df.top20$ID, levels = EC.REACTOME.df.top20$ID)

ggplot(EC.REACTOME.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "REACTOME top 20 enriched gene sets in epithelial cells")

## GSEA with selected datasets
mucosal_healing <- 
  read.gmt("/references/MSigDB_GMTs/custom_genesets_mucosal_healing.gmt")

Fenton_UC_up <- read.gmt("/references/MSigDB_GMTs/custom_genesets_Fenton_UC.gmt")

Fenton_Control_up <- 
  read.gmt("/references/MSigDB_GMTs/custom_genesets_Fenton_Control.gmt")

Fenton_Remission_up <- 
  read.gmt("/references/MSigDB_GMTs/custom_genesets_Fenton_Remission.gmt")

#run GSEA with custom mucosal healing gene sets
EC.mucosal.healing <- GSEA(ranked_genes_EC, TERM2GENE = mucosal_healing, seed=TRUE)

head(EC.mucosal.healing, n = 20)

print(paste0(EC.mucosal.healing@result$ID, " - NES:  ",
              EC.mucosal.healing@result$NES,  " - padj:  ",
             EC.mucosal.healing@result$p.adjust))

EC.mucosal.healing.df <- data.frame(ID = EC.mucosal.healing@result$ID, 
                             NES = EC.mucosal.healing@result$NES, 
                             padj = EC.mucosal.healing@result$p.adjust)

ggplot(EC.mucosal.healing.df, aes(x = ID, y = NES, fill = padj)) + geom_col() + ylim(-2,2) +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "Mucosal healing enriched gene sets in epithelial cells")


#run GSEA with Fenton et al. UC gene set
EC.Fenton.UC <- GSEA(ranked_genes_EC, TERM2GENE = Fenton_UC_up, seed=TRUE)

head(EC.Fenton.UC, n = 20)

print(paste0(EC.Fenton.UC@result$ID, " - NES:  ",
              EC.Fenton.UC@result$NES,  " - padj:  ",
             EC.Fenton.UC@result$p.adjust))

EC.Fenton.UC.df <- data.frame(ID = EC.Fenton.UC@result$ID, 
                             NES = EC.Fenton.UC@result$NES, 
                             padj = EC.Fenton.UC@result$p.adjust)

ggplot(EC.Fenton.UC.df, aes(x = ID, y = NES, fill = padj)) + geom_col() + ylim(-2.5,2.5) + 
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "Gene sets enriched in UC from Fenton et al. in epithelial cells")


#run GSEA with Fenton et al. UC in remission gene set
EC.Fenton.Remission <- GSEA(ranked_genes_EC, TERM2GENE = Fenton_Remission_up, seed=TRUE)

head(EC.Fenton.Remission, n = 20)

print(paste0(EC.Fenton.Remission@result$ID, " - NES:  ",
              EC.Fenton.Remission@result$NES,  " - padj:  ",
             EC.Fenton.Remission@result$p.adjust))

EC.Fenton.Remission.df <- data.frame(ID = EC.Fenton.Remission@result$ID, 
                             NES = EC.Fenton.Remission@result$NES, 
                             padj = EC.Fenton.Remission@result$p.adjust)

ggplot(EC.Fenton.Remission.df, aes(x = ID, y = NES, fill = padj)) + geom_col() + ylim(-2,2) +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "Gene sets enriched in UC remission from Fenton et al. in epithelial cells")


#run GSEA with Fenton et al. Healthy gene set
EC.Fenton.Healthy <- GSEA(ranked_genes_EC, TERM2GENE = Fenton_Control_up, seed=TRUE)

head(EC.Fenton.Healthy, n = 20)

print(paste0(EC.Fenton.Healthy@result$ID, " - NES:  ",
              EC.Fenton.Healthy@result$NES,  " - padj:  ",
             EC.Fenton.Healthy@result$p.adjust))

EC.Fenton.Healthy.df <- data.frame(ID = EC.Fenton.Healthy@result$ID, 
                             NES = EC.Fenton.Healthy@result$NES, 
                             padj = EC.Fenton.Healthy@result$p.adjust)

ggplot(EC.Fenton.Healthy.df, aes(x = ID, y = NES, fill = padj)) + geom_col() + ylim(-2,2) +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "Gene sets enriched in healthy vs UC from Fenton et al. in epithelial cells")
```


```{r GSEA summary plot Epithelzellen, fig.width=8, fig.height=8}
all.GSEA.EC <- rbind(EC.Hallmark.df.top20, EC.GO.BP.df.top20)
all.GSEA.EC$DB <- ifelse(grepl("HALLMARK", all.GSEA.EC$ID), "HALLMARK", NA)
all.GSEA.EC$DB <- ifelse(grepl("GOBP", all.GSEA.EC$ID), "GOBP", all.GSEA.EC$DB)
all.GSEA.EC$DB <- factor(all.GSEA.EC$DB, levels = c("HALLMARK", "GOBP"))

all.GSEA.EC$Status <- ifelse(all.GSEA.EC$NES < 0, "post CART", NA)
all.GSEA.EC$Status <- ifelse(all.GSEA.EC$NES > 0, "pre CART", all.GSEA.EC$Status)
all.GSEA.EC$Status <- factor(all.GSEA.EC$Status, levels = c("pre CART", "post CART"))

all.GSEA.EC$ID <- gsub("GOBP", "", all.GSEA.EC$ID)
all.GSEA.EC$ID <- gsub("HALLMARK", "", all.GSEA.EC$ID)
all.GSEA.EC$ID <- gsub("_", " ", all.GSEA.EC$ID)

all.GSEA.EC <- all.GSEA.EC %>% mutate(log.10.padj = -log10(padj))

All.GSEA.EC.gg <- ggplot2::ggplot(all.GSEA.EC, aes(x = Status, y = reorder_within(ID, NES, DB))) + 
  geom_point(aes(color = log.10.padj, size = abs(NES))) + scale_y_reordered() + scale_size(range = c(3,6)) +
  scale_x_discrete(expand = c(0.5,0.5)) +
  ggforce::facet_col(~DB, scales = "free_y", space = "free") +
   scale_color_gradient(low = "darkblue", high = "purple") + 
  labs(color = "-log10\n(adj. P-value)", size = "NES") +
  theme_classic(base_size = 13) +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_text(size = 10),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "right", 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

All.GSEA.EC.gg

ggplot2::ggsave("/Figures/GSEA_summary_Epithelial_cells.tiff",
       plot = All.GSEA.EC.gg, device = "tiff", dpi = 300, width = 8, height = 8, units = "in", compression = "lzw")
```

## GSEA for TPH and regulatory T cells

```{r DEG und GSEA für T Zell Subsets}
Tcells$Annotation <- Idents(Tcells)
Tcells.subpop.pseudo <- AggregateExpression(Tcells, assays = "RNA", 
                                        group.by = c("date", "Annotation"), slot = "counts",
                                        return.seurat = F)

# Manipulate the matrix to fit DESeq2 input
matrix.t <- t(Tcells.subpop.pseudo$RNA)

count.data.frame <- as.data.frame(matrix.t)
splitRows <- gsub('.*_', '', rownames(count.data.frame))
cell.type.list <- split.data.frame(count.data.frame, f = factor(splitRows))

#Retranspose the matrix and fix column names for all elements of the list
cell.type.list.t <- lapply(cell.type.list, function(x){
	rownames(x) <- gsub('(.*)_.*', '\\1', rownames(x))
 	t(x)
 })

# Create meta.data for count matrix
col.data <- data.frame(samples = colnames(cell.type.list.t$`Treg-1`))

col.data$treatment <- ifelse(col.data$samples == "g2024-07-15" | col.data$samples == "g2024-11-28" |
                               col.data$samples == "g2025-02-17", "pre_CART", "post_CART")

# Perform DESeq2 for Treg1 cells
count.data.Treg1 <- cell.type.list.t$`Treg-1`
IG.transcripts <- c(which(grepl("^IGH", rownames(count.data.Treg1)) == T), 
                    which(grepl("^IGL", rownames(count.data.Treg1)) == T), 
                    which(grepl("^IGK", rownames(count.data.Treg1)) == T))
count.data.Treg1 <- count.data.Treg1[-IG.transcripts,]
class(count.data.Treg1) <- "integer"


dds <- DESeqDataSetFromMatrix(countData = count.data.Treg1, 
				colData =  col.data,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds)

resultsNames(dds)

#Generate results object
res <- results(dds, name = "treatment_pre_CART_vs_post_CART")
res 

res_df <- as.data.frame(res)

# Sort dataframe for adjusted p value and show top20 results 
res_df_sort <- res_df[order(res_df$padj),]
head(res_df_sort, 20)


res_df_sort$sign.log.10.p <-
  sign(res_df_sort$log2FoldChange)*(-log10(res_df_sort$pvalue))
res_df_sort <- res_df_sort %>% filter(!is.na(sign.log.10.p))
res_df_sort <- res_df_sort[order(res_df_sort$sign.log.10.p, decreasing = T),]

ranked_genes_T <- as.vector(res_df_sort$sign.log.10.p)
names(ranked_genes_T) <- rownames(res_df_sort)

Treg1.volcano.data <- res_df_sort
Treg1.volcano.data$log10.pvalue <- -log10(Treg1.volcano.data$pvalue)
Treg1.volcano.data$diffexpressed <- ifelse(Treg1.volcano.data$log10.pvalue > -log10(0.05) & 
                                        abs(Treg1.volcano.data$log2FoldChange) > 1, "Up or Down", "NS")
Treg1.volcano.data$gene <- rownames(Treg1.volcano.data)

# Volcano plot
Treg1_volcano_plot <- ggplot(Treg1.volcano.data, aes(x = log2FoldChange, y = log10.pvalue, 
                       color = diffexpressed, text = gene)) +
  geom_point(size = 1.5) +
  scale_color_manual(values = c("lightgrey", "darkblue")) + 
  labs(title = "Volcano Plot", x = "Log2(Fold Change)", y = "-log10(P-value)") + 
  theme_minimal()

ggplotly(Treg1_volcano_plot, tooltip = "text")
```


```{r Run GSEA for Treg1, fig.height=8, fig.width =10}
#run GSEA
Tcell.GOBP <- GSEA(ranked_genes_T, TERM2GENE = GO_BP, TERM2NAME = GO_BP_t2n, seed=TRUE)

Tcell.GOBP.df <- data.frame(ID = Tcell.GOBP@result$ID, 
                             NES = Tcell.GOBP@result$NES, 
                             padj = Tcell.GOBP@result$p.adjust)

Tcell.GOBP.df.top10.up <- Tcell.GOBP.df %>% filter(NES > 0) %>% slice_min(padj, n = 10) %>%
  slice_max(NES, n = 10)
Tcell.GOBP.df.top10.dn <- Tcell.GOBP.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
Tcell.GOBP.df.top20 <- rbind(Tcell.GOBP.df.top10.up, Tcell.GOBP.df.top10.dn)

Tcell.GOBP.df.top20$ID <- ifelse(Tcell.GOBP.df.top20$ID ==
 "GOBP_POSITIVE_REGULATION_OF_PRODUCTION_OF_MOLECULAR_MEDIATOR_OF_IMMUNE_RESPONSE",
 "GOBP_POS_REGULATION_PRODUCTION_OF_MEDIATORS_IMMUNE_RESPONSE", Tcell.GOBP.df.top20$ID)

Tcell.GOBP.df.top20 <- Tcell.GOBP.df.top20[order(Tcell.GOBP.df.top20$NES),]
Tcell.GOBP.df.top20$ID <- factor(Tcell.GOBP.df.top20$ID, levels = Tcell.GOBP.df.top20$ID)

ggplot(Tcell.GOBP.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "GOBP Top 20 enriched Gene Sets in T cells")


Tcell.REACTOME <- GSEA(ranked_genes_T, TERM2GENE = REACTOME, TERM2NAME = REACTOME_t2n, seed=TRUE)

Tcell.REACTOME.df <- data.frame(ID = Tcell.REACTOME@result$ID, 
                             NES = Tcell.REACTOME@result$NES, 
                             padj = Tcell.REACTOME@result$p.adjust)

Tcell.REACTOME.df.top10.up <- Tcell.REACTOME.df %>% filter(NES > 0) %>% slice_min(padj, n = 10) %>%
  slice_max(NES, n = 10)
Tcell.REACTOME.df.top10.dn <- Tcell.REACTOME.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
Tcell.REACTOME.df.top20 <- rbind(Tcell.REACTOME.df.top10.up, Tcell.REACTOME.df.top10.dn)

Tcell.REACTOME.df.top20 <- Tcell.REACTOME.df.top20[order(Tcell.REACTOME.df.top20$NES),]
Tcell.REACTOME.df.top20$ID <- factor(Tcell.REACTOME.df.top20$ID, levels = Tcell.REACTOME.df.top20$ID)

ggplot(Tcell.REACTOME.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "REACTOME Top 20 enriched Gene Sets in T cells")
```

```{r check for Treg2}
# Create meta.data for count matrix
col.data <- data.frame(samples = colnames(cell.type.list.t$`Treg-2`))

col.data$treatment <- ifelse(col.data$samples == "g2024-07-15" | col.data$samples == "g2024-11-28" |
                               col.data$samples == "g2025-02-17", "pre_CART", "post_CART")

# Perform DESeq2 for Treg2 cells
count.data.Treg2 <- cell.type.list.t$`Treg-2`
IG.transcripts <- c(which(grepl("^IGH", rownames(count.data.Treg2)) == T), 
                    which(grepl("^IGL", rownames(count.data.Treg2)) == T), 
                    which(grepl("^IGK", rownames(count.data.Treg2)) == T))
count.data.Treg2 <- count.data.Treg2[-IG.transcripts,]
class(count.data.Treg2) <- "integer"


dds <- DESeqDataSetFromMatrix(countData = count.data.Treg2, 
				colData =  col.data,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds)

resultsNames(dds)

#Generate results object
res <- results(dds, name = "treatment_pre_CART_vs_post_CART")
res 

res_df <- as.data.frame(res)

# Sort dataframe for adjusted p value and show top20 results 
res_df_sort <- res_df[order(res_df$padj),]
head(res_df_sort, 20)


res_df_sort$sign.log.10.p <-
  sign(res_df_sort$log2FoldChange)*(-log10(res_df_sort$pvalue))
res_df_sort <- res_df_sort %>% filter(!is.na(sign.log.10.p))
res_df_sort <- res_df_sort[order(res_df_sort$sign.log.10.p, decreasing = T),]

ranked_genes_T <- as.vector(res_df_sort$sign.log.10.p)
names(ranked_genes_T) <- rownames(res_df_sort)

# Volcano Plot
Treg2.volcano.data <- res_df_sort
Treg2.volcano.data$log10.pvalue <- -log10(Treg2.volcano.data$pvalue)
Treg2.volcano.data$diffexpressed <- ifelse(Treg2.volcano.data$log10.pvalue > -log10(0.05) & 
                                        abs(Treg2.volcano.data$log2FoldChange) > 1, "Up or Down", "NS")
Treg2.volcano.data$gene <- rownames(Treg2.volcano.data)

# Volcano plot
Treg2_volcano_plot <- ggplot(Treg2.volcano.data, aes(x = log2FoldChange, y = log10.pvalue, 
                       color = diffexpressed, text = gene)) +
  geom_point(size = 1.5) +
  scale_color_manual(values = c("lightgrey", "darkblue")) + 
  labs(title = "Volcano Plot", x = "Log2(Fold Change)", y = "-log10(P-value)") + 
  theme_minimal()

ggplotly(Treg2_volcano_plot, tooltip = "text")

```


```{r Run GSEA for Treg2 cluster, fig.height = 8, fig.width = 10}
#run GSEA
Tcell.GOBP <- GSEA(ranked_genes_T, TERM2GENE = GO_BP, TERM2NAME = GO_BP_t2n, seed=TRUE)

Tcell.GOBP.df <- data.frame(ID = Tcell.GOBP@result$ID, 
                             NES = Tcell.GOBP@result$NES, 
                             padj = Tcell.GOBP@result$p.adjust)

Tcell.GOBP.df.top10.up <- Tcell.GOBP.df %>% filter(NES > 0) %>% slice_min(padj, n = 10) %>%
  slice_max(NES, n = 10)
Tcell.GOBP.df.top10.dn <- Tcell.GOBP.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
Tcell.GOBP.df.top20 <- rbind(Tcell.GOBP.df.top10.up, Tcell.GOBP.df.top10.dn)

Tcell.GOBP.df.top20 <- Tcell.GOBP.df.top20[order(Tcell.GOBP.df.top20$NES),]
Tcell.GOBP.df.top20$ID <- factor(Tcell.GOBP.df.top20$ID, levels = Tcell.GOBP.df.top20$ID)

ggplot(Tcell.GOBP.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "GOBP Top 20 enriched Gene Sets in T cells")

#run GSEA
Tcell.REACTOME <- GSEA(ranked_genes_T, TERM2GENE = REACTOME, TERM2NAME = REACTOME_t2n, seed=TRUE)

Tcell.REACTOME.df <- data.frame(ID = Tcell.REACTOME@result$ID, 
                             NES = Tcell.REACTOME@result$NES, 
                             padj = Tcell.REACTOME@result$p.adjust)

Tcell.REACTOME.df.top10.up <- Tcell.REACTOME.df %>% filter(NES > 0) %>% slice_min(padj, n = 10) %>%
  slice_max(NES, n = 10)
Tcell.REACTOME.df.top10.dn <- Tcell.REACTOME.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
Tcell.REACTOME.df.top20 <- rbind(Tcell.REACTOME.df.top10.up, Tcell.REACTOME.df.top10.dn)

Tcell.REACTOME.df.top20 <- Tcell.REACTOME.df.top20[order(Tcell.REACTOME.df.top20$NES),]
Tcell.REACTOME.df.top20$ID <- factor(Tcell.REACTOME.df.top20$ID, levels = Tcell.REACTOME.df.top20$ID)

ggplot(Tcell.REACTOME.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "REACTOME Top 20 enriched Gene Sets in T cells")
```


```{r Compare TPH pre and post CART}
# Create meta.data for count matrix
col.data <- data.frame(samples = colnames(cell.type.list.t$`TPH-like`))

col.data$treatment <- ifelse(col.data$samples == "g2024-07-15" | col.data$samples == "g2024-11-28" |
                               col.data$samples == "g2025-02-17", "pre_CART", "post_CART")

# Perform DESeq2 for TPH-like cells
count.data.TPH <- cell.type.list.t$`TPH-like`
IG.transcripts <- c(which(grepl("^IGH", rownames(count.data.TPH)) == T), 
                    which(grepl("^IGL", rownames(count.data.TPH)) == T), 
                    which(grepl("^IGK", rownames(count.data.TPH)) == T))
count.data.TPH <- count.data.TPH[-IG.transcripts,]
class(count.data.TPH) <- "integer"


dds <- DESeqDataSetFromMatrix(countData = count.data.TPH, 
				colData =  col.data,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds)

resultsNames(dds)

#Generate results object
res <- results(dds, name = "treatment_pre_CART_vs_post_CART")
res 

res_df <- as.data.frame(res)

# Sort dataframe for adjusted p value and show top20 results 
res_df_sort <- res_df[order(res_df$padj),]
head(res_df_sort, 20)


res_df_sort$sign.log.10.p <-
  sign(res_df_sort$log2FoldChange)*(-log10(res_df_sort$pvalue))
res_df_sort <- res_df_sort %>% filter(!is.na(sign.log.10.p))
res_df_sort <- res_df_sort[order(res_df_sort$sign.log.10.p, decreasing = T),]

ranked_genes_T <- as.vector(res_df_sort$sign.log.10.p)
names(ranked_genes_T) <- rownames(res_df_sort)

# Volcano Plot
TPH.volcano.data <- res_df_sort
TPH.volcano.data$log10.pvalue <- -log10(TPH.volcano.data$pvalue)
TPH.volcano.data$diffexpressed <- ifelse(TPH.volcano.data$log10.pvalue > -log10(0.05) & 
                                        abs(TPH.volcano.data$log2FoldChange) > 1, "Up or Down", "NS")
TPH.volcano.data$gene <- rownames(TPH.volcano.data)

# Volcano plot
TPH_volcano_plot <- ggplot(TPH.volcano.data, aes(x = log2FoldChange, y = log10.pvalue, 
                       color = diffexpressed, text = gene)) +
  geom_point(size = 1.5) +
  scale_color_manual(values = c("lightgrey", "darkblue")) + 
  labs(title = "Volcano Plot", x = "Log2(Fold Change)", y = "-log10(P-value)") + 
  theme_minimal()

ggplotly(TPH_volcano_plot, tooltip = "text")
```


```{r Run GSEA for TPH like cells, fig.width=10, fig.height=8}
#run GSEA
Tcell.GOBP <- GSEA(ranked_genes_T, TERM2GENE = GO_BP, TERM2NAME = GO_BP_t2n, seed=TRUE)

Tcell.GOBP.df <- data.frame(ID = Tcell.GOBP@result$ID, 
                             NES = Tcell.GOBP@result$NES, 
                             padj = Tcell.GOBP@result$p.adjust)

Tcell.GOBP.df.top10.up <- Tcell.GOBP.df %>% filter(NES > 0) %>% slice_min(padj, n = 10) %>%
  slice_max(NES, n = 10)
Tcell.GOBP.df.top10.dn <- Tcell.GOBP.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
Tcell.GOBP.df.top20 <- rbind(Tcell.GOBP.df.top10.up, Tcell.GOBP.df.top10.dn)

Tcell.GOBP.df.top20 <- Tcell.GOBP.df.top20[order(Tcell.GOBP.df.top20$NES),]
Tcell.GOBP.df.top20$ID <- factor(Tcell.GOBP.df.top20$ID, levels = Tcell.GOBP.df.top20$ID)

ggplot(Tcell.GOBP.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "GOBP Top 20 enriched Gene Sets in T cells")

#run GSEA
Tcell.REACTOME <- GSEA(ranked_genes_T, TERM2GENE = REACTOME, TERM2NAME = REACTOME_t2n, seed=TRUE)

Tcell.REACTOME.df <- data.frame(ID = Tcell.REACTOME@result$ID, 
                             NES = Tcell.REACTOME@result$NES, 
                             padj = Tcell.REACTOME@result$p.adjust)

Tcell.REACTOME.df.top10.up <- Tcell.REACTOME.df %>% filter(NES > 0) %>% slice_min(padj, n = 10) %>%
  slice_max(NES, n = 10)
Tcell.REACTOME.df.top10.dn <- Tcell.REACTOME.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
Tcell.REACTOME.df.top20 <- rbind(Tcell.REACTOME.df.top10.up, Tcell.REACTOME.df.top10.dn)

Tcell.REACTOME.df.top20 <- Tcell.REACTOME.df.top20[order(Tcell.REACTOME.df.top20$NES),]
Tcell.REACTOME.df.top20$ID <- factor(Tcell.REACTOME.df.top20$ID, levels = Tcell.REACTOME.df.top20$ID)

ggplot(Tcell.REACTOME.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "REACTOME Top 20 enriched Gene Sets in T cells")
```
