---
title: "Myeloid_Cells_Subclustering"
author: "Julia Hitschfel"
date: "2026-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(Seurat)
library(SeuratObject)
library(dplyr)
library(ggplot2)
library(colorspace)
library(RColorBrewer)
library(SingleR)
library(celldex)

farbenblind15 <- c("#000000","#004949","#009292","#FF6DB6","#FFB6DB",
                  "#490092","#006DDB","#B66DFF","#6DB6FF","#B6DBFF",
                  "#920000","#924900","#DB6D00","#24FF24","#FFFF6D")
```

## Myeloid cells subclustering

```{r Import data}
scobject <- readRDS(file = "integrated_seurat_refined_annotations.rds")

DimPlot(scobject, reduction = "umap", label = T)
DimPlot(scobject, reduction = "umap", group.by = "SCT_snn_res.0.4", label = T)

# Monocytes are cluster 5, myeloid cells is cluster 12

# Top marker in Cluster 5: TYROBP, AIF1, IFI30, C1QA, FCER1G, HLA class II, CD68, MS4A7, LYZ, MS4A6A, IL1B

# Top marker in Cluster 12: CORO1A, HLA-DQA1, TMSB4X, PFN1, RPL and RPS transkripte, LCP1, CD52

All_Myeloid_comb <- WhichCells(scobject, idents = c("Myeloid Cells_1", "Myeloid Cells_2"))

FeaturePlot(scobject, features = "CD14", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "FCER1G", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "LYZ", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "IFI30", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "TYROBP", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "HLA-DPB1", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "CD68", slot = "counts", max.cutoff = 5)

FeaturePlot(scobject, features = "CORO1A", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "HLA-DQA1", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "TMSB4X", slot = "counts", max.cutoff = 35)
FeaturePlot(scobject, features = "LCP1", slot = "counts", max.cutoff = 5)
FeaturePlot(scobject, features = "CD52", slot = "counts", max.cutoff = 5)
```

```{r Check filters}
genes_to_filter <- c('FCER1G', 'LYZ', 'IFI30', 'HLA-DPB1')
Non_Myeloid_Cells <- WhichCells(scobject, expression = FCER1G < 3 & LYZ < 3 & IFI30 < 2 &
                                     `HLA-DPB1` < 3 & TYROBP < 2, slot = "counts")

Non_Myeloid_Seurat <- subset(scobject, cells = Non_Myeloid_Cells)

DimPlot(Non_Myeloid_Seurat, reduction = "umap")

#Classical monocyte markers
FeaturePlot(Non_Myeloid_Seurat, features = "CD14")
FeaturePlot(Non_Myeloid_Seurat, features = "CD68")
FeaturePlot(Non_Myeloid_Seurat, features = "FCGR1A")

#Classical macrophage markers
FeaturePlot(Non_Myeloid_Seurat, features = "ITGAM")
FeaturePlot(Non_Myeloid_Seurat, features = "FCGR3A")
FeaturePlot(Non_Myeloid_Seurat, features = "CCR5")
```

```{r Read in Data from myeloid cells}
All.data <- list.dirs("D:/Sequencing_data/data/", recursive = F)

list.10x <- list()
counter <- 1
for(i in All.data){
  rnaseq.data <- Read10X_h5(paste0(i, "/count/sample_filtered_feature_bc_matrix.h5"))
  all.cells <- CreateSeuratObject(counts = rnaseq.data, project = "Colitis", min.cells = 3, 
                             min.features = 200)
  all.cells <- RenameCells(all.cells, new.names = paste0(colnames(all.cells), "_", counter))
  monos <- subset(all.cells, cells = All_Myeloid_comb)
  list.10x[[i]] <- monos
  counter <- counter + 1
}


print("Myeloid Cell Numbers before Filtering:")

for(i in list.10x){print(nrow(i@meta.data))}
```


```{r Filter myeloid cells with marker genes}
for(i in 1:length(list.10x)){  
  monos <- list.10x[[i]]
  monos@meta.data$date <- substr(names(list.10x)[i], 25, 100)
  Myeloid_Cells <- WhichCells(monos, expression = FCER1G > 2 | LYZ > 2 | IFI30 > 1 | `HLA-DPB1` > 2 | 
                                TYROBP > 1, slot = "counts")
  Bcells <- vector()
  if("MS4A1" %in% rownames(monos@assays$RNA$counts)){
    MS4A1_Bcells <- WhichCells(monos, expression = MS4A1 > 0, slot = "counts")
  }
    if(exists("Bcells")){Bcells <- c(Bcells, MS4A1_Bcells)}
  
  if("IGHM" %in% rownames(monos@assays$RNA$counts)){
    IGHM_Bcells <- WhichCells(monos, expression = IGHM > 0, slot = "counts")
  }
   if(exists("Bcells")){Bcells <- c(Bcells, IGHM_Bcells)}
  
  if("MZB1" %in% rownames(monos@assays$RNA$counts)){
    MZB1_Bcells <- WhichCells(monos, expression = MZB1 > 0, slot = "counts")
  }
   if(exists("Bcells")){Bcells <- c(Bcells,  MZB1_Bcells)}
  
  if("CD19" %in% rownames(monos@assays$RNA$counts)){
    CD19_Bcells <- WhichCells(monos, expression = CD19 > 0, slot = "counts")
  }
   if(exists("Bcells")){Bcells <- c(Bcells,  CD19_Bcells)}
  
  if("CD79A" %in% rownames(monos@assays$RNA$counts)){
    CD79A_Bcells <- WhichCells(monos, expression = CD79A > 0, slot = "counts")
  }
  if(exists("Bcells")){Bcells <- c(Bcells,  CD79A_Bcells)}
  
  Tcells <- WhichCells(monos, expression = CD3D > 0 | CD3E > 0, slot = "counts")
  
  Myeloid_Cells <- setdiff(Myeloid_Cells, Bcells)
  Myeloid_Cells <- setdiff(Myeloid_Cells, Tcells)
  monos <- subset(monos, cells = Myeloid_Cells)
  monos@meta.data$percent.mt <- PercentageFeatureSet(monos, pattern = "^MT")
  monos@meta.data$percent.rps <- PercentageFeatureSet(monos, pattern = "^RPS")
  monos@meta.data$percent.rpl <- PercentageFeatureSet(monos, pattern = "^RPL")
  monos <- SCTransform(monos, verbose = FALSE)
  list.10x[[i]] <- monos
}

print("Myeloid Cell Numbers after Filtering:")

for(i in list.10x){print(nrow(i@meta.data))}
```

The CCA Integration workflow does not work anymore, since there are fewer than 100 myeloid cells in the first sample. Instead, the RPCA Integration workflow will be used, since it allows adjusting k.weights. 

The worklow for RPCA will be used as describe in: https://satijalab.org/seurat/articles/integration_rpca.html

```{r Integrate Data}
features <- SelectIntegrationFeatures(object.list = list.10x, nfeatures = 3000)
list.10x <- PrepSCTIntegration(object.list = list.10x, anchor.features = features)
list.10x <- lapply(X = list.10x, FUN = RunPCA, features = features)

immune.anchors <- FindIntegrationAnchors(object.list = list.10x, normalization.method = "SCT",
    anchor.features = features, dims = 1:20, reduction = "rpca", k.anchor = 20)
scMyeloid <- IntegrateData(anchorset = immune.anchors, normalization.method = "SCT", 
                                     dims = 1:20, k.weight = 62)

scMyeloid <- RunPCA(scMyeloid, verbose = FALSE)
scMyeloid <- RunUMAP(scMyeloid, reduction = "pca", dims = 1:20)

# Visualization
p1 <- DimPlot(scMyeloid, reduction = "umap", group.by = "date")
p1 


scMyeloid <- FindNeighbors(scMyeloid, reduction = "pca", dims = 1:20)
scMyeloid <- FindClusters(scMyeloid, resolution = c(0.2, 0.4, 0.6))

DimPlot(scMyeloid, reduction = "umap", label = T, group.by = "integrated_snn_res.0.4",
        cols = farbenblind15, label.size = 6)

DimPlot(scMyeloid, reduction = "umap", group.by = "integrated_snn_res.0.2", label = T,
        cols = farbenblind15, label.size = 6)

DimPlot(scMyeloid, reduction = "umap", group.by = "date", split.by = "date", ncol = 3)
```


```{r QC plots}
Idents(scMyeloid) <- "integrated_snn_res.0.4"
sum_table <- scMyeloid@meta.data %>% group_by(date) %>% 
  summarise(RNA.count.median = median(nCount_RNA), RNA.feature.median = median(nFeature_RNA), 
            MT.freq.median = median(percent.mt), RPS.freq.median = median(percent.rps), 
            RPL.freq.median = median(percent.rpl))

sum_table

DimPlot(scMyeloid, reduction = "pca") + NoLegend()

FeaturePlot(scMyeloid, reduction = "pca", features = "percent.mt")
FeaturePlot(scMyeloid, reduction = "pca", features = "percent.rps")
FeaturePlot(scMyeloid, reduction = "pca", features = "percent.rpl")
FeaturePlot(scMyeloid, features = "nCount_RNA", reduction = "pca", max.cutoff = 5000) 
DimHeatmap(scMyeloid, dims = 1:15, cells = 500, balanced = TRUE)
```

```{r Examine clusters}
# Cluster Resolution 0.4
table(scMyeloid@meta.data$date, scMyeloid@meta.data$integrated_snn_res.0.4)

Cluster_over_time_04 <- scMyeloid@meta.data %>% group_by(date, integrated_snn_res.0.4) %>% 
  summarise(n = n()) %>% mutate(Freq = n/sum(n))

ggplot(Cluster_over_time_04, aes(x = date, y = Freq, fill = integrated_snn_res.0.4)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15)

# Cluster Resolution 0.2
table(scMyeloid@meta.data$date, scMyeloid@meta.data$integrated_snn_res.0.2)

Cluster_over_time_04 <- scMyeloid@meta.data %>% group_by(date, integrated_snn_res.0.2) %>% 
  summarise(n = n()) %>% mutate(Freq = n/sum(n))

ggplot(Cluster_over_time_04, aes(x = date, y = Freq, fill = integrated_snn_res.0.2)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15)
```

```{r Find marker, fig.height=14, fig.width=8}
scMyeloid <- JoinLayers(scMyeloid, assay = "RNA", layers = "counts")

# For better visualization, I will normalize the counts of the RNA slot to have a data slot 
scMyeloid <- NormalizeData(scMyeloid, assay = "RNA")
scMyeloid <- ScaleData(scMyeloid, assay = "RNA")

DefaultAssay(scMyeloid) <- "RNA"

# Cluster Resolution 0.4
Cluster_marker_04 <- FindAllMarkers(scMyeloid, logfc.threshold = 0.25, min.pct = 0.25, 
                                    only.pos = T)

write.csv(Cluster_marker_04, file = "Myeloid_cells_marker_genes_res04.csv")

top10_genes <- Cluster_marker_04 %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 10) %>% ungroup()

DoHeatmap(scMyeloid, features = unique(top10_genes$gene), assay = "RNA", slot = "scale.data") +
  NoLegend()

DotPlot(scMyeloid, features = unique(top10_genes$gene), group.by = "integrated_snn_res.0.4") +
  coord_flip()

Cluster_marker_04_noRPS <- Cluster_marker_04[which(grepl("^RPS*", Cluster_marker_04$gene) == F),]
Cluster_marker_04_noRPL <- Cluster_marker_04_noRPS[which(grepl("^RPL*", Cluster_marker_04_noRPS$gene) == F),]
top10_genes_noRPS <- Cluster_marker_04_noRPL %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 10) %>% ungroup()

DoHeatmap(scMyeloid, features = unique(top10_genes_noRPS$gene), assay = "RNA", slot = "scale.data") +
  NoLegend() + labs(title = "Heatmap without RPS transcripts")

DotPlot(scMyeloid, features = unique(top10_genes_noRPS$gene), group.by = "integrated_snn_res.0.4") +
  coord_flip() + labs(title = "Dotplot without RPS transcripts")

# Top 50 Marker without RPS/RPL for Sam
top50_genes_04 <- Cluster_marker_04_noRPL %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 50) %>% ungroup()

write.csv(top50_genes_04, file = "Myeloid_cells_top50_marker_genes_res04.csv")

# Cluster Resolution 0.2
Idents(scMyeloid) <- "integrated_snn_res.0.2"
Cluster_marker_02 <- FindAllMarkers(scMyeloid, logfc.threshold = 0.25, min.pct = 0.25, 
                                    only.pos = T)

write.csv(Cluster_marker_02, file = "Myeloid_cells_marker_genes_res02.csv")

top10_genes <- Cluster_marker_02 %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 10) %>% ungroup()

DoHeatmap(scMyeloid, features = unique(top10_genes$gene), assay = "RNA", slot = "scale.data") +
  NoLegend()

DotPlot(scMyeloid, features = unique(top10_genes$gene), group.by = "integrated_snn_res.0.2") +
  coord_flip()

Cluster_marker_02_noRPS <- Cluster_marker_02[which(grepl("^RPS*", Cluster_marker_02$gene) == F),]
Cluster_marker_02_noRPL <- Cluster_marker_02_noRPS[which(grepl("^RPL*", Cluster_marker_02_noRPS$gene) == F),]
top10_genes_noRPS <- Cluster_marker_02_noRPL %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 10) %>% ungroup()

DoHeatmap(scMyeloid, features = unique(top10_genes_noRPS$gene), assay = "RNA", slot = "scale.data") +
  NoLegend() + labs(title = "Heatmap without RPS transcripts")

DotPlot(scMyeloid, features = unique(top10_genes_noRPS$gene), group.by = "integrated_snn_res.0.4") +
  coord_flip() + labs(title = "Dotplot without RPS transcripts")

#Top 50 Marker for Sam
top50_genes_02 <- Cluster_marker_02_noRPL %>% group_by(cluster) %>% dplyr::filter(avg_log2FC > 1) %>% 
  slice_head(n = 50) %>% ungroup()

write.csv(top50_genes_02, file = "Myeloid_cells_top50_marker_genes_res04.csv")
```

```{r Annotations}
Myo.sce <- as.SingleCellExperiment(scMyeloid, assay = "integrated")
Myo.sce
ls('package:celldex')
ref.set <- celldex::HumanPrimaryCellAtlasData()
head(unique(ref.set$label.fine),20)
pred.cnts <- SingleR::SingleR(test = Myo.sce, ref = ref.set, labels = ref.set$label.fine)

lbls.keep <- table(pred.cnts$labels)>10
scMyeloid$SingleR.labels <- ifelse(lbls.keep[pred.cnts$labels], pred.cnts$labels, 'Other')

DimPlot(scMyeloid, reduction='umap', group.by='SingleR.labels', label = T) + NoLegend()

# Potential Idents for Res 0.4
Myos_annotations <- scMyeloid@meta.data %>% group_by(integrated_snn_res.0.4, SingleR.labels) %>% 
  summarise(potential.ident = n())

Myos_top_ann_04 <- Myos_annotations %>% group_by(integrated_snn_res.0.4) %>% 
  slice_max(potential.ident, n = 3)

print(Myos_top_ann_04)

# Potential Idents for Res 0.2
Myos_annotations_02 <- scMyeloid@meta.data %>% group_by(integrated_snn_res.0.2, SingleR.labels) %>% 
  summarise(potential.ident = n())

Myos_top_ann_02 <- Myos_annotations_02 %>% group_by(integrated_snn_res.0.2) %>% 
  slice_max(potential.ident, n = 3)

print(Myos_top_ann_02)


saveRDS(scMyeloid, file = "Myeloid_Cells.rds")
```

## Marker Gene Resolution 0.4

### Monocyte Marker Genes Res 0.4

```{r Violin Plots Monocyte markers}
Idents(scMyeloid) <- "integrated_snn_res.0.4"

#Monocyte marker
VlnPlot(scMyeloid, features = "CD14", layer = "data")
VlnPlot(scMyeloid, features = "FCGR3A", layer = "data") #CD16a (NK cells, macrophages)
VlnPlot(scMyeloid, features = "S100A8", layer = "data")
```

### Activated innate immun cells 0.4

```{r Monocyte markers 2}
VlnPlot(scMyeloid, features = "NEAT1", layer = "data")
VlnPlot(scMyeloid, features = "CD83", layer = "data")
VlnPlot(scMyeloid, features = "TNFAIP3", layer = "data")
VlnPlot(scMyeloid, features = "NFKB1", layer = "data")
```

### Macrophagen Marker Gene

```{r Macrophage Marker genes}
VlnPlot(scMyeloid, features = "CCR1", layer = "data")
VlnPlot(scMyeloid, features = "LYZ", layer = "data")
VlnPlot(scMyeloid, features = "APOE", layer = "data")
VlnPlot(scMyeloid, features = "CD4", layer = "data")
VlnPlot(scMyeloid, features = "ITGAM", layer = "data") #CD11b
VlnPlot(scMyeloid, features = "CSF1R", layer = "data") #CD115
VlnPlot(scMyeloid, features = "CD163", layer = "data") # (M2 Macs)
VlnPlot(scMyeloid, features = "MRC1", layer = "data") #CD206(M2 Mac)
VlnPlot(scMyeloid, features = "CD209", layer = "data") #(M2 Mac)
VlnPlot(scMyeloid, features = "TGFB1", layer = "data") #(M2 Mac)
VlnPlot(scMyeloid, features = "MAF", layer = "data") #(M2 Mac)
```


### DC Marker Gene

```{r Violins for DC marker genes}
VlnPlot(scMyeloid, features = "ID2") # cDC1 and cDC2 marker

VlnPlot(scMyeloid, features = "IRF8") # cDC1 marker
VlnPlot(scMyeloid, features = "CLEC9A") # cDC1 marker
VlnPlot(scMyeloid, features = "LY75") #CD205 # cDC1 marker
VlnPlot(scMyeloid, features = "ITGAE") #CD103 # cDC1 marker
VlnPlot(scMyeloid, features = "CD8A") #cDC1 marker
VlnPlot(scMyeloid, features = "CADM1") #cDC1 marker
VlnPlot(scMyeloid, features = "ITGAX") #CD11C

VlnPlot(scMyeloid, features = "IRF4") # cDC2 marker
VlnPlot(scMyeloid, features = "CD1C") #cDC2 marker
VlnPlot(scMyeloid, features = "FCER1A") # cDC2 Marker
VlnPlot(scMyeloid, features = "CLEC10A") # cDC2 Marker
VlnPlot(scMyeloid, features = "FCGR2B") #CD32b # cDC2 Marker
```


### pDC Marker Gene

```{r pDC marker}
VlnPlot(scMyeloid, features = "IL3RA") #CD123 
VlnPlot(scMyeloid, features = "NRP1") #CD304
```


### Neutrophil marker

```{r Neutrophil violins}
VlnPlot(scMyeloid, features = "FCGR3B") #CD16b
VlnPlot(scMyeloid, features = "CXCR2")
VlnPlot(scMyeloid, features = "IL17RA")
```

### B cell Marker Gene

```{r B cell marker}
VlnPlot(scMyeloid, features = "CD52")
VlnPlot(scMyeloid, features = "CD37")
VlnPlot(scMyeloid, features = "IGHA1")
```


#Inflammatory Marker Gene

```{r inflammation marker}
VlnPlot(scMyeloid, features = "IL1B")
VlnPlot(scMyeloid, features = "IL6")
VlnPlot(scMyeloid, features = "CCL24")
VlnPlot(scMyeloid, features = "MMP12")
```


```{r Rename Idents}
scMyeloid <- RenameIdents(scMyeloid, '0' = 'M2 Macrophages_1',
                          '1' = 'Inflammatory monocyte-derived Macrophages', 
                          '2' = 'M1 Macrophages', 
                          '3' = 'M2 Macrophages_2', 
                          '4' = 'cDC1', 
                          '5' = 'cDC2', 
                          '6' = 'Neutrophils')

DimPlot(scMyeloid, reduction = "umap", label = T)

scMyeloid$Annotation <- Idents(scMyeloid)

Cluster_over_time_ann_04 <- scMyeloid@meta.data %>% group_by(date, Annotation) %>% 
  summarise(n = n()) %>% mutate(Freq = (n/sum(n))*100)

ggplot(Cluster_over_time_ann_04, aes(x = date, y = Freq, fill = Annotation)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15) + labs(y = "% of myeloid cells") + 
  theme_minimal()

DimPlot(scMyeloid, reduction = "umap", cols = farbenblind15) + coord_equal(ratio = 1.5)

ggsave("/Figures/Myeloid_cells_umap_res04.tiff", plot = last_plot(),
       device = "tiff", dpi = 300, width = 10, height = 7, units = "in", compression = "lzw")

saveRDS(scMyeloid, file = "Myeloid_Cells_annotated.rds")
```

## Marker Gene Resolution 0.2

### Monocyte Marker Resolution 0.2

```{r Monocyte Marker}
Idents(scMyeloid) <- "integrated_snn_res.0.2"

#Monocyte marker
VlnPlot(scMyeloid, features = "CD14", layer = "data")
VlnPlot(scMyeloid, features = "FCGR3A", layer = "data") #CD16a (NK cells, macrophages)
VlnPlot(scMyeloid, features = "CD163", layer = "data")
VlnPlot(scMyeloid, features = "S100A8", layer = "data")
```

```{r Activated innate immune cells}
VlnPlot(scMyeloid, features = "NEAT1", layer = "data")
VlnPlot(scMyeloid, features = "CD83", layer = "data")
VlnPlot(scMyeloid, features = "TNFAIP3", layer = "data")
VlnPlot(scMyeloid, features = "NFKB1", layer = "data")
```

### Macrophages Marker 0.2

```{r Macs 0.2}
VlnPlot(scMyeloid, features = "C1QC", layer = "data")
VlnPlot(scMyeloid, features = "CCR1", layer = "data")
VlnPlot(scMyeloid, features = "LYZ", layer = "data")
VlnPlot(scMyeloid, features = "APOE", layer = "data")
VlnPlot(scMyeloid, features = "CD4", layer = "data")
VlnPlot(scMyeloid, features = "ITGAM", layer = "data") #CD11b
VlnPlot(scMyeloid, features = "CSF1R", layer = "data") #CD115
VlnPlot(scMyeloid, features = "MRC1", layer = "data") #CD206(M2 Mac)
VlnPlot(scMyeloid, features = "CD209", layer = "data") #(M2 Mac)
VlnPlot(scMyeloid, features = "TGFB1", layer = "data") #(M2 Mac)
VlnPlot(scMyeloid, features = "MAF", layer = "data") #(M2 Mac)
```
```{r DC marker 0.2}
VlnPlot(scMyeloid, features = "ID2") # cDC1 and cDC2 marker

VlnPlot(scMyeloid, features = "IRF8") # cDC1 marker
VlnPlot(scMyeloid, features = "CLEC9A") # cDC1 marker
VlnPlot(scMyeloid, features = "LY75") #CD205 # cDC1 marker
VlnPlot(scMyeloid, features = "ITGAE") #CD103 # cDC1 marker
VlnPlot(scMyeloid, features = "CD8A") #cDC1 marker
VlnPlot(scMyeloid, features = "CADM1") #cDC1 marker
VlnPlot(scMyeloid, features = "ITGAX") #CD11C

VlnPlot(scMyeloid, features = "IRF4") # cDC2 marker
VlnPlot(scMyeloid, features = "CD1C") #cDC2 marker
VlnPlot(scMyeloid, features = "FCER1A") # cDC2 Marker
VlnPlot(scMyeloid, features = "CLEC10A") # cDC2 Marker
VlnPlot(scMyeloid, features = "FCGR2B") #CD32b # cDC2 Marker
```

```{r pDC marker 02}
VlnPlot(scMyeloid, features = "IL3RA") #CD123 
VlnPlot(scMyeloid, features = "NRP1") #CD304
```


```{r Neutrophils 0.2}
VlnPlot(scMyeloid, features = "FCGR3B") #CD16b
VlnPlot(scMyeloid, features = "CXCR2")
VlnPlot(scMyeloid, features = "IL17RA")
VlnPlot(scMyeloid, features = "CEACAM8")
```

```{r B and T cell Marker 0.2}
VlnPlot(scMyeloid, features = "CD52")
VlnPlot(scMyeloid, features = "CD37")
VlnPlot(scMyeloid, features = "IGHA1")
```


```{r Inflammatory marker 0.2}
VlnPlot(scMyeloid, features = "IL1B")
VlnPlot(scMyeloid, features = "IL6")
VlnPlot(scMyeloid, features = "CCL24")
VlnPlot(scMyeloid, features = "MMP12")
```


```{r Idents Res 0.2}
Idents(scMyeloid) <- "integrated_snn_res.0.2" 

scMyeloid_02 <- RenameIdents(scMyeloid, '0' = 'M1 Macrophages',
                          '1' = 'Inflammatory monocyte-derived macrophage', 
                          '2' = 'M2 Macrophages', 
                          '3' = 'DCs', 
                          '4' = 'Neutrophils')

DimPlot(scMyeloid_02, reduction = "umap", label = T)

scMyeloid_02$Annotation <- Idents(scMyeloid_02)

Cluster_over_time_ann_02 <- scMyeloid_02@meta.data %>% group_by(date, Annotation) %>% 
  summarise(n = n()) %>% mutate(Freq = (n/sum(n))*100)

ggplot(Cluster_over_time_ann_02, aes(x = date, y = Freq, fill = Annotation)) + 
  geom_col(color = "black") + scale_fill_manual(values = farbenblind15) + labs(y = "% of myeloid cells") + 
  theme_minimal()

DimPlot(scMyeloid_02, reduction = "umap", cols = farbenblind15) + coord_equal(ratio = 1.5)

ggsave("/Figures/Myeloid_cells_umap_res02.tiff", plot = last_plot(),
       device = "tiff", dpi = 300, width = 10, height = 7, units = "in", compression = "lzw")

saveRDS(scMyeloid_02, "Myeloid_Cells_res02_annotated.rds")
```

```{r Frequencies in all cells}
All_cells_TP <- scobject@meta.data %>% group_by(date) %>% summarise(All.n = n())
All_cells_TP

## All Myeloid cells
Myeloid_over_time <- scMyeloid@meta.data %>% group_by(date) %>% 
  summarise(Myeloid.n = n()) 

# Merge data.frames
Freq_of_myeloid_in_all <- full_join(All_cells_TP, Myeloid_over_time, by = "date")

Freq_of_myeloid_in_all <- Freq_of_myeloid_in_all %>% mutate(Freq_in_all = Myeloid.n/All.n*100)

ggplot(Freq_of_myeloid_in_all, aes(x = date, y = Freq_in_all)) + 
  geom_col(fill = "#924900") + labs(x = "Date" , y = "% of total cells") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave(filename = "/Figures/Percentage_all_myeloid_cells_in_total_cells.tiff",
  plot = last_plot(), device = "tiff", dpi = 300, width = 10, height = 7, units = "in", 
       compression = "lzw")

# Frequencies for myeloid clusters res 0.4 
# Add Zeros for missing clusters

Cluster_over_time_res_04 <- scMyeloid@meta.data %>% group_by(date, integrated_snn_res.0.4) %>% 
  summarise(Myeloid_n = n()) %>% mutate(Freq = (Myeloid_n/sum(Myeloid_n))*100)

Idents(scMyeloid) <- "Annotation"
Myeloid_Clusters <- seq(0,6,1)
date.list <- list()
for(i in unique(Cluster_over_time_res_04$date)){
  Myeloid_date <- Cluster_over_time_res_04 %>% subset(Cluster_over_time_res_04$date == i)
  if(nrow(Myeloid_date) < 7){
    missing_Myeloid_Clusters <- setdiff(Myeloid_Clusters, Myeloid_date$integrated_snn_res.0.4)
    Zero_Clusters <- data.frame(date = rep(i, length(missing_Myeloid_Clusters)),
                                integrated_snn_res.0.4 = as.factor(missing_Myeloid_Clusters),
                                Myeloid_n = rep(0, length(missing_Myeloid_Clusters)),
                                Freq = rep(0, length(missing_Myeloid_Clusters)))
    Myeloid_date <- rbind(Myeloid_date, Zero_Clusters)
  }
  date.list[[i]] <- Myeloid_date
}

Myeloid_Cluster_over_time <- do.call(rbind, date.list)

# Name Myeloid cell clusters
Myeloid_cluster_annotations <- data.frame("integrated_snn_res.0.4" = seq(0,6,1),
                                        "Annotation" = levels(scMyeloid$Annotation))

Myeloid_Cluster_over_time <- merge(Myeloid_Cluster_over_time, Myeloid_cluster_annotations,
                                      by = "integrated_snn_res.0.4")

write.csv(Myeloid_Cluster_over_time, file = "Myeloid_cell_freq_res04.csv")

# Lineplot
ggplot(Myeloid_Cluster_over_time, aes(x = date, y = Freq, group = Annotation,
                                 color = Annotation)) + geom_line(linewidth = 1.5) +
  scale_color_manual(values = farbenblind15) + labs(x = "Date" , y = "% of Myeloid Cells") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave(filename = "/Figures/Cluster_percentage_myeloid_cells_res04.tiff", 
       plot = last_plot(), device = "tiff", dpi = 300, width = 10, height = 7, units = "in", 
       compression = "lzw")

## Percentage of myeloid cell clusters in all cells (resolution 0.4)
Freq_of_myeloid_cluster_in_all <- full_join(Myeloid_Cluster_over_time, All_cells_TP, by = "date")

Freq_of_myeloid_cluster_in_all <- Freq_of_myeloid_cluster_in_all %>% mutate(Freq_in_all = Myeloid_n/All.n*100)

write.csv(Freq_of_myeloid_cluster_in_all, file = "Myeloid_cell_freq_in_total_cells_res04.csv")


ggplot(Freq_of_myeloid_cluster_in_all, aes(x = date, y = Freq_in_all, group = Annotation, color = Annotation)) + 
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = farbenblind15) + labs(x = "Date" , y = "% of total cells") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave(filename = "/Figures/Cluster_percentage_myeloid_cells_in_total_cells_res04.tiff",
  plot = last_plot(), device = "tiff", dpi = 300, width = 10, height = 7, units = "in", compression = "lzw")



## Resolution 0.2
# Add Zeros for missing clusters

Cluster_over_time_res_02 <- scMyeloid@meta.data %>% group_by(date, integrated_snn_res.0.2) %>% 
  summarise(Myeloid_n = n()) %>% mutate(Freq = (Myeloid_n/sum(Myeloid_n))*100)

Myeloid_Clusters_02 <- seq(0,4,1)
date.list.02 <- list()
for(i in unique(Cluster_over_time_res_02$date)){
  Myeloid_date_02 <- Cluster_over_time_res_02 %>% subset(Cluster_over_time_res_02$date == i)
  if(nrow(Myeloid_date_02) < 5){
    missing_Myeloid_Clusters <- setdiff(Myeloid_Clusters_02, Myeloid_date_02$integrated_snn_res.0.2)
    Zero_Clusters_02 <- data.frame(date = rep(i, length(missing_Myeloid_Clusters)),
                                integrated_snn_res.0.2 = as.factor(missing_Myeloid_Clusters),
                                Myeloid_n = rep(0, length(missing_Myeloid_Clusters)),
                                Freq = rep(0, length(missing_Myeloid_Clusters)))
    Myeloid_date_02 <- rbind(Myeloid_date_02, Zero_Clusters_02)
  }
  date.list.02[[i]] <- Myeloid_date_02
}

Myeloid_Cluster_over_time_02 <- do.call(rbind, date.list.02)

# Name Myeloid cell clusters
Myeloid_cluster_annotations_02 <- data.frame("integrated_snn_res.0.2" = seq(0,4,1),
                                        "Annotation" = levels(Idents(scMyeloid_02)))

Myeloid_Cluster_over_time_02 <- merge(Myeloid_Cluster_over_time_02, Myeloid_cluster_annotations_02,
                                      by = "integrated_snn_res.0.2")

write.csv(Myeloid_Cluster_over_time_02, file = "Myeloid_cell_freq_res02.csv")

# Lineplot
ggplot(Myeloid_Cluster_over_time_02, aes(x = date, y = Freq, group = Annotation,
                                 color = Annotation)) + geom_line(linewidth = 1.5) +
  scale_color_manual(values = farbenblind15) + labs(x = "Date" , y = "% of Myeloid Cells") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave(filename = "/Figures/Cluster_percentage_myeloid_cells_res02.tiff", 
       plot = last_plot(), device = "tiff", dpi = 300, width = 10, height = 7, units = "in", 
       compression = "lzw")

## Percentage of myeloid cell clusters in all cells (resolution 0.2)
Freq_of_myeloid_cluster_in_all_02 <- full_join(Myeloid_Cluster_over_time_02, All_cells_TP, by = "date")

Freq_of_myeloid_cluster_in_all_02 <- Freq_of_myeloid_cluster_in_all_02 %>% mutate(Freq_in_all = Myeloid_n/All.n*100)

write.csv(Freq_of_myeloid_cluster_in_all_02, file = "Myeloid_cell_freq_in_total_cells_res02.csv")


ggplot(Freq_of_myeloid_cluster_in_all_02, aes(x = date, y = Freq_in_all, group = Annotation, color = Annotation)) + 
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = farbenblind15) + labs(x = "Date" , y = "% of total cells") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave(filename = "/Figures/Cluster_percentage_myeloid_cells_in_total_cells_res02.tiff",
  plot = last_plot(), device = "tiff", dpi = 300, width = 10, height = 7, units = "in", 
       compression = "lzw")
```
