---
title: "DESeq2_Myeloid_Cells"
author: "Julia Hitschfel"
date: "2026-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Seurat)
library(dplyr)
library(DESeq2)
library(ggplot2) 
library(clusterProfiler)
library(tidyverse)
library(msigdbr)
library(ggrepel)
library(ggforce)
library(tidytext)
```


## DESeq2 Pseudobulk Timepoints


```{r DESeq2 for all myeloid cells}
myeloid.seurat <- readRDS("Myeloid_Cells_annotated.rds")

all.myeloid.pseudo <- AggregateExpression(myeloid.seurat, assays = "RNA", 
                                        group.by = "date", slot = "counts",
                                        return.seurat = F)

# Create meta.data for count matrix
Myeloid.numbers <- myeloid.seurat@meta.data %>% group_by(date) %>% summarise(Myeloid.count = n())
Myeloid.numbers$date <- paste0("g", Myeloid.numbers$date)
Myeloid.numbers$treatment <- ifelse(Myeloid.numbers$date == "g2024_07_15" | 
                                    Myeloid.numbers$date == "g2024_11_28" |
                               Myeloid.numbers$date == "g2025_02_17", "pre_CART", "post_CART")

# Visualize cell counts
ggplot(Myeloid.numbers, aes(x=date, y=Myeloid.count, fill=treatment)) +
    geom_bar(stat="identity", color="black") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(x="Timepoint", y="Number of Myeloid cells") +
    geom_text(aes(label=Myeloid.count), vjust=-0.5)

# Perform DESeq2 for Myeloid pseudobulk
# To prevent the B cell pathways from popping up, I am removing all the IG transcripts at this step:
count.data.all.myeloid <- as.matrix(all.myeloid.pseudo$RNA)

IG.transcripts <- c(which(grepl("^IGH", rownames(count.data.all.myeloid)) == T), 
                    which(grepl("^IGL", rownames(count.data.all.myeloid)) == T), 
                    which(grepl("^IGK", rownames(count.data.all.myeloid)) == T))

count.data.all.myeloid.noIG <- count.data.all.myeloid[-IG.transcripts,]
class(count.data.all.myeloid.noIG) <- "integer"

# myeloid numbers now has underscores, while the matrix has dashes.... change to dashes in Myeloid numbers:
Myeloid.numbers$date <- gsub("_", "-", Myeloid.numbers$date)

dds.all.myeloid <- DESeqDataSetFromMatrix(countData = count.data.all.myeloid.noIG, 
				colData =  Myeloid.numbers,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds.all.myeloid)) >= 10
dds.all.myeloid <- dds.all.myeloid[keep,]

dds.all.myeloid <- DESeq(dds.all.myeloid)

resultsNames(dds.all.myeloid)

#Generate results object
res.all.myeloid <- results(dds.all.myeloid, name = "treatment_pre_CART_vs_post_CART")
res.all.myeloid 

res.all.myeloid_df <- as.data.frame(res.all.myeloid)

# Sort dataframe for adjusted p value and show top20 results 
res.all.myeloid_df_sort <- res.all.myeloid_df[order(res.all.myeloid_df$padj),]
head(res.all.myeloid_df_sort, 20)

write.csv(res.all.myeloid_df_sort, "/DESeq2/DEGs_Myeloid_cells.csv")

# Violin Plot of DEGs
res.all.myeloid_df_volcano <- res.all.myeloid_df
res.all.myeloid_df_volcano$neg.log10.padj <- -log10(res.all.myeloid_df_volcano$padj)
res.all.myeloid_df_volcano <- res.all.myeloid_df_volcano %>% filter(!is.na(neg.log10.padj))


res.all.myeloid_df_volcano <- res.all.myeloid_df_volcano %>%
  mutate(gene_type = case_when(log2FoldChange >= 1 & padj <= 0.05 ~ "up",
                               log2FoldChange <= -1 & padj <= 0.05 ~ "down",
                               TRUE ~ "ns"))  

cols <- c("up" = "#ffad73", "down" = "#26b3ff", "ns" = "grey") 
sizes <- c("up" = 2, "down" = 2, "ns" = 1) 
alphas <- c("up" = 1, "down" = 1, "ns" = 0.5)

significant.genes <- res.all.myeloid_df_volcano %>% filter(padj <= 0.05)
significant.genes$geneID <- rownames(significant.genes)
  
res.all.myeloid_df_volcano %>%
  ggplot(aes(x = log2FoldChange,
             y = neg.log10.padj,
             fill = gene_type,    
             size = gene_type,
             alpha = gene_type)) + 
  geom_point(shape = 21, # Specify shape and colour as fixed local parameters    
             colour = "black") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),
             linetype = "dashed") +
   geom_label_repel(data = significant.genes, # Add labels last to appear as the top layer  
                   aes(label = geneID),
                   force = 2,
                   nudge_y = 0.25) +
  scale_fill_manual(values = cols) + # Modify point colour
  scale_size_manual(values = sizes) + # Modify point size
  scale_alpha_manual(values = alphas) + # Modify point transparency
  scale_x_continuous(breaks = c(seq(-8, 8, 2)),       
                     limits = c(-8, 8)) + 
  labs(x = "log2(Fold Change)", y = "-log10(adj. P value)", 
       title = "Differentially expressed genes in myeloid cells pre and post CAR T") + theme_bw()
```

## Pre-ranked GSEA mit clusterProfiler

Ranking was done with signed -log10(p value)


### GSEA für alle myeloischen Zellen mit HALLMARK Gensets

```{r GSEA mir clusterProfiler}
res.all.myeloid_df_sort$sign.log.10.p <-
  sign(res.all.myeloid_df_sort$log2FoldChange)*(-log10(res.all.myeloid_df_sort$pvalue))
res.all.myeloid_df_sort <- res.all.myeloid_df_sort %>% filter(!is.na(sign.log.10.p))
res.all.myeloid_df_sort <- res.all.myeloid_df_sort[order(res.all.myeloid_df_sort$sign.log.10.p, 
                                                         decreasing = T),]

ranked_genes_M <- as.vector(res.all.myeloid_df_sort$sign.log.10.p)
names(ranked_genes_M) <- rownames(res.all.myeloid_df_sort)

# Import the MSigDB Hallmark Collection
hallmark<- msigdbr(collection = "H") %>% 
  dplyr::select(gs_name, gene_symbol)

head(hallmark)

h2 <- msigdbr(collection = "H") %>% 
  dplyr::select(gs_name, gs_description)

#set seed
set.seed(123)
#run GSEA
All.Myeloid.HALLMARK <- GSEA(ranked_genes_M, TERM2GENE = hallmark, TERM2NAME = h2, seed=TRUE)

head(All.Myeloid.HALLMARK)

print(paste0(All.Myeloid.HALLMARK@result$ID, " - NES:  ", All.Myeloid.HALLMARK@result$NES,  " - padj:  ",
             All.Myeloid.HALLMARK@result$p.adjust))

All.Myeloid.Hallmark.df <- data.frame(ID = All.Myeloid.HALLMARK@result$ID, 
                             NES = All.Myeloid.HALLMARK@result$NES, 
                             padj = All.Myeloid.HALLMARK@result$p.adjust)

All.Myeloid.Hallmark.df <- All.Myeloid.Hallmark.df[order(All.Myeloid.Hallmark.df$NES),]
All.Myeloid.Hallmark.df$ID <- factor(All.Myeloid.Hallmark.df$ID, levels = All.Myeloid.Hallmark.df$ID)

ggplot(All.Myeloid.Hallmark.df, aes(x = ID, y = NES, fill = padj)) + geom_col() + 
  scale_fill_gradient() + coord_flip() + theme_minimal()

```


### GO and REACTOME GSEA for myeloid cells 

```{r GO and REACTOME all myeloid cells}
GO_BP <- msigdbr(collection = "C5", subcollection = "GO:BP") %>% 
  dplyr::select(gs_name, gene_symbol)

GO_BP_t2n <- msigdbr(collection = "C5", subcollection = "GO:BP") %>% 
  dplyr::select(gs_name, gs_description)

#run GSEA
All.Myeloid.GO.BP <- GSEA(ranked_genes_M, TERM2GENE = GO_BP, TERM2NAME = GO_BP_t2n, seed=TRUE)

All.Myeloid.GO.BP.df <- data.frame(ID = All.Myeloid.GO.BP@result$ID, 
                             NES = All.Myeloid.GO.BP@result$NES, 
                             padj = All.Myeloid.GO.BP@result$p.adjust)

All.Myeloid.GO.BP.df.top10.up <- All.Myeloid.GO.BP.df %>% filter(NES > 0) %>% slice_min(padj, n = 10)
All.Myeloid.GO.BP.df.top10.dn <- All.Myeloid.GO.BP.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
All.Myeloid.GO.BP.df.top20 <- rbind(All.Myeloid.GO.BP.df.top10.up, All.Myeloid.GO.BP.df.top10.dn)

All.Myeloid.GO.BP.df.top20$ID <- ifelse(All.Myeloid.GO.BP.df.top20$ID ==
"GOBP_ADAPTIVE_IMMUNE_RESPONSE_BASED_ON_SOMATIC_RECOMBINATION_OF_IMMUNE_RECEPTORS_BUILT_FROM_IMMUNOGLOBULIN_SUPERFAMILY_DOMAINS", "GOBP_SOMATIC_RECOMBINATION_OF_IMMUNE_RECEPTORS_FROM_IG_SUPERFAMILY_DOMAINS", All.Myeloid.GO.BP.df.top20$ID)

All.Myeloid.GO.BP.df.top20 <- All.Myeloid.GO.BP.df.top20[order(All.Myeloid.GO.BP.df.top20$NES),]
All.Myeloid.GO.BP.df.top20$ID <- factor(All.Myeloid.GO.BP.df.top20$ID, levels = All.Myeloid.GO.BP.df.top20$ID)

ggplot(All.Myeloid.GO.BP.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() + 
  scale_fill_gradient() + coord_flip() + labs(title = "GO:BP Top 20 enriched Gene Sets in Myeloid cells") + theme_minimal() + 
  theme(axis.text.x = element_text(size = 7), axis.text.y = element_text(size = 7))
  


## GSEA with REACTOME
REACTOME <- msigdbr(collection = "C2", subcollection = "CP:REACTOME") %>% 
  dplyr::select(gs_name, gene_symbol)

REACTOME_t2n <- msigdbr(collection = "C2", subcollection = "CP:REACTOME") %>% 
  dplyr::select(gs_name, gs_description)

#run GSEA
All.Myeloid.REACTOME <- GSEA(ranked_genes_M, TERM2GENE = REACTOME, TERM2NAME = REACTOME_t2n, seed=TRUE)

All.Myeloid.REACTOME.df <- data.frame(ID = All.Myeloid.REACTOME@result$ID, 
                             NES = All.Myeloid.REACTOME@result$NES, 
                             padj = All.Myeloid.REACTOME@result$p.adjust)

All.Myeloid.REACTOME.df.top10.up <- All.Myeloid.REACTOME.df %>% filter(NES > 0) %>% slice_min(padj, n = 10) %>%
  slice_max(NES, n = 10)
All.Myeloid.REACTOME.df.top10.dn <- All.Myeloid.REACTOME.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
All.Myeloid.REACTOME.df.top20 <- rbind(All.Myeloid.REACTOME.df.top10.up, All.Myeloid.REACTOME.df.top10.dn)

All.Myeloid.REACTOME.df.top20$ID <- ifelse(All.Myeloid.REACTOME.df.top20$ID ==
    "REACTOME_ANTIGEN_ACTIVATES_B_CELL_RECEPTOR_BCR_LEADING_TO_GENERATION_OF_SECOND_MESSENGERS",
    "REACTOME_SECOND_MESSENGER_GENERATION_AFTER_ANTIGENIC_BCR_ACTIVATION", All.Myeloid.REACTOME.df.top20$ID)

All.Myeloid.REACTOME.df.top20 <- All.Myeloid.REACTOME.df.top20[order(All.Myeloid.REACTOME.df.top20$NES),]
All.Myeloid.REACTOME.df.top20$ID <- factor(All.Myeloid.REACTOME.df.top20$ID, levels = All.Myeloid.REACTOME.df.top20$ID)

ggplot(All.Myeloid.REACTOME.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "REACTOME Top 20 enriched Gene Sets in Myeloid cells") + 
  theme(axis.text.x = element_text(size = 7), axis.text.y = element_text(size = 7))
```


### DESeq2 for myeloid cell subsets (Resolution 0.4)

```{r }
myeloid.pseudo <- AggregateExpression(myeloid.seurat, assays = "RNA", 
                                        group.by = c("date", "Annotation"), slot = "counts",
                                        return.seurat = F)

# Manipulate the matrix to fit DESeq2 input
matrix.t <- t(myeloid.pseudo$RNA)
count.data.frame <- as.data.frame(matrix.t)
splitRows <- gsub('.*_', '', rownames(count.data.frame))
cell.type.list <- split.data.frame(count.data.frame, f = factor(splitRows))

#Retranspose the matrix and fix column names for all elements of the list + remove IG transcripts
cell.type.list.t <- lapply(cell.type.list, function(x){
	rownames(x) <- gsub('(.*)_.*', '\\1', rownames(x))
	t(x)
})

cell.type.list.t.noIG <- lapply(cell.type.list.t, function(x){
    x <- x[-IG.transcripts,]
})

# Create meta.data for count matrix
col.data <- data.frame(samples = colnames(cell.type.list.t.noIG$`M1 Macrophages`))

col.data$treatment <- ifelse(col.data$samples == "g2024-07-15" | col.data$samples == "g2024-11-28" |
                               col.data$samples == "g2025-02-17", "pre_CART", "post_CART")

# Perform DESeq2 for M1_Macs
count.data.M1Macs <- cell.type.list.t.noIG$`M1 Macrophages`
class(count.data.M1Macs) <- "integer"


dds <- DESeqDataSetFromMatrix(countData = count.data.M1Macs, 
				colData =  col.data,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds)

resultsNames(dds)

#Generate results object
res <- results(dds, name = "treatment_pre_CART_vs_post_CART")
res 

res_df <- as.data.frame(res)

# Sort dataframe for adjusted p value and show top20 results 
res_df_sort <- res_df[order(res_df$padj),]
head(res_df_sort, 20)

## violin Plot for M1 Macrophages
# Violin Plot of DEGs
res_df_volcano <- res_df
res_df_volcano$neg.log10.padj <- -log10(res_df_volcano$padj)
res_df_volcano <- res_df_volcano %>% filter(!is.na(neg.log10.padj))


res_df_volcano <- res_df_volcano %>%
  mutate(gene_type = case_when(log2FoldChange >= 1 & padj <= 0.05 ~ "up",
                               log2FoldChange <= -1 & padj <= 0.05 ~ "down",
                               TRUE ~ "ns"))  

significant.genes <- res_df_volcano %>% filter(padj <= 0.05)
significant.genes$geneID <- rownames(significant.genes)
  
res_df_volcano %>%
  ggplot(aes(x = log2FoldChange,
             y = neg.log10.padj,
             fill = gene_type,    
             size = gene_type,
             alpha = gene_type)) + 
  geom_point(shape = 21, # Specify shape and colour as fixed local parameters    
             colour = "black") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),
             linetype = "dashed") +
   geom_label_repel(data = significant.genes, # Add labels last to appear as the top layer  
                   aes(label = geneID),
                   force = 2,
                   nudge_y = 0.25) +
  scale_fill_manual(values = cols) + # Modify point colour
  scale_size_manual(values = sizes) + # Modify point size
  scale_alpha_manual(values = alphas) + # Modify point transparency
  scale_x_continuous(breaks = c(seq(-8, 8, 2)),
                     limits = c(-8, 8)) +
  labs(x = "log2(Fold Change)", y = "-log10(adj. P value)", 
       title = "Differentially expressed genes in M1 macrophages pre and post CAR T") + theme_bw()
```

```{r DESeq2 for M2 Macs}
# Create meta.data for count matrix
col.data.M2 <- data.frame(samples = colnames(cell.type.list.t.noIG$`M2 Macrophages-1`))

col.data.M2$treatment <- ifelse(col.data.M2$samples == "g2024-07-15" | col.data.M2$samples == "g2024-11-28" |
                               col.data.M2$samples == "g2025-02-17", "pre_CART", "post_CART")

# Perform DESeq2 for M1_Macs
count.data.M2Macs1 <- cell.type.list.t.noIG$`M2 Macrophages-1`
class(count.data.M2Macs1) <- "integer"


dds <- DESeqDataSetFromMatrix(countData = count.data.M2Macs1, 
				colData =  col.data.M2,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds)

resultsNames(dds)

#Generate results object
res_M2Macs1 <- results(dds, name = "treatment_pre_CART_vs_post_CART")
res_M2Macs1 

res_M2Macs1_df <- as.data.frame(res_M2Macs1)

# Sort dataframe for adjusted p value and show top20 results 
res_M2Macs1_df_sort <- res_M2Macs1_df[order(res_M2Macs1_df$padj),]
head(res_M2Macs1_df_sort, 20)

## violin Plot for M2 Macrophages 1
# Violin Plot of DEGs
res_M2Macs1_df_volcano <- res_M2Macs1_df
res_M2Macs1_df_volcano$neg.log10.padj <- -log10(res_M2Macs1_df_volcano$padj)
res_M2Macs1_df_volcano <- res_M2Macs1_df_volcano %>% filter(!is.na(neg.log10.padj))


res_M2Macs1_df_volcano <- res_M2Macs1_df_volcano %>%
  mutate(gene_type = case_when(log2FoldChange >= 1 & padj <= 0.05 ~ "up",
                               log2FoldChange <= -1 & padj <= 0.05 ~ "down",
                               TRUE ~ "ns"))  

significant.genes.M2 <- res_M2Macs1_df_volcano %>% filter(padj <= 0.05)
significant.genes.M2$geneID <- rownames(significant.genes.M2)
  
res_M2Macs1_df_volcano %>%
  ggplot(aes(x = log2FoldChange,
             y = neg.log10.padj,
             fill = gene_type,    
             size = gene_type,
             alpha = gene_type)) + 
  geom_point(shape = 21, # Specify shape and colour as fixed local parameters    
             colour = "black") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),
             linetype = "dashed") +
   geom_label_repel(data = significant.genes.M2, # Add labels last to appear as the top layer  
                   aes(label = geneID),
                   force = 2,
                   nudge_y = 0.25) +
  scale_fill_manual(values = cols) + # Modify point colour
  scale_size_manual(values = sizes) + # Modify point size
  scale_alpha_manual(values = alphas) + # Modify point transparency
  scale_x_continuous(breaks = c(seq(-8, 8, 2)),
                     limits = c(-8, 8)) +
  labs(x = "log2(Fold Change)", y = "-log10(adj. P value)", 
       title = "Differentially expressed genes in M2 macrophages (1) pre and post CAR T") + theme_bw()
```

```{r DESeq2 for M2 Macs 2}
# Create meta.data for count matrix
col.data.M2.2 <- data.frame(samples = colnames(cell.type.list.t.noIG$`M2 Macrophages-2`))

col.data.M2.2$treatment <- ifelse(col.data.M2.2$samples == "g2024-07-15" | 
                                    col.data.M2.2$samples == "g2024-11-28" |
                               col.data.M2.2$samples == "g2025-02-17", "pre_CART", "post_CART")

# Perform DESeq2 for M1_Macs
count.data.M2Macs2 <- cell.type.list.t.noIG$`M2 Macrophages-2`
class(count.data.M2Macs2) <- "integer"


dds <- DESeqDataSetFromMatrix(countData = count.data.M2Macs2, 
				colData =  col.data.M2.2,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds)

resultsNames(dds)

#Generate results object
res_M2Macs2 <- results(dds, name = "treatment_pre_CART_vs_post_CART")
res_M2Macs2 

res_M2Macs2_df <- as.data.frame(res_M2Macs2)

# Sort dataframe for adjusted p value and show top20 results 
res_M2Macs2_df_sort <- res_M2Macs2_df[order(res_M2Macs2_df$padj),]
head(res_M2Macs2_df_sort, 20)

## violin Plot for M2 Macrophages 2
# Violin Plot of DEGs
res_M2Macs2_df_volcano <- res_M2Macs2_df
res_M2Macs2_df_volcano$neg.log10.padj <- -log10(res_M2Macs2_df_volcano$padj)
res_M2Macs2_df_volcano <- res_M2Macs2_df_volcano %>% filter(!is.na(neg.log10.padj))


res_M2Macs2_df_volcano <- res_M2Macs2_df_volcano %>%
  mutate(gene_type = case_when(log2FoldChange >= 1 & padj <= 0.05 ~ "up",
                               log2FoldChange <= -1 & padj <= 0.05 ~ "down",
                               TRUE ~ "ns"))  

significant.genes.M2.2 <- res_M2Macs2_df_volcano %>% filter(padj <= 0.05)
significant.genes.M2.2$geneID <- rownames(significant.genes.M2.2)
  
res_M2Macs2_df_volcano %>%
  ggplot(aes(x = log2FoldChange,
             y = neg.log10.padj,
             fill = gene_type,    
             size = gene_type,
             alpha = gene_type)) + 
  geom_point(shape = 21, # Specify shape and colour as fixed local parameters    
             colour = "black") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),
             linetype = "dashed") +
   geom_label_repel(data = significant.genes.M2.2,
                   aes(label = geneID),
                   force = 2,
                   nudge_y = 0.25) +
  scale_fill_manual(values = cols) + # Modify point colour
  scale_size_manual(values = sizes) + # Modify point size
  scale_alpha_manual(values = alphas) + # Modify point transparency
  scale_x_continuous(breaks = c(seq(-8, 8, 2)),       
                     limits = c(-8, 8)) + 
  labs(x = "log2(Fold Change)", y = "-log10(adj. P value)", 
       title = "Differentially expressed genes in M2 macrophages (2) pre and post CAR T") + theme_bw()
```
Durch den besseren Ausschluss der B Zellen ist diese DEG Analyse jetzt bereinigt. 

```{r DESeq2 for DCs}
# Create meta.data for count matrix
col.data.cDC2 <- data.frame(samples = colnames(cell.type.list.t.noIG$`cDC2`))

col.data.cDC2$treatment <- ifelse(col.data.cDC2$samples == "g2024-07-15" | 
                                    col.data.cDC2$samples == "g2024-11-28" |
                               col.data.cDC2$samples == "g2025-02-17", "pre_CART", "post_CART")

# Perform DESeq2 for M1_Macs
count.data.cDC2 <- cell.type.list.t.noIG$`cDC2`
class(count.data.cDC2) <- "integer"


dds <- DESeqDataSetFromMatrix(countData = count.data.cDC2, 
				colData =  col.data.cDC2,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds)

resultsNames(dds)

#Generate results object
res_cDC2 <- results(dds, name = "treatment_pre_CART_vs_post_CART")
res_cDC2 

res_cDC2_df <- as.data.frame(res_cDC2)

# Sort dataframe for adjusted p value and show top20 results 
res_cDC2_df_sort <- res_cDC2_df[order(res_cDC2_df$padj),]
head(res_cDC2_df_sort, 20)

## violin Plot for M2 Macrophages 2
# Violin Plot of DEGs
res_cDC2_df_volcano <- res_cDC2_df
res_cDC2_df_volcano$neg.log10.padj <- -log10(res_cDC2_df_volcano$padj)
res_cDC2_df_volcano <- res_cDC2_df_volcano %>% filter(!is.na(neg.log10.padj))


res_cDC2_df_volcano <- res_cDC2_df_volcano %>%
  mutate(gene_type = case_when(log2FoldChange >= 1 & padj <= 0.05 ~ "up",
                               log2FoldChange <= -1 & padj <= 0.05 ~ "down",
                               TRUE ~ "ns"))  

significant.genes.cDC2 <- res_cDC2_df_volcano %>% filter(padj <= 0.05)
significant.genes.cDC2$geneID <- rownames(significant.genes.cDC2)
  
res_cDC2_df_volcano %>%
  ggplot(aes(x = log2FoldChange,
             y = neg.log10.padj,
             fill = gene_type,    
             size = gene_type,
             alpha = gene_type)) + 
  geom_point(shape = 21, # Specify shape and colour as fixed local parameters    
             colour = "black") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),
             linetype = "dashed") +
   geom_label_repel(data = significant.genes.cDC2,
                   aes(label = geneID),
                   force = 2,
                   nudge_y = 0.25) +
  scale_fill_manual(values = cols) + # Modify point colour
  scale_size_manual(values = sizes) + # Modify point size
  scale_alpha_manual(values = alphas) + # Modify point transparency
  # scale_x_continuous(breaks = c(seq(-8, 8, 2)),       
  #                    limits = c(-8, 8)) + 
  labs(x = "log2(Fold Change)", y = "-log10(adj. P value)", 
       title = "Differentially expressed genes in cDC2 pre and post CAR T") + theme_bw()
```

### DEG Analysis for Myeloid cell Subpopulations (Resolution 0.2)


```{r DESeq2 for M1 Macrophages}
myeloid.02.seurat <- readRDS("Myeloid_Cells_res02_annotated.rds")

myeloid.02.pseudo <- AggregateExpression(myeloid.02.seurat, assays = "RNA", 
                                        group.by = c("date", "Annotation"), slot = "counts",
                                        return.seurat = F)

# Manipulate the matrix to fit DESeq2 input
matrix.02.t <- t(myeloid.02.pseudo$RNA)
count.02.data.frame <- as.data.frame(matrix.02.t)
splitRows <- gsub('.*_', '', rownames(count.02.data.frame))
cell.type.list.02 <- split.data.frame(count.02.data.frame, f = factor(splitRows))

#Retranspose the matrix and fix column names for all elements of the list
cell.type.list.02.t <- lapply(cell.type.list.02, function(x){
	rownames(x) <- gsub('(.*)_.*', '\\1', rownames(x))
	t(x)
})

cell.type.list.02.t.noIG <- lapply(cell.type.list.02.t, function(x){
    x <- x[-IG.transcripts,]
})

# Create meta.data for count matrix
col.data.02 <- data.frame(samples = colnames(cell.type.list.02.t.noIG$`M1 Macrophages`))

col.data.02$treatment <- ifelse(col.data.02$samples == "g2024-07-15" | col.data.02$samples == "g2024-11-28" |
                               col.data.02$samples == "g2025-02-17", "pre_CART", "post_CART")

# Perform DESeq2 for M1_Macs
count.data.M1 <- cell.type.list.02.t.noIG$`M1 Macrophages`
class(count.data.M1) <- "integer"


dds <- DESeqDataSetFromMatrix(countData = count.data.M1, 
				colData =  col.data.02,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds)

resultsNames(dds)

#Generate results object
res <- results(dds, name = "treatment_pre_CART_vs_post_CART")
res 

res_M1_df <- as.data.frame(res)

# Sort dataframe for adjusted p value and show top20 results 
res_M1_df_sort <- res_M1_df[order(res_M1_df$padj),]
head(res_M1_df_sort, 20)

## violin Plot for M1 Macrophages
# Violin Plot of DEGs
res_M1_df_volcano <- res_M1_df
res_M1_df_volcano$neg.log10.padj <- -log10(res_M1_df_volcano$padj)
res_M1_df_volcano <- res_M1_df_volcano %>% filter(!is.na(neg.log10.padj))


res_M1_df_volcano <- res_M1_df_volcano %>%
  mutate(gene_type = case_when(log2FoldChange >= 1 & padj <= 0.05 ~ "up",
                               log2FoldChange <= -1 & padj <= 0.05 ~ "down",
                               TRUE ~ "ns"))  

significant.genes.M1 <- res_M1_df_volcano %>% filter(padj <= 0.05)
significant.genes.M1$geneID <- rownames(significant.genes.M1)
  
res_M1_df_volcano %>%
  ggplot(aes(x = log2FoldChange,
             y = neg.log10.padj,
             fill = gene_type,    
             size = gene_type,
             alpha = gene_type)) + 
  geom_point(shape = 21, # Specify shape and colour as fixed local parameters    
             colour = "black") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),
             linetype = "dashed") +
   geom_label_repel(data = significant.genes.M1, # Add labels last to appear as the top layer  
                   aes(label = geneID),
                   force = 2,
                   nudge_y = 0.25) +
  scale_fill_manual(values = cols) + # Modify point colour
  scale_size_manual(values = sizes) + # Modify point size
  scale_alpha_manual(values = alphas) + # Modify point transparency
  scale_x_continuous(breaks = c(seq(-8, 8, 2)),
                     limits = c(-8, 8)) +
  labs(x = "log2(Fold Change)", y = "-log10(adj. P value)", 
       title = "Differentially expressed genes in M1 macrophages pre and post CAR T") + theme_bw()
```


```{r DESeq2 for M2 Macs Resolution 02}
# Create meta.data for count matrix
col.data.02.InflMo <- data.frame(samples = 
                               colnames(cell.type.list.02.t.noIG$`Inflammatory monocyte-derived macrophage`))

col.data.02.InflMo$treatment <- ifelse(col.data.02.InflMo$samples == "g2024-07-15" | 
                                     col.data.02.InflMo$samples == "g2024-11-28" |
                               col.data.02.InflMo$samples == "g2025-02-17", "pre_CART", "post_CART")

# Perform DESeq2 for M1_Macs
count.data.InflMo <- cell.type.list.02.t.noIG$`Inflammatory monocyte-derived macrophage`
class(count.data.InflMo) <- "integer"


dds <- DESeqDataSetFromMatrix(countData = count.data.InflMo, 
				colData =  col.data.02.InflMo,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds)

resultsNames(dds)

#Generate results object
res <- results(dds, name = "treatment_pre_CART_vs_post_CART")
res 

res_InflMo_df <- as.data.frame(res)

# Sort dataframe for adjusted p value and show top20 results 
res_InflMo_df_sort <- res_InflMo_df[order(res_InflMo_df$padj),]
head(res_InflMo_df_sort, 20)

## violin Plot for InflMo Macrophages
# Violin Plot of DEGs
res_InflMo_df_volcano <- res_InflMo_df
res_InflMo_df_volcano$neg.log10.padj <- -log10(res_InflMo_df_volcano$padj)
res_InflMo_df_volcano <- res_InflMo_df_volcano %>% filter(!is.na(neg.log10.padj))


res_InflMo_df_volcano <- res_InflMo_df_volcano %>%
  mutate(gene_type = case_when(log2FoldChange >= 1 & padj <= 0.05 ~ "up",
                               log2FoldChange <= -1 & padj <= 0.05 ~ "down",
                               TRUE ~ "ns"))  

significant.genes.InflMo <- res_InflMo_df_volcano %>% filter(padj <= 0.05)
significant.genes.InflMo$geneID <- rownames(significant.genes.InflMo)
  
res_InflMo_df_volcano %>%
  ggplot(aes(x = log2FoldChange,
             y = neg.log10.padj,
             fill = gene_type,    
             size = gene_type,
             alpha = gene_type)) + 
  geom_point(shape = 21, # Specify shape and colour as fixed local parameters    
             colour = "black") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),
             linetype = "dashed") +
   geom_label_repel(data = significant.genes.InflMo, # Add labels last to appear as the top layer  
                   aes(label = geneID),
                   force = 2,
                   nudge_y = 0.25) +
  scale_fill_manual(values = cols) + # Modify point colour
  scale_size_manual(values = sizes) + # Modify point size
  scale_alpha_manual(values = alphas) + # Modify point transparency
  scale_x_continuous(breaks = c(seq(-25, 25, 5)),
                     limits = c(-25, 25)) +
  labs(x = "log2(Fold Change)", y = "-log10(adj. P value)", 
       title = "Differentially expressed genes in InflMo macrophages pre and post CAR T") + theme_bw()
```

### Test Custom IgG Immune Complex Signature from Uzzan et al. 


```{r Test Custom IgG Signature}
Macrophage.Idents <- levels(Idents(myeloid.seurat))[grep("*Macrophage*", levels(Idents(myeloid.seurat)))]
Myeloid.macrophage <- subset(myeloid.seurat, idents = Macrophage.Idents)

Myeloid.macrophage.pseudo <- AggregateExpression(Myeloid.macrophage, assays = "RNA", 
                                        group.by = "date", slot = "counts",
                                        return.seurat = F)

# Create meta.data for count matrix
Myeloid.macrophage.numbers <- Myeloid.macrophage@meta.data %>% group_by(date) %>% summarise(Macro.count = n())
Myeloid.macrophage.numbers$date <- paste0("g", Myeloid.macrophage.numbers$date)
Myeloid.macrophage.numbers$treatment <- ifelse(Myeloid.macrophage.numbers$date == "g2024_07_15" | 
                                    Myeloid.macrophage.numbers$date == "g2024_11_28" |
                               Myeloid.macrophage.numbers$date == "g2025_02_17", "pre_CART", "post_CART")

# Visualize cell counts
ggplot(Myeloid.macrophage.numbers, aes(x=date, y=Macro.count, fill=treatment)) +
    geom_bar(stat="identity", color="black") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(x="Timepoint", y="Number of Macrophages") +
    geom_text(aes(label=Macro.count), vjust=-0.5)

# Perform DESeq2 for Myeloid pseudobulk
# To prevent the B cell pathways from popping up, I am removing all the IG transcripts at this step:
count.data.all.macrophages <- as.matrix(Myeloid.macrophage.pseudo$RNA)

IG.transcripts <- c(which(grepl("^IGH", rownames(count.data.all.macrophages)) == T), 
                    which(grepl("^IGL", rownames(count.data.all.macrophages)) == T), 
                    which(grepl("^IGK", rownames(count.data.all.macrophages)) == T))

count.data.all.macrophages.noIG <- count.data.all.macrophages[-IG.transcripts,]
class(count.data.all.macrophages.noIG) <- "integer"

Myeloid.macrophage.numbers$date <- gsub("_", "-", Myeloid.macrophage.numbers$date)

dds.all.macrophages <- DESeqDataSetFromMatrix(countData = count.data.all.macrophages.noIG, 
				colData =  Myeloid.macrophage.numbers,
 				design = ~ treatment)

#Remove all genes with fewer than 10 reads 
keep <- rowSums(counts(dds.all.macrophages)) >= 10
dds.all.macrophages <- dds.all.macrophages[keep,]

dds.all.macrophages <- DESeq(dds.all.macrophages)

resultsNames(dds.all.macrophages)

#Generate results object
res.all.macrophages <- results(dds.all.macrophages, name = "treatment_pre_CART_vs_post_CART")
res.all.macrophages 

res.all.macrophages_df <- as.data.frame(res.all.macrophages)

# Sort dataframe for adjusted p value and show top20 results 
res.all.macrophages_df_sort <- res.all.macrophages_df[order(res.all.macrophages_df$padj),]
head(res.all.macrophages_df_sort, 20)

write.csv(res.all.macrophages_df_sort, "U:/Analysen/2025_Colitis_All_Timepoints/DESeq2/DEGs_macrophages.csv")

# Violin Plot of DEGs
res.all.macrophages_df_volcano <- res.all.macrophages_df
res.all.macrophages_df_volcano$neg.log10.padj <- -log10(res.all.macrophages_df_volcano$padj)
res.all.macrophages_df_volcano <- res.all.macrophages_df_volcano %>% filter(!is.na(neg.log10.padj))


res.all.macrophages_df_volcano <- res.all.macrophages_df_volcano %>%
  mutate(gene_type = case_when(log2FoldChange >= 1 & padj <= 0.05 ~ "up",
                               log2FoldChange <= -1 & padj <= 0.05 ~ "down",
                               TRUE ~ "ns"))  

cols <- c("up" = "#ffad73", "down" = "#26b3ff", "ns" = "grey") 
sizes <- c("up" = 2, "down" = 2, "ns" = 1) 
alphas <- c("up" = 1, "down" = 1, "ns" = 0.5)

significant.genes <- res.all.macrophages_df_volcano %>% filter(padj <= 0.05)
significant.genes$geneID <- rownames(significant.genes)
  
res.all.macrophages_df_volcano %>%
  ggplot(aes(x = log2FoldChange,
             y = neg.log10.padj,
             fill = gene_type,    
             size = gene_type,
             alpha = gene_type)) + 
  geom_point(shape = 21, # Specify shape and colour as fixed local parameters    
             colour = "black") + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),
             linetype = "dashed") +
   geom_label_repel(data = significant.genes, # Add labels last to appear as the top layer  
                   aes(label = geneID),
                   force = 2,
                   nudge_y = 0.25) +
  scale_fill_manual(values = cols) + # Modify point colour
  scale_size_manual(values = sizes) + # Modify point size
  scale_alpha_manual(values = alphas) + # Modify point transparency
  scale_x_continuous(breaks = c(seq(-8, 8, 2)),       
                     limits = c(-8, 8)) + 
  labs(x = "log2(Fold Change)", y = "-log10(adj. P value)", 
       title = "Differentially expressed genes in macrophages pre and post CAR T") + theme_bw()


res.all.macrophages_df_sort$sign.log.10.p <-
  sign(res.all.macrophages_df_sort$log2FoldChange)*(-log10(res.all.macrophages_df_sort$pvalue))
res.all.macrophages_df_sort <- res.all.macrophages_df_sort %>% filter(!is.na(sign.log.10.p))
res.all.macrophages_df_sort <- res.all.macrophages_df_sort[order(res.all.macrophages_df_sort$sign.log.10.p, 
                                                         decreasing = T),]

ranked_genes_M <- as.vector(res.all.macrophages_df_sort$sign.log.10.p)
names(ranked_genes_M) <- rownames(res.all.macrophages_df_sort)

#run GSEA
All.macrophages.HALLMARK <- GSEA(ranked_genes_M, TERM2GENE = hallmark, TERM2NAME = h2, seed=TRUE)

head(All.macrophages.HALLMARK)

print(paste0(All.macrophages.HALLMARK@result$ID, " - NES:  ", All.macrophages.HALLMARK@result$NES,  " - padj:  ",
             All.macrophages.HALLMARK@result$p.adjust))

All.macrophages.Hallmark.df <- data.frame(ID = All.macrophages.HALLMARK@result$ID, 
                             NES = All.macrophages.HALLMARK@result$NES, 
                             padj = All.macrophages.HALLMARK@result$p.adjust)

All.macrophages.Hallmark.df <- All.macrophages.Hallmark.df[order(All.macrophages.Hallmark.df$NES),]
All.macrophages.Hallmark.df$ID <- factor(All.macrophages.Hallmark.df$ID, levels = All.macrophages.Hallmark.df$ID)

ggplot(All.macrophages.Hallmark.df, aes(x = ID, y = NES, fill = padj)) + geom_col() + 
  scale_fill_gradient() + coord_flip() + theme_minimal()

#run GSEA with GO:BP 
All.macrophages.GO <- GSEA(ranked_genes_M, TERM2GENE = GO_BP, TERM2NAME = GO_BP_t2n, seed=TRUE)

head(All.macrophages.GO)

print(paste0(All.macrophages.GO@result$ID, " - NES:  ", All.macrophages.GO@result$NES,  " - padj:  ",
             All.macrophages.GO@result$p.adjust))

All.macrophages.GO.df <- data.frame(ID = All.macrophages.GO@result$ID, 
                             NES = All.macrophages.GO@result$NES, 
                             padj = All.macrophages.GO@result$p.adjust)

All.macrophages.GO.df.top10.up <- All.macrophages.GO.df %>% filter(NES > 0) %>% slice_min(padj, n = 10) %>%
  slice_max(NES, n = 10)
All.macrophages.GO.df.top10.dn <- All.macrophages.GO.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
All.macrophages.GO.df.top20 <- rbind(All.macrophages.GO.df.top10.up, All.macrophages.GO.df.top10.dn)

All.macrophages.GO.df.top20 <- All.macrophages.GO.df.top20[order(All.macrophages.GO.df.top20$NES),]
All.macrophages.GO.df.top20$ID <- factor(All.macrophages.GO.df.top20$ID, levels = All.macrophages.GO.df.top20$ID)

ggplot(All.macrophages.GO.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() + 
  scale_fill_gradient() + coord_flip() + theme_minimal()


#run GSEA
All.macrophages.REACTOME <- GSEA(ranked_genes_M, TERM2GENE = REACTOME, TERM2NAME = REACTOME_t2n, seed=TRUE)

All.macrophages.REACTOME.df <- data.frame(ID = All.macrophages.REACTOME@result$ID, 
                             NES = All.macrophages.REACTOME@result$NES, 
                             padj = All.macrophages.REACTOME@result$p.adjust)

All.macrophages.REACTOME.df.top10.up <- All.macrophages.REACTOME.df %>% filter(NES > 0) %>% slice_min(padj, n = 10) %>%
  slice_max(NES, n = 10)
All.macrophages.REACTOME.df.top10.dn <- All.macrophages.REACTOME.df %>% filter(NES < 0) %>% slice_min(padj, n = 10)
All.macrophages.REACTOME.df.top20 <- rbind(All.macrophages.REACTOME.df.top10.up, All.macrophages.REACTOME.df.top10.dn)

All.macrophages.REACTOME.df.top20 <- All.macrophages.REACTOME.df.top20[order(All.macrophages.REACTOME.df.top20$NES),]
All.macrophages.REACTOME.df.top20$ID <- factor(All.macrophages.REACTOME.df.top20$ID, 
                                               levels = All.macrophages.REACTOME.df.top20$ID)

ggplot(All.macrophages.REACTOME.df.top20, aes(x = ID, y = NES, fill = padj)) + geom_col() +
  scale_fill_gradient() + coord_flip() + theme_minimal() + 
  labs(title = "REACTOME Top 20 enriched Gene Sets in macrophages") + 
  theme(axis.text.x = element_text(size = 7), axis.text.y = element_text(size = 7))
```


```{r GSEA summary figure, fig.width=8, fig.height=8}
all.GSEA.mac <- rbind(All.macrophages.Hallmark.df, All.macrophages.GO.df.top20, All.macrophages.REACTOME.df.top20)
all.GSEA.mac$DB <- ifelse(grepl("HALLMARK", all.GSEA.mac$ID), "HALLMARK", NA)
all.GSEA.mac$DB <- ifelse(grepl("GOBP", all.GSEA.mac$ID), "GOBP", all.GSEA.mac$DB)
all.GSEA.mac$DB <- ifelse(grepl("REACTOME", all.GSEA.mac$ID), "REACTOME", all.GSEA.mac$DB)
all.GSEA.mac$DB <- factor(all.GSEA.mac$DB, levels = c("HALLMARK", "GOBP", "REACTOME"))

all.GSEA.mac$Status <- ifelse(all.GSEA.mac$NES < 0, "post CART", NA)
all.GSEA.mac$Status <- ifelse(all.GSEA.mac$NES > 0, "pre CART", all.GSEA.mac$Status)

all.GSEA.mac.preCAR <- all.GSEA.mac %>% filter(Status == "pre CART") %>% group_by(DB) %>% slice_max(NES, n = 5)
all.GSEA.mac.postCAR <- all.GSEA.mac %>% filter(Status == "post CART") %>% group_by(DB) %>% slice_min(NES, n = 5)
all.GSEA.mac.top5 <- rbind(all.GSEA.mac.preCAR, all.GSEA.mac.postCAR)
all.GSEA.mac.top5$Status <- factor(all.GSEA.mac.top5$Status, levels = c("pre CART", "post CART"))

all.GSEA.mac.top5$ID <- gsub("GOBP", "", all.GSEA.mac.top5$ID)
all.GSEA.mac.top5$ID <- gsub("REACTOME", "", all.GSEA.mac.top5$ID)
all.GSEA.mac.top5$ID <- gsub("HALLMARK", "", all.GSEA.mac.top5$ID)
all.GSEA.mac.top5$ID <- gsub("_", " ", all.GSEA.mac.top5$ID)

all.GSEA.mac.top5 <- all.GSEA.mac.top5 %>% mutate(log.10.padj = -log10(padj))

ggplot(all.GSEA.mac.top5, aes(x = Status, y = reorder_within(ID, NES, DB))) + 
  geom_point(aes(color = log.10.padj, size = abs(NES))) + scale_y_reordered() + scale_size(range = c(3,6)) +
  scale_x_discrete(expand = c(0.5,0.5)) +
  ggforce::facet_col(~DB, scales = "free_y", space = "free") +
   scale_color_gradient(low = "darkblue", high = "purple") + 
  labs(color = "-log10\n(adj. P-value)", size = "NES") +
  theme_classic(base_size = 13) +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_text(size = 10),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "right", 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("/Figures/GSEA_summary_Macrophages.tiff", 
       plot = last_plot(), device = "tiff", dpi = 300, width = 8, height = 8, units = "in", compression = "lzw")


## Without REACTOME
all.GSEA.mac.noREACT <- rbind(All.macrophages.Hallmark.df, All.macrophages.GO.df.top20)
all.GSEA.mac.noREACT$DB <- ifelse(grepl("HALLMARK", all.GSEA.mac.noREACT$ID), "HALLMARK", NA)
all.GSEA.mac.noREACT$DB <- ifelse(grepl("GOBP", all.GSEA.mac.noREACT$ID), "GOBP", all.GSEA.mac.noREACT$DB)
all.GSEA.mac.noREACT$DB <- factor(all.GSEA.mac.noREACT$DB, levels = c("HALLMARK", "GOBP"))

all.GSEA.mac.noREACT$Status <- ifelse(all.GSEA.mac.noREACT$NES < 0, "post CART", NA)
all.GSEA.mac.noREACT$Status <- ifelse(all.GSEA.mac.noREACT$NES > 0, "pre CART", all.GSEA.mac.noREACT$Status)

all.GSEA.mac.noREACT.preCAR <- all.GSEA.mac.noREACT %>% filter(Status == "pre CART") %>% group_by(DB) %>% 
  slice_max(NES, n = 5)
all.GSEA.mac.noREACT.postCAR <- all.GSEA.mac.noREACT %>% filter(Status == "post CART") %>% group_by(DB) %>% 
  slice_min(NES, n = 5)
all.GSEA.mac.noREACT.top5 <- rbind(all.GSEA.mac.noREACT.preCAR, all.GSEA.mac.noREACT.postCAR)
all.GSEA.mac.noREACT.top5$Status <- factor(all.GSEA.mac.noREACT.top5$Status, levels = c("pre CART", "post CART"))

all.GSEA.mac.noREACT.top5$ID <- gsub("GOBP", "", all.GSEA.mac.noREACT.top5$ID)
all.GSEA.mac.noREACT.top5$ID <- gsub("HALLMARK", "", all.GSEA.mac.noREACT.top5$ID)
all.GSEA.mac.noREACT.top5$ID <- gsub("_", " ", all.GSEA.mac.noREACT.top5$ID)

all.GSEA.mac.noREACT.top5 <- all.GSEA.mac.noREACT.top5 %>% mutate(log.10.padj = -log10(padj))

ggplot(all.GSEA.mac.noREACT.top5, aes(x = Status, y = reorder_within(ID, NES, DB))) + 
  geom_point(aes(color = log.10.padj, size = abs(NES))) + scale_y_reordered() + scale_size(range = c(3,6)) +
  scale_x_discrete(expand = c(0.5,0.5)) +
  ggforce::facet_col(~DB, scales = "free_y", space = "free") +
   scale_color_gradient(low = "darkblue", high = "purple") + 
  labs(color = "-log10\n(adj. P-value)", size = "NES") +
  theme_classic(base_size = 13) +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_text(size = 10),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "right", 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("/Figures/GSEA_summary_Macrophages_noREACTOME.tiff", 
       plot = last_plot(), device = "tiff", dpi = 300, width = 6, height = 6, units = "in", compression = "lzw")
```


```{r Custom Gene Sets}
# Custom gene set 
Uzzan_IgG_gene_set <- c("IL1RN", "INHBA", "EGR2", "TGM2", "SPP1", "PLEK", "DUSP7", "SOCS3", "GDF15", "CPEB4",
                        "BCL2L1", "PDPN", "PTGS2", "IL1A", "EDN1", "SLAMF1", "GALNT7", "GBP2", "DNAJC19", "GSDMD",
                        "MYO1F", "ARHGEF10L", "APOL3", "BMF", "TLR8")

Uzzan_LPS_gene_set <- c("HIF1A", "SOD2", "HCAR3", "FLT1", "IL1B", "CFB", "DDIT4", "G0S2", "TREM1", "CLEC4E", "IL15RA",
                        "RND1", "REV3L", "SLCO4A1", "CXCL9", "AGRN", "SLC25A37", "TIMP1", "IRAK3", "IFITM1", "C4orf3",
                        "SPP1", "CALCRL", "SPON1", "SLAMF1", "TGM2", "SMAD3", "MMP3", "GPC1", "PTCH1", "CYP27A1", "ST18",
                        "FKBP4", "THEMIS2", "SULF2", "PLAU", "IRF2BP2")

custom_gene_set <- list(Uzzan_IgG_gene_set, Uzzan_LPS_gene_set)
names(custom_gene_set) <- c("Uzzan_IgG_complex", "Uzzan_LPS_stim")

term2gene <- stack(custom_gene_set) %>%
    dplyr::select(term = ind, gene = values)

#run GSEA
All.macrophages.Uzzan.IgG <- GSEA(ranked_genes_M, TERM2GENE = term2gene, seed=TRUE)

head(All.macrophages.Uzzan.IgG)

print(paste0(All.macrophages.Uzzan.IgG@result$ID, " - NES:  ", All.macrophages.Uzzan.IgG@result$NES,  " - padj:  ",
             All.macrophages.Uzzan.IgG@result$p.adjust))

enrichplot::gseaplot2(All.macrophages.Uzzan.IgG, geneSetID = 1, pvalue_table = TRUE, color = "#86B875")
p <- enrichplot::gseaplot2(All.macrophages.Uzzan.IgG, geneSetID = 1, pvalue_table = TRUE, color = "#86B875")

png("/Figures/GSEA_Macrophages_IgG.png", 
    width = 2000, height = 1500, res = 300)
print(p)  # Important: use print() for ggplot objects
dev.off()
```


